---
title: "Math 120 Lec 01"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```



# Week 1

## Wednesday Feb 3


This is Rstudio!

I can write stuff, but I can also do math:

```{r}
1+1
```

Sometimes you'll see me use code, like this:  let's look at the famous "iris" dataset. 

```{r}
head(iris)
```


You don't need to know any R code, but you ABSOLUTELY SHOULD understand all mathematical computations.



## 1.2: Data Basics

What's a dataset?

There are a several ways to represent data.  In 120, we'll always use **rectangular data**.

Data is rectangular if:

- Each row is an individual (case)
- Each column is a variable, i.e., a quality of the individuals that we're measuring
- Each cell/entry contins a specivic observation/value

Goal:  look at a dataset, and say smart things about the individuals in it.

### Types of Data

Not all data is created equal!

Need to be **very** careful -- different data types require different tools to analyze.


First major division:  some data are numbers, some are not!

Ex:  makes sense to find mean sepal length, but makes no sense to find "mean species".

Broadly speaking:

-quantitative/numerical data:  it's a number
-qualitative/categorical data:  it's not a number

We can subdivide further:

Quantitative data can be either:

- discrete: countable.  whole numbers.  Ex:  number of students in this class.  # of siblings you have.  Etc.  
- continuous:  measurable.  could take any real number (within reasonable bounds).  Ex:  height, GPA, temperature, weight.  In general:  any physical measurement of the universe.  

Caveats:

- Height.  We usually round our height to whole number.  But it's still cts!  Depends on the nature of the variable, not how we round it.  

- Money.  Let's just agree that money is continuous.



Qualitative data can also be subdivided into two types:

- nominal data: no intrinsic order.  Ex: color,  species, gender, nationality, language, 
- ordinal data: DO have intrinsic order.  Ex:  Age, Month, Alphabet, Income level, Educational attainment,  minutes in the our, year in school, seasons

Caveats:

- Age:  might seem like quantitative, but careful, it's a category!  Same with minutes in the hour.  

- We can always ASSIGN and ordering.  Like color:  we -could- order by wavelength.  


### Metadata

Metadata is data about data.

Almost always, we study data collected by someone else.  Not always clear what the variables represent.  Ex:  "mpg dataset"

```{r}
head(mpg)
```

Metadata tells:

- what the cases are (rows)
- what variables mean (columns)
- what the units are (when applicable)









































