---
title: "Math 120 Week 03"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```



1) Find $\frac{dy}{dx}$ for 

$$x^3+y^3=3xy$$


2) Find the equation of the tangent line $L(X)$ at the point the above curve at the point $\left(\frac{3}{2},\frac{3}{2}\right)$


3) Find $\frac{dy}{dx}$, and also all coordinates where the curve has a vertical tangent:

$$x=\frac{1}{3}y^3-\frac{1}{2}y^2-6y$$

4)  Find $\frac{dy}{dx}$ for 

$$6x^2y^2+\sin(xy)=2x$$











```{r}
(17-18)/2.5
```


```{r}
(20-18)/2.5
```


```{r}
.7881-.3446
```





```{r}
1.04*2.5+18
```





1)  The average wait time at a particular bus stop is 12 min, with stdev 2.1 min.  If we take a sample of 35 people, what's the probability that their mean wait time is at least 11.5 min?


```{r}
1-pnorm(11.5, 12, 2.1/sqrt(35))
```


2)  The average wait time at a particular bus stop is 12 min, with stdev 2.1 min.  If we take a sample of 50 people, how high must their average wait time be in order to be in the top 5% of sample mean wait times?

```{r}
qnorm(.95, 12, 2.1/sqrt(50))
```


3)  This year at DU, 51% of students are female.  In a sample of 45 students, what's the probability that at least half of them are female?


```{r}
1-pnorm(.5, .51, sqrt(.51*.49/45))
```




