---
title: "Math 120 HW"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 1
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```


ARC

Jacob Brown
Max




- random process/experiment
- sample space
- event
- relations
- mututally exclusive
- 3 axioms of probability
- P(empty)
- law of large numbers
- complement
- general addition rule
- equally likely outcomes


















 can do Wednesdays, 4:30pm-6:30pm, and Sundays, 4:00pm-7:00pm. I hope this works for everyone.

Best,
Ted


sunday 1-4pm, monday 4-6


\



```{r}
choose(13,2)*choose(4,2)*choose(4,2)*choose(11,1)*choose(4,1)
```



38

```{r}
x <- c(120, 125, 140, 130, 115, 120, 110, 130, 135)

median(x)
```

42)  a) add constant
     b) multiply by constant
     
     
a)  mean({y1, y2, .., y3})


44)  

```{r}
x <- c(180.5, 181.7, 180.9, 181.6, 182.6, 181.6, 181.3, 182.1, 182.1, 180.3, 181.7, 180.5)

sum(x)^2
```

```{r}
.5/15
```

```{r}
182.6-180.3
```

```{r}
1/15
```

2.64
3.32












6.22:


```{r}
data<-c(.92, .79, .90, .65, .86, .47, .73, .97, .94, .77)
```

method of moments:

```{r}
(2*mean(data)-1)/(1-mean(data))
```

MLE:
```{r}
-1-10/sum(log(data))
```


Q:  which is better?  MOM or MLE?

1)  Not all functions are differentiable!  MLE doesn't work!

2)  The MLE is "almost unbiased". As n increases, limits to unbiased.














