---
title: "Math 120 HW"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 1
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```


# HW 2 - What's In Your Breakfast Cereal?

Due Friday Oct 1 before the beginning of class.  Submit as .pdf document, just like HW 1.

For this HW you'll be looking at data on a selection of common breakfast cereals.  Find the data on our [shared Google Drive folder](https://drive.google.com/drive/folders/1C1XHN3PGe4lMQXS_hq49btBM6QaiEuBW?usp=sharing) (you can click the link here).

(For the record, I'm a Honey Bunches of Oats woman, myself.)

Your goal is to create and analyze linear models for the following:

- Predict `Fiber` based on `Potassium`
- Predict `Protein` based on `Sugar`.

For each, perform the following:

  1) Make a scatterplot, and describe its key features using language from lecture.
  2) Compute the 5 key summary statistics:
    $\bar{x}$, $\bar{y}$, $s_x$, $s_y$, $r$
  3) Compute the equation of the least-squares line.
  4) Carefully interpret the slope of the line, as practiced in lecture.
  4) Make, and interpret, a residual plot.
  5) Compute and interpret $R^2$.

Everything should be neatly and clearly presented in a Google Docs report in complete English sentences and paragraphs.  Exported to .pdf.  

Late HW will not be accepted for any reason.  Students may turn in HW early, of course.





A researcher collects data on 22 study participants.  She measures both their daily coffee consumption (in ounces) and their mean nightly sleep time (in hours) for the duration of the study.

Afterwards, she computes a linear model for predicting sleep time based on coffee consumption:

yhat = 7.834 - .104*x

r = -0.391

SSResid = 6.54

sum( (x-xbar)^2 ) = 98.3

Is there evidence of a linear relationship between coffee consumption and sleep time?  Perform a complete hyp test.

```{r}
-.391*sqrt(20)/sqrt(1-(-.391)^2)
```


```{r}
-0.391*sqrt(20)/sqrt(1-(-0.391^2)) 
```


```{r}
2*pt(-1.63, 20)
```


```{r}
-.104 - 2.85*sqrt(6.54/20)/sqrt(98.3)
```

```{r}
-.104+2.85*sqrt(6.54/20)/sqrt(98.3)
```



```{r}
-0.104-2.82*sqrt(6.54/20)/sqrt(98.3)
```

```{r}
1+1
```


```{r}
-0.391*sqrt(22-2)/sqrt(1-(-0.391^2)) 
```




```{r}
df <- 20
1-pt(.391*sqrt(df)/sqrt(1-.391^2), df)
```

```{r}
xbar=8
sx=3
ybar=7
sy=.8
r=-.391

(r*sy/sx)
```

```{r}
sqrt(6.54/20)/sqrt(98)
```

1)  r=0.416,  n=38.  Evidence of a linear relationship?

2)  

 - yhat = 15.2 - 2.4x
 - SSResid = 24.91
 - sum( (x-xbar)^2 ) = 6780
 - n = 35
 
 Compute 99% CI for slope.