---
title: "Lec 02"
author: "Ali Miller"
date: "9/01/2019"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}
###########################################################################
################# DON'T DELETE THIS PART!!!!!##############################
###########################################################################
                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans"))                                 #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #
### Below is where you read and store your data.                          #
                                                                          #
deathPenaltyData <- read_csv("death penalty data.csv")
                                                                          #
                                                                          #
                                                                          #
############# ^ ################################# ^ #######################
############# | ## DON'T DELETE THIS PART!!!!!### | #######################
############# | ################################# | #######################
```

## Wakeup Challenge

Star wars with yellow eyes:

```{r}

starwars %>% filter(eye_color == "yellow") %>% nrow()



```


## Back to death penalty

Basic plots:  age, sex, race

```{r}


qplot(
      deathPenaltyData$Age,
      xlab = "Age",
      ylab = "Count",
      main = "Age of executed prisoners"
      )


qplot(Sex, data=deathPenaltyData,
      xlab="Sex",
      ylab="Count",
      main = "Sex of executed prisoners")


```

Question:  what percent of executed prisoners are women?

Answer:  filter and nrow!

```{r}

#how many in whole data set?

nrow(deathPenaltyData)


head(deathPenaltyData)

deathPenaltyData %>% filter(Sex == "Female") %>% nrow


#find the percent
16/1442

```

Wow!  Of the 1442 people who've received the death penalty, only 16 of them (1.1%) were women!



## Using select()

Select lets us choose which variables we want:

```{r}

deathPenaltyData %>% head()

deathPenaltyData %>% select(Date, Age, Sex, Race, State, Method, `Victim Count`)

#We picked our variables, but don't forget to save them with <-   !!!

deathPenaltyData %>% 
  select(Date, Age, Sex, Race, State, Method, `Victim Count`) ->
  myDeathPenaltyData

#Did it work?  Does myDeath..as;dlfja have the right stuff?

#Let's look with the head() commmand:  it shows you the first 6 rows

myDeathPenaltyData %>% head()

#is there other stuff head() can do?  Use the help() command!

help("head")


```



## Summarise and Group_by

These let us "collapse" data along a variable.  Let me show you!

```{r}


#Find total victim count in each state

#i.e., group by state, then summarise by total:

myDeathPenaltyData %>% group_by(State) %>% 
                       summarise(totalVictimCount = sum(`Victim Count`))

# what about average victim count?  Use mean() instead of sum()

myDeathPenaltyData %>% group_by(State) %>%
                       summarise(avgVictimCount = mean(`Victim Count`), totalVictimCount = sum(`Victim Count`))


```


Useful shortcut:  "count" means both group_by and summarise with total:

```{r}

myDeathPenaltyData %>% count(Race, State)





```


## Faceting Graphs

Often, we want break up our visual summaries by group.  Ex:

```{r}


qplot(myDeathPenaltyData$State)

#Ok, what about by race?

qplot(data=myDeathPenaltyData, State, fill=Race)



```

The above technique (using fill=Race) is called "faceting".  It breaks up our visuals by some categorical variable.

Q:  would it make sense to facet by the variable:  Age?

```{r}

qplot(data=myDeathPenaltyData, State, fill=Age)

```

It doesn't make sense to facet on quantitative variables!!!!  Need categorical, and hopefully, not too many categories!









