---
title: "DA 101 Lec 10"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```

## Wake-up challenge: dplyr

We'll use the built-in `mpg` data.  

```{r}
mpg %>% head
```

Your challenge, shgould you choose to accept it (and you will!):

1) Make a new table that shows the average mpg for each class.  (Very similar to part of your lab).  

Which dplyr function?

Not select, since select lets you choose variables:

```{r}
mpg %>% select(year)
```

Needed:  `summarise()` !!  But first, `group_by()`

```{r}
#first, group_by.  Then, summarize avg cty mpg.

mpg %>% group_by(class) %>% 
        summarise(meanCityMpg = mean(cty))
```



2) Add to the existing table the average mpg for each class.

Which function?  `mutate()`!

```{r}
mpg %>% group_by(class) %>% mutate(meanCtyMpg = mean(cty))
```




## CLT Demo with simulations: `sample` and `replicate`

Simulations are awesome!  They let us perform experiments with just our computers.  They can't tell us everything, but they're helpful for investigations.

First, let's look at sample:

```{r}
#demo of sample() for dice experiment

mySample <- sample(1:6, 100, replace = TRUE)

#let's look:

qplot(mySample, geom="histogram")



```

Seems reasonable;  we expect 1, 2, 3, 4, 5, 6 to be equal, but we know there's random variation.

Challenge:  plot a histo for a sample of size 5000

```{r}
mySample2 <- sample(1:6, 5000, replace = TRUE)

qplot(mySample2, geom="histogram")
```

```{r}
mySample2 <- sample(1:6, 50000, replace = TRUE)

qplot(mySample2, geom="histogram")
```

Conclusion:  as sample size increases, the proportions get closer and closer to the expected 1/6.   

**Example with means**

Let's replicate the full experiment:  100 samples of size 34

Use `replicate()`.

```{r}

mySampleMeans <- replicate(100, mean(sample(1:6, 34, replace = TRUE)))

#let's look:

qplot(mySampleMeans, geom="histogram", bins=12)

```

Observation:  even with a relatively low sample size (n=34) and not too many samples (100), we all can see the pattern.  Sample means follow a normal distribution.  Bigger n, closer to normal (bell curve).

Let's bump up:  5000 samples of size 1000


```{r}
mySampleMeans2 <- replicate(5000, mean(sample(1:6, 1000, replace = TRUE)))

#let's look:

qplot(mySampleMeans2, geom="histogram", bins=30)
```

Woooooah!  For large samples, we see that the means follow a normal/bell curve distribution.  Amazing fact:  THIS ALWAYS HAPPENS!!!  THE CENTRAL LIMIT THEOREM (CLT)!!!!!

No matter what our original population is, even if it's a super weird crazy non-normal population, sample means (with large n) **always** follow a normal dist.

This fact is the reason that we can find p-vals, since sample means are **very predictable**.

## Picture that connects the p-val to the null hypothesis

Here's how p-vals work:

1)  We make an assumption about the population (H0).  That value is the **center of our bell curve**

p is the probability of observing a sample result as or more extreme than the one we got.

In other words:  We know that the means follow a bell curve (normal).  We must make an assumption about the center (H0), and then, find the area out past our test stat.

TAIL AREA = PVAL!!!!!

If your test stat is waaaaay out in the tails, then our ovbserved data is unlikely according to the null hyp.  We trust our data, so we reject H0.  


## Left-tail, right-tail, two-tail

There are three possibilities for Ha.

H0 is *always* a statement of equality (=)

Ha depends on what we want to show:

- Ha:  mu1 > mu2  (right tail test)
- Ha:  mu1 < mu2  (left tail test)
- Ha:  mu1 != mu2

## Examples:  `mpg` data


## Error Types


## Power