---
title: "DA 101 Lec 04"
author: "[Your Name Here!]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}
###########################################################################
################# DON'T DELETE THIS PART!!!!!##############################
###########################################################################
                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans"))                                 #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #
### Below is where you read and store your data.                          #
                                                                          #
youShouldRenameThis <- read_csv("death penalty data.csv")
                                                                          #
                                                                          #
                                                                          #
############# ^ ################################# ^ #######################
############# | ## DON'T DELETE THIS PART!!!!!### | #######################
############# | ################################# | #######################
```

## Wakeup Challenge

- Using the built-in `starwars` dataset (be sure to `library(tidyverse)` first), use the `count()` command to tally up gender and eye color (eg, how many starwars females have brown eyes?



```{r}
starwars %>% head()

starwars %>% count(gender, eye_color) %>%
             group_by(gender) %>%
             mutate(percent = n/sum(n))

```



- Use `group_by()` and `mutate` to find out the percentage of each eye color within each gender (eg, what percent of female characters have black eyes?)



## Tidy Data, Revisited

- Meaning of [Tidy Data](https://r4ds.had.co.nz/tidy-data.html) 
- Example:  [Dr. Supp's Data](https://ndownloader.figshare.com/files/2252083)
   -  How to Make it Tidy?
   
   
Question:  notice any problems with this dataset?  Differences between 2013 and 2014?
   

## Tidy Data Worksheet

## `gather()` and `spread()`

Gather makes tables longer, spread makes tables shorter and wider.


### Ex: gathering `table4a` and `table4b`

```{r}
table4a
table4b

table4a %>% gather(`1999`, `2000`, key = "year", value="cases" ) -> tidy4a

```

Cool!  Now, our data is tidy:  each variable has a column (here, we created "year" and every row is a seperate observation).

Now, you tidy up table4b:

```{r}
#note: table 4b contains "population data"

#don't forget to name them!

table4b %>% 
  gather(`1999`, `2000`, key="year", value="population") -> 
  tidy4b



```

One further:  combine the two into one big tidy dataset!

```{r}

tidy4a
tidy4b

#useful new function alert!!! left_join()

left_join(tidy4a, tidy4b) %>% mutate(rate = cases/population)

```


### Ex: spreading `table2`

## In-class Challenges

### Tidy up the Billboard Data



