---
title: "DA 101 Lec 06"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```

## Data Scrubbing Overview

There's no one way to wrangle data; each dataset is unique (and uniquely messy!).  However, there are some "best practices" to follow:

### 1) Need Tidy Data

- Useful functions:  `gather()` and `spread()`


### 2) Variables Have Correct "Type"

- Useful functions:  "as functions":  as.Date(), as.int(), as.numeric(), as.str() etc

Ex:  dates shoul be in date format so that we can do math
Ex:  In the CIRP Data:

```{r}
cirpData <- read_csv("CIRP2014Data.csv")
head(cirpData)

```
See:  sex is wrong type.  It should be a factor!  Use as.factor()


### 3) Values Have Sensible Format

Example:  cirp data should have "male" and "female", not 1 and 2.

- Useful functions:  `recode()` changes "factor levels", and `sub()` is useful to edit formatting.

I.e, `1` = "Male", `2` = "Female.


### 4) Variable Names Are Human-Readable

- Useful functions:  `rename()`

```{r}
cirpData %>% head
cirpData %>% rename(YrEnglish = YRSTUDY1)

# In class, I forgot the exact syntax of the `rename()` command.  Here's how I reminded myself:

help("rename")

```

The cirp data has all kinds of wrangling opportunities.  Changing sex to facor.  Changing factor levels ot male/female.  Changing HSGPA labels to actual grades (A, B, ...) 






## In-Class Challenge

Can you tidy these datasets?  Each dataset gives a diffrent report about departments on several campuses.

```{r}
liberalArtsFunding <- read_csv("Liberal Arts Funding.csv")
liberalArtsGenderRatio <- read_csv("Liberal Arts Gender Ratio.csv")
liberalArtsRanking <- read_csv("Liberal Arts Ranking.csv")
liberalArtsStudents <- read_csv("Liberal Arts Students.csv")
liberalArtsFaculty <- read_csv("Liberal Arts Faculty.csv")
#Funding Table

#gather the data
tidyFunding <- 
  liberalArtsFunding %>% 
  gather(Math, `Poli Sci`, Econ, English,
         key="Department", value="Funding")

#fix format issues
tidyFunding$Funding <- 
  sub("000","",tidyFunding$Funding)

tidyFunding$Funding <- 
  sub("k","",tidyFunding$Funding)


#Gender Ratio Table


tidyGenderRatio <- 
  liberalArtsGenderRatio %>% 
  gather(Math, `Poli Sci`, Econ, English, 
         key="Department", value="Gender Ratio")

tidyGenderRatio$`Gender Ratio` <-
  sub("%", "", tidyGenderRatio$`Gender Ratio`)


#Ranking Table

tidyRanking <- 
  liberalArtsRanking %>% 
  gather(Math, `Poli Sci`, Econ, English, 
         key="Department", value="Ranking")

#clean up!

tidyRanking$Ranking <- 
  sub("th","", tidyRanking$Ranking)

tidyRanking$Ranking <- 
  sub("rd","", tidyRanking$Ranking)

tidyRanking$Ranking <- 
  sub("nd","", tidyRanking$Ranking)



#Students and Faculty Tables

tidyStudents <- 
  liberalArtsStudents %>% 
  gather(Math, Psych, Econ, English, 
         key="Department", value="Students")

tidyFaculty <- 
  liberalArtsFaculty %>% 
  gather(Math, Psych, Econ, English, 
         key="Department", value="Faculty")


#Join The Tables

tidyTable <- tidyFaculty %>% 
  full_join(tidyFunding) %>% 
  full_join(tidyGenderRatio) %>% 
  full_join(tidyRanking) %>%
  full_join(tidyStudents) 

tidyTable

```

## Where are we now?




## Follow-up Questions:

1) How many departments have more than 20000 in funding?  Hint: `filter()`

2) How many departments have more than 20000 in funding and a gender ratio higher than 10%?  Hint:  `filter()`

3) Make a table that contains the total number of students at each school. (`group_by()` and `summarise()`)

4) Add a column that shows student/teacher ratio. (`mutate()`)

5) Sort the table by funding. (`arrange()`)

