---
title: "Lec 03"
author: "[Your Name Here!]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: journal
    df_print: paged
  
---

```{r setup, include=FALSE}
###########################################################################
################# DON'T DELETE THIS PART!!!!!##############################
###########################################################################
                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans"))                                 #
  )                                                                       #
                                                                          #
library(kableExtra)                                                                          #
                                                                          #
### Below is where you read and store your data.                          #
                                                                          #
deathPenaltyData <- read_csv("death penalty data.csv")
                                                                          #
                                                                          #
                                                                          #
############# ^ ################################# ^ #######################
############# | ## DON'T DELETE THIS PART!!!!!### | #######################
############# | ################################# | #######################
```

## Mastering dplyr

First, fix date:

```{r}
as.Date(deathPenaltyData$Date, 
        format="%m/%d/%Y") %>% 
        format("%Y") ->
        deathPenaltyData$Date

deathPenaltyData %>% head 




```


# Goal for today:  practice group_by, summarize, and mutate

```{r}
# Graph state:

qplot(State, data=deathPenaltyData)

#Let's examine the race dist in each state

qplot(State, data=deathPenaltyData, fill=Race)


```

Problem:  we want percent, not totals!  Let's find them using count.

```{r}


deathPenaltyData %>% count(State, Race) %>% head

#compare to original:

deathPenaltyData %>% head


```
## count and mutate

Count() is a combination of group_by and summarize by sum.


```{r}

#let's use mutate to add a proportion column.  New data always goes to end

stateSummary <- deathPenaltyData %>%                    #look at challenge below!
  count(State, Race) %>% 
  group_by(State) %>% 
  mutate(Proportion = n/sum(n))   ##<- woah dude!  important!  sum(n) by category
  

```

Example observation:  43.1% of all people who were executed in AL were black.  

Now, let's put this together in a column graph:

```{r}



qplot(x=State,                                     #challenge second part
      y=Proportion,
      data=stateSummary,
      geom = "col",                 #<--- note, col uses y for height of bar
      fill = Race) + theme(axis.text.x = element_text(angle=-90))





```


Oooh!  Now, instead of looking at "counts", we're looking at proportions -- that way we can compare all of the states, even though some (TX!!) have many, many more.

Observation:  the distribution varies widely from state to state!  Ex:  everyone who received the death penalty in Pensylvania was white, while over half of those executed in NE were black!

Observation:  Asian americans only seem to get the death penalty in CA, NV, OK


In-class challenge:  Try the same thing, but for -method-, not race:

- count up the method in each state  (count)
- find the proprtion using mutate   (mutate)
- plot usin `col` plot                (qplot)


```{r}
# count

deathPenaltyData %>% count(State, Method) %>%
                     group_by(State) %>%              #necessary!  need % in each state!
                     mutate(Prop = n/sum(n)) ->        #note, sum is by state, since group_by
                     methodStateSummary
#plot:

qplot(data=methodStateSummary,
      x=State,
      y=Prop,
      geom = "col",
      fill=Method)+ theme(axis.text.x = element_text(angle=-90))

```

