---
title: "DA 101 Week 3"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: cerulean
    df_print: paged
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```




## Cool example:  Death Penalty Data

First, download and load the csv.

```{r}
deathData <- read.csv("death penalty data.csv")
head(deathData)
```

About the data:  the death penalty was illegal for decades after the civil war.  A supreme cour case in the 70s (1977) legalized it again.  This data is a list of everyone who's received the death penalty since then.

Q:  What about race?  

Idea:  histogram

Can't do histo, not numeric!

Scatterplot doesn't make sense, that's for two numerical variables.



```{r}
#ggplot(deathData, aes(x=Race))+geom_histogram()
```
SInce we're working with categorical data, need a barplot!

```{r}
ggplot(deathData, aes(Race, fill=Race))+geom_bar()
```

Woah!  Does this match up with US demographics?  I think not!  Follow-up question:  
- what are the demographics of the us
- what prorortions are we actually seeing here
- is there a statistically significant difference from population?
- are particular groups disproportionately receiving the death penalty?

Technique:  prop tables

They're easy in R:  use table()

```{r}
table(deathData$Race) %>% prop.table()
```







# Monday

## More examples with death penalty

```{r}

deathData <- read.csv("death penalty data.csv")
head(deathData)

```

```{r}
library(dplyr)
library(ggplot2)

ggplot(deathData, aes(Race))+geom_bar()
```


WHat about a visual that shows a comparison/examination of the relationship of Race and Method (of execution)?

Let's use a stacked barplot.  Set "fill" in aes()


```{r}
ggplot(deathData, aes(Race, fill=Method))+geom_bar(position=position_dodge())
```

## Example:  State

Use a visual to show -where- executions are occuring.

Idea:  use a barplot for the (categorical) State variable!

```{r}
ggplot(deathData, aes(State))+geom_bar()
```

Woah!  TX, what's up?  So many more exectutions!

Moreover, seems like there's just a handful of places where the death penalty is used.

Q:  What if I want to study those states where death penalty is most common?  I could use a filter!

Let's try these states:  AL, FL, GA, MO, OK, TX, VA

How can get a dataset with just these states?  Filter!  We can add many conditions with "or":  |

```{r}
highStates <- filter(deathData, State == "AL" | State == "FL"
                                | State ==  "GA"
                                | State == "MO"
                                | State == "OK"
                                | State == "TX"
                                | State == "VA")

head(highStates)

## did it work?  check with "unique"

unique(highStates$State)


```


We've got it narrowed down.  Let's study those states.  Q:  Method.  In those states, what methods are used?

```{r}
ggplot(highStates, aes(Method))+geom_bar()
```

In these states, where capitol punishmen is used the most, we see that electrocution and Lethal Injection are the only methods used. 

What about Race in those states?  Let's compare both Race and State.  

```{r}
ggplot(highStates, aes(State, fill=Race))+geom_bar(position="fill")
```

Notice here, we can compare ratios/percentages directly!  

Interesting observation:  why are so many (relatively) asian americans receiving the death penalty in OK???  Is there a story there?

## Example: WHen???

Q:  Can we plot time trends in the death penalty?

```{r}
head(deathData)
```

Idea:  barplot?

```{r}
ggplot(deathData, aes(Date))+geom_bar()
```

Bad plot.  Basically:  too many dates, and most only have one execution.  

Problem:  we want the YEAR, not the individual day.  This is an example of "data wrangling".  Our dates aren't in the right format.

```{r}
head(deathData)
```

Problem:  see "<fctr>"?  R thinks the dates are "categorical" variables!  We want numbers.  

R has built-in date functions.  Use:  as.Date()

Problem:  we need to teach it the format.  


```{r}
as.Date(deathData$Date, "%m/%d/%Y")

## We actually want to save this.  Be careful!  Don't destroy your data!  Make a copy.

myDeathData <- deathData  #backup

myDeathData$Date <- as.Date(deathData$Date, "%m/%d/%Y")

head(myDeathData)

# one more step:  format for just the year

myDeathData$Date <-  format(myDeathData$Date, "%Y")

head(myDeathData)

ggplot(myDeathData, aes(Date))+geom_bar()

```

## Here is a cool example table

| Var 1 | Var 2 | Var 3|
| ----- | ------|------|
| thing 1|thing2|thing3|

Example:  mean of iris sepal length

```{r}
mean(iris$Sepal.Length)
```

Instead, you can include it in text.  Use `backticks`  Example: the mean sepal length of iris data is `r mean(iris$Sepal.Length)`








