---
title: "Math 120 Week 03"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: cerulean
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```









# Monday


## Example (Z Scores)

On average, giraffes are 18' tall with std dev 2'.  Elephants have mean height 10' with std dev 0.8'.

If a particular giraffe is 16.5', and a specific elephant is 9.3' tall, which is taller relative to their species?

Answer:  z scores!!  Recall:  a z score for a particular obs is the distance between the observation and the mean, measured in std devs.  I.e., how many std devs above/below the mean are you?  

Giraffe:

```{r}
(16.5-18)/2
```

Elephant:

```{r}
(9.3-10)/.8
```


Both are below the mean (negative z score), but giraffe had bigger z score.  Thus it's taller.

## Empirical Rule (68-95-99.7 Rule)

**IF** our data follows a roughly symmetric, bell-shaped unimodal distribution (normal distribution):

- 68% of the data lies within +/- 1 std dev of the mean (z score btwn -1 and 1).
- 95% of the data lies within +/- 2 std dev of the mean (z score btwn -2 and 2).
- 99.7 of the data lies within +/- 3 std dev of the mean (z btwn -3 and 3).

The empirical rule is a "heuristic"  for predicting "where" we expect data to be.  In what ranges?

## Empirical Example

Human height follows a normal dist.  For men, the mean is 69" and the std dev is 2.7".  About what percent of men are between 66.3" and 74.4"?  

Z scores!

```{r}
(66.3-69)/2.7
```

```{r}
(74.4-69)/2.7
```

```{r}
34+34+13.5
```

About 81.5% of men have height in that range.

About what percent of men are 77.1" tall or less?

```{r}
(77.1-69)/2.7
```


Hard way:  add all to the right of z=3

.15+2.35+13.5+34+34+13.5+2.35 = something

Smart way:  1 - .15 = 99.85% of men are 77.1" tall or below.

## Conclusions about Empirical Rule

Empirical rule lets us make predictions about where we expect data to lie (only in a normal dist).  

Empirical rule lets you have feelings about z scores!  We know how common/uncommon a particular measurement is!

If z score between -2 and 2, that's pretty common (middle 95%).  Z bigger than 2 is "unusual", and z bigger than 3 is crazy! Only 0.15% of the population that range!

## Descriptive statistics

We just spent two weeks learning "descriptive statistics".  I.e., what is data, what kinds of data, how do we view data, and what do we say about data?

Q:  What if I want to know the average systolic blood pressure of **all** 18-22 yo Americans?  

Problem:  it's impractical/impossible to track down all gazillion 18-22 year olds and administer a sphygmomanometer.  Finite time, money, resources.  

In reality, we're forced to work with samples.  We hope they're Representative, i.e., the sample statistic is the similar to the population parameter.  We have to balance these two "levels" of our data:

A **statistic** is a numerical summary of a sample.  I.e., the things we have.

Examples:  
- sample mean is xbar.
- sample std dev is "s"
- sample proportion is "phat"

For each, there's a population version.  A **parameter** is a numerical summary of a population.  We hope to know these, but never will!  Big problem, since sample data is often limited!

Examples:
- pop mean is mu
- population std dev is sigma
- population proportion is "p" (unusual Latin instead of Greek)

We'll never know these parameters, but we hope to say something smart about them based on our statistics (i.e. our sample data).  This process is called **statistical inference**:  making conclusions about population parameters based on sample statistics.

## 4 Sampling Strategies

(Note:  reading in 1.3 of your text)

Note:  the population in question is up to YOU to determine!  Ex:  all Americans, all university students, all men, all 18-22 year olds, etc.  

1) Simple Random Sample (SRS).  This is the best one!  In an SRS, all members of the population are equally likely to be chosen.  Kinda tough!  We need a list (in some order) of every member of the population, then use a random number generator to select our sample.  Usually pretty hard to do!

2) Convenience sample.  Worst kind!  Just sample whoever or whatever is convenient!  Sadly, quite common.  No reason to assume that a convenience sample is representative.  

3) Stratified sample.  Divide and conquer:  break population into strata (groups with commonalities, i.e. gender, age group, income level, education attainment, etc), then do SRS of each one!  Advantage:  get representation from large, diverse populations even with small samples.

4) Cluster sample.  Break population into clusters, randomly select clusters (SRS of the clusters), then sample all individuals in the cluster.  Frequently used over large geographic areas.

Note:  see pictures in text on p26, 28

# Tuesday

## Exercise 1.19 (p30)

- SRS:  Might not  get coverage of all sections.  (bad!)
- Strata are the sections!  Guarantees representation! (good!)
- If sections are cluster, we only choose a sample of them. Some sections get skipped entirely!  (bad)

## Example 1.15 (p28)

- It's hard to get around the densely tropical jungle.  In SRS and Stratified, we're likely to go to many villages.  Here, clusters are more efficient.  Clusters generally are geographic units, ex, city block, village, county.  



## Correlation and Linear Regression

Recall, scatterplots compare two quant variables.  It's an x-y plot.  

Example:  displacement (engine size) vs hwy mpg:

```{r}
plot(mpg$displ, mpg$hwy)
```

Here, looks like as displ gets bigger, fuel efficiency gets smaller.  We want to study this relationship, specifially LINEAR relationships.

## Important qualities of linear relationship

1) Strength.  How close to the "best fit" line are the points?  Above, we see a moderately strong relationsip.

2) Direction.  Up or down?  Pos or neg?  If positive relationship, both variables increase together.  If negative, as x increases, y decreases.  Above, we see a negative relationship.  

## Linear correlation coefficient

We need an objective measure of the "strength" of a linear relationship.  It's called the linear correlation coefficient, "r".  (Not the same as R, our software).

R tells us the strength and direction of a linear relationship.

## Properties of R

- CLose to 1 means strong pos rel
- Close to -1 means strong neg rel
- Close to 0 means no/weak linear relationship (Warning!  Could still be a nonlinear relationship!  Think:  smiley face.)
- R doesn't depend on units (forunate!)  I.e, could switch from lbs to kg, or F to C, or mi to km, doesn't change r! 




# Wednesday

## Finding the equation for the least squares line

Idea: the best fit line is the one that minimizes the square residuals, i.e. the square vertical distance between the points and the line.

Fortunately, there's always only one line that accomplishes this!  

Like all lines, there are two parts:  slope and intercept.

In high school, you learned 

- y = mx + b

Here, we use different letters for the slope and intercept, but the meaning is the same:

- yhat = b0 + b1x

where

- b0 = y-intercept
- b1 = slope

They're easy to find:

1)  Find the slope.  The equation is

$$b_1 = r\cdot \frac{s_y}{s_x}$$

Two notes:

- Similar to change in y over change in x.  Makes sense!
- Stronger correlation means steeper slope!  I.e., stronger correlation means that x has a more dramatic effect on y.

2) Find intercept.  We use the **very cool fact** that the least-squares line always touches the point $(\bar{x}, \bar{y})$ :

$$ b_0 = \bar{y} - b_1 \cdot \bar{x} $$

## Example

Suppose xbar = 1, ybar = 2, sx = 3,  sy = 6, and r = 1/2.  

Then, the equation of the line:

b1 = (1/2)*6/3 = 1

b0 = 2 - 1*1 = 1

Thus yhat = 1 + x




















