---
title: "Math 120 Week 02"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: cerulean
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```












# Tuesday

Goals:

- Center, Shape, and Spread for Quantitative Variables
- Outliers, boxplots

Measures of center:  tell what's "normal".  

Definition:  a measure is *robust* if it's not strongly affected by outliers/skew.

Example: median is robust (not affected by outliers), but mean is *NOT* robust.  Example: 1, 1, 1, 1, 1, 19.  

## Shape of a distribution

Common shapes:

- Symmetric.  Evenly distributed on either side of the middle.
- Left skew (left tail)
- Right skew (right tail)

### Exercise:  What shape?

Based on the following, what shape does the variable have and how do the mean and median compare?

1)  Variable:  Exam scores in a large class for an easy exam.  Shape:  left skew, most students scored well, only a few scored poorly.  We'd expect the mean score to lower than the median.  

2)  Variable:  salaries in a large corporation.  Strong right skew!  Most employees earn a "normal" amount, a select few earn much more.  Ex:  In 2018, revealed that the CEO of Activision earned 300x the average salary!  The mean salary will be a lot larger, could skew our perception of wealth in the corporation.

3)  Variable:  height of adult men in the US.  Symmetric dist!  Most people have "average" height, only a few are exceptionally tall or short.  We'd expect that the mean and median are similar.

4)  Variable:  heights of all adults in the US, men and women.  Shape:   bimodal.  Still fairly symmetric, since 50/50 gender.  

## Measures of Spread (quantitative variable)

I.e., how much variability?  How far away do we expect observations to be from the center?

Three big measures:

1) Range:  max - min

Example:  range of sepal lengths:

```{r}
summary(iris$Sepal.Length)
```

Range:  3.6cm.  Advantage:  easy!  Big disadvantage:  super, duper, extra not robust!!!  Even one outlier wrecks the range.  

2) Standard deviation.  The "deviation" for an observation is just it's distance from the mean:  x - xbar.  For std deviation, we find all deviations for all observations, and (roughly) take their average.  Std dev = average distance from the center. 

Formula:

$$s=\sqrt{\sum\frac{(x-\bar{x})^2}{n-1}}$$

3) Interquartile Range = The distance between the quartiles Q1 and Q3.  Note:  Q1 is the median of the lower half.  Q3 is the median of the upper half). IQR gives us the range/spread of the middle 50% of the data.  

Interesting note:  all three of these are totally different approaches, but they're all reasonable ways to measure spread.  

Q:  Which of the three measures of spread is "robust"?  
A: Only the IQR, since it's determined by the middle stuff.  Outliers have no effect.  

## Outliers

Outliers are "unusual" measurements!  Couple of different methods for determining.  Here's ours:  

We'll define an outlier to be any measurement that's more than 1.5 IQRs away from Q1 or Q3.

Low cutoff:  $$Q1 - 1.5\cdot IQR$$

high cutoff: $$Q3 + 1.5\cdot IQR$$

**Example**  Verify that 20 is an outlier in 1,2,3,4,5,20.

## Boxplots

Boxplots visualize the IQR.  Super useful for comparing multiple quant. variables, since it's easy to have multiple boxplots on the same axes.

Example:  Iris dataset

```{r}
boxplot(iris$Sepal.Length)
```


Comparing all variables:
```{r}
boxplot(iris)
```

Comparing by species:
```{r}
boxplot(data=iris, Sepal.Length~Species)
```



# Wednesday

## Goals

- Weird Plots
- Empirical Rule
- Meaning of "Statistical Inference"

## Weird plots

We've seen some "typical" plots.  Example, iris data:

```{r}
boxplot(data=iris, Sepal.Length~Species)
```

Observations:  Q1 for versicolor about 5.7ish.  

Q:  which species has the highest center?  Virginica!

Q:  which species has the smallest spread? Setosa sepal lengths have least variability.  

Q:  Outliers?  Looks like virginica has one low outlier.  

## Weird example: Cirp Survey Data

Take a look:

```{r}
cirpData <- read.csv("CIRP_DATA_2014_CLEAN.csv")
head(cirpData)

```

Note:  totally need metadata.  

Q:  what about english?  How much english have DU students had?

```{r}
boxplot(cirpData$YRSTUDY1)
```

Analysis:  almost everyone responded 6, i.e., the vast majority of incoming DU students took 4 years of english in HS.  Q1 = M = Q3 = 6.


## Another weird example:  scatterplot

Q: Is there a relationship between RATE01 (academic ability) and RATE03 (competitiveness)?  

```{r}
qplot(cirpData$RATE01, cirpData$RATE03)
```

Woah!  weird!  What's going on?  Discrete data often doesn't work well with scatterplots.  Can't tell how many points!

Neat trick:  jitter.  "Jittering" means adding a small amount of random noise.

```{r}
qplot(cirpData$RATE01, cirpData$RATE03)+geom_jitter()

```
Ah!  Now we can see a roughly positive pattern.  

Moral of the story:  think carefully about any visualition you do.  What's it mean?



## Example:  Exam scores

66, 67, 71, 71, 74, 76, 76, 78, 80, 99

Check for outliers, make a boxplot, describe shape.

IQR: 7, Q1 = 71, Q3 = 78

Low cutoff:  


```{r}
71 - 1.5*7
```

High cutoff:  

```{r}
78+1.5*7
```

Note:  99 is an outlier!

Warning for boxplot:  draw axis first!

Shape:  aside from outlier, there's a little left skew.

```{r}
boxplot(c(66, 67, 71, 71, 74, 76, 76, 78, 80, 99), horizontal = TRUE)

```


## Empirical Rule and Z scores

We know about outliers, but what about everything else?  Common question:  how "unusual" is a measurement in our data?

Major tool:  z scores (standard scores).  

Z scores measure the distance from a point to the mean, in units of standard deviation!  I.e., how many standard deviations above or below the mean are you?

Example:  Adult women in the US have mean height 64" with std dev 2.4"  (true).  Prof miller 75".  Compute her z score!

```{r}
(75-64)/2.4
```

Note about rounding:  three decimals is enough.

Why is this useful?  It lets us compare elements of different datasets!  It accounts for the center and spread of the respective distributions.

Example:  Exam scores.  On ACT, the mean score for test takers is 24, std dev 3.  On SAT, the mean score is 680, with std dev 40.

If Alex got a 28 on the ACT, and Beth got 720 on the SAT, who scored better?

Z scores!

Alex:

```{r}
(28- 24)/3
```

Beth:

```{r}
(720-680)/40
```















