---
title: "DA 101 Lec 21"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```


## Chi Square Goodness of Fit Test

So far, we've seen hyp tests for one and two proportions.  We can answer question like:

Is there evidence that more than half in the population belong to a category?

Is there evidence of a difference in proprtion between two populations?


Chi square goodness of fit test expands this conversation:  what if we want to know about an entire distribution for all categories, not just one in particular?

Example:  The M&M manufacturer (Mars Co) claims a specific distribution for the colors of M&Ms.  If we take a particular bag as a sample, is there evidence that the distribution doesn't conform?  Many categories: 

- blue
- brown
- yellow
- orange
- green

### Classic example:  Baby Toys

An investigator wonders:  do babies show a preference for a certain color of toy?  The experiment:  100 babies are presented with 4 identical toys that only differ in color:

- blue
- yellow
- green
- red

We record which toy they reach for first (indicating their preference).  Results:

- 21 choose blue
- 15 choose yellow
- 34 choose red
- 30 choose green 


Clearly, the results deviate a bit from our expectation (25 each).  Is this deviation big enough/extreme enough to conclude, with statistical significance, that the proportions differ for each color.


#### Hypotheses

H0:  p_blue = .25, p_yellow=.25, p_green = .25, p_red = .25
Ha:  at least one of the above differs


#### Test Stat

chisq = sum( (O-E)^2/E )  [Note: similar to std dev]

stdev = sum ( (x - xbar)^2 )/(n-1)

Here:

chisq = (21-25)^2/25 + (15-25)^2/25+(34-25)^2/25+(30-25)^2/25

```{r}
(21-25)^2/25 + (15-25)^2/25+(34-25)^2/25+(30-25)^2/25
```

#### p-val

The chi square test stat doesn't follow a normal dist!  (Std dev must always be positive).  It has its own dist:  chi square dist!  It's like a distro for std dev.

In r: pchisq

```{r}
1-pchisq(8.88, 4-1)  # df = categories - 1
```

Note:  1- is for right tail, since cdfs always give left

Here, small p-val means reject H0, there's strong evidence that babies exhibit some color preference (i.e. not all proprtions are equal).

### Death Penalty Data

Read it:

```{r}
deathData <- read_csv("death penalty data.csv")
head(deathData)
```


Q: Race is a variable.  What's up?

```{r}
deathData %>% ggplot(aes(Race))+geom_bar()
```

Q:  does this distro follow the race demographics of the USA, -or-, is there evidence that the proportion of people receiving the death penalty differs from the national demographics.

According to the internet:  66% of americans are white, 13% are black, 5% are asian, 0.5% are NA, and 15% are latino, and .5% other.

Observed totals:

```{r}
table(deathData$Race)
```

```{r}
# store it:
# asian, black, latino, na, other, white
totals <- c(6, 495, 120, 16, 2, 803)

props <- c(.05, .13, .15, .005, .005, .66)

chisq.test(totals, p = props)
```

