---
title: "DA 101 Lec 07"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```

## From Last Time:

```{r}
liberalArtsFunding <- read_csv("Liberal Arts Funding.csv")
liberalArtsGenderRatio <- read_csv("Liberal Arts Gender Ratio.csv")
liberalArtsRanking <- read_csv("Liberal Arts Ranking.csv")
liberalArtsStudents <- read_csv("Liberal Arts Students.csv")
liberalArtsFaculty <- read_csv("Liberal Arts Faculty.csv")
#Funding Table

#gather the data
tidyFunding <- 
  liberalArtsFunding %>% 
  gather(Math, `Poli Sci`, Econ, English,
         key="Department", value="Funding")

#fix format issues
tidyFunding$Funding <- 
  sub("000","",tidyFunding$Funding)

tidyFunding$Funding <- 
  sub("k","",tidyFunding$Funding)


#Gender Ratio Table


tidyGenderRatio <- 
  liberalArtsGenderRatio %>% 
  gather(Math, `Poli Sci`, Econ, English, 
         key="Department", value="Gender Ratio")

tidyGenderRatio$`Gender Ratio` <-
  sub("%", "", tidyGenderRatio$`Gender Ratio`)


#Ranking Table

tidyRanking <- 
  liberalArtsRanking %>% 
  gather(Math, `Poli Sci`, Econ, English, 
         key="Department", value="Ranking")

#clean up!

tidyRanking$Ranking <- 
  sub("th","", tidyRanking$Ranking)

tidyRanking$Ranking <- 
  sub("rd","", tidyRanking$Ranking)

tidyRanking$Ranking <- 
  sub("nd","", tidyRanking$Ranking)



#Students and Faculty Tables

tidyStudents <- 
  liberalArtsStudents %>% 
  gather(Math, Psych, Econ, English, 
         key="Department", value="Students")

tidyFaculty <- 
  liberalArtsFaculty %>% 
  gather(Math, Psych, Econ, English, 
         key="Department", value="Faculty")


#Join The Tables

tidyTable <- tidyFaculty %>% 
  full_join(tidyFunding) %>% 
  full_join(tidyGenderRatio) %>% 
  full_join(tidyRanking) %>%
  full_join(tidyStudents) 

tidyTable




```

## Where are we now?

Hey, why are there nas?  Well, let's look at the data

```{r}
liberalArtsFaculty
liberalArtsRanking
```
Observation: the full join uses **all** data from all tables.  If there's not a match in one table, then those entries are filled with NA.

Challenge:  instead of full_join, try inner_join.  What happens?

```{r}
tidyFaculty %>% 
  inner_join(tidyFunding) %>% 
  inner_join(tidyGenderRatio) %>% 
  inner_join(tidyRanking) %>%
  inner_join(tidyStudents) 
```



Inner join only takes data that's in **Both** tables.  I.e., has to be in all of them.

Note:  there's also,

- `left_join` uses everything in the first dataset, only what matches from the second.  Useful for adding on data that matchs.


```{r}
left_join(tidyFaculty, tidyRanking)
```

This result has all data from tidyfaculty, and sticks NAs in for rankins where missing.

Cool observation: 

- Full join gave us 25 rows (all)
- Inner join gave us 15 rows (only those that match)
- Left_join gave us 20 rows:  all in the left, only matching on the right.

## Follow-up Questions:


```{r}
tidyTable
```



#### 1) How many departments have more than 20000 in funding?  Hint: `filter()`


```{r}
tidyTable %>% filter(Funding > 20) %>% nrow
```


Woah!  Notice that funding is still a character type!

Fix:

```{r}
tidyTable %>% filter(as.integer(Funding) > 20) %>% nrow
```


Wooooah!  Using correct data types is super important!  R tried to guess how the character works, but it guessed wrong!!


#### 2) How many departments have more than 20000 in funding and a gender ratio higher than 10%?  Hint:  `filter()`

```{r}
tidyTable %>% filter(as.integer(Funding) > 20  & as.integer(`Gender Ratio`)>10 )
```



#### 3) Make a table that contains the total number of students at each school. (`group_by()` and `summarise()`)

```{r}
tidyTable %>% group_by(School) %>% summarize(StudentTotal = sum(Students, na.rm = TRUE))
```

Challenge:  same, but average number of students:


```{r}
tidyTable %>% group_by(School) %>% summarize(StudentMean= mean(Students, na.rm = TRUE))

```


#### 4) Add a column that shows student/teacher ratio. (`mutate()`)

```{r}

tidyTable %>% mutate(studentTeacherRatio = Students/Faculty)

```


Challenge:  Find average student teacher ratio at the schools (by department)

```{r}
tidyTable %>% group_by(Department) %>% mutate(AvgRatio = mean(Students/Faculty, na.rm = TRUE))
```




#### 5) Sort the table by funding. (`arrange()`)


```{r}
help(arrange)


arrange(tidyTable, Funding)


tidyTable %>% arrange(Funding)

```

