---
title: "DA 101 Lec 20"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```


## Hyp tests for categorical data

Our tests so far, like t test and cor test are about numerical data.

T tests are about means (xbar), and cor tests are about correlation coeffs (r).  

What if we want to study a categorical variable?  Ex:  hair color, year in school, nationality, political party, etc.  You can't find the "average" of a category, so what then?

What summary stat would you use for categories?  Count/freq is an obvious one, but it also has an obvious limitation:  we need to know count relative to sample size:  percent!!! I.e., "proportion".

### Hyp test for proprtions

Recall, CLT:  tells us that our sample statistic follow normal distributions as long as n is large.  (More on that later).   For t-test, we need n at least 30  (made up, but useful!). 

All hyp tests follow the same four basic steps:  

- hypotheses
- test stat
- p-val
- conclusion

Here, our hypotheses are about proportions.  We need a test stat, I.e., a z-score (standard score).  

Recall:  z = (obs - expected)/(std dev)

For numerical data:  z = (xbar - mu)/sigma

For sample proportions, phat, the CLT tells us the expected proportion is p, the population proportion.

The std dev for phat is sqrt(p(1-p)/n)

We need the center and spread to use the normal dist.

Example:  Prof Miller suspects that more than half of residents in Columbus are registered Democrats.  (note:  proportion 1/2)  To test this, she takes a sample 100 voters in columbus, and finds that 59 of them are Dem.  Has she found statistically significant evicence to support her claim?

1)  Hypotheses

My claim:  the proprtion of dem is greater than 1/2

H0:  p=1/2  (always a statement of = about pop)
Ha:  p>1/2

2) Observed:  phat = 59/100
   Expected:  p = 1/2
   Std Dev:   sqrt((1/2)*(1-1/2)/100)
   
   To the z score!
   
```{r}
# (obs - exp)/std dev

(59/100-1/2)/sqrt(1/2*(1-1/2)/100)

```
   
Test stat:  z = 1.8.

3)  P-value.  The p-val is the probability of obvserving a sample proportion that's as big or bigger than the one we got (59/100), assuming that the null hypothesis is true (p= 1/2).


ClT says this follows a normal dist.  R (and every other software ever) has a built-in normal calculator:  `pnorm`:

```{r}
pnorm(1.8)
```

The above is the area to the left of z=1.8, but we need tail area.  
```{r}
#  p = 1-.964

1-.964

```

4)  Since p<.05, we reject H0.  There's strong evidence to suggest that the proportion of Dem voters in columbus really is greater than 1/2.  



Your turn!

To pass a change to the faculty handbook, a 2/3 majority is needed.  Prof Miller has a change she'd like to propose, but she wants to make sure that it's supported beforehand.  To investigate, she takes a sample of 60 faculty members, and finds that 44 of them support her proposal.

Is there statistically significant evidence to support her claim?

H0:  p=2/3
Ha:  p>2/3

z = (phat - p)/sqrt(p*(1-p)/n)

```{r}
(44/60 - 2/3)/sqrt(2/3*(1-2/3)/60)
```

Use pnorm for p-val:

```{r}
1-pnorm(1.095445)
```





```{r}
(44/60-2/3)/sqrt((2/3)*(1-2/3)/60)

1-pnorm(1.0945)
```



## More about Categorical Inference (Fri)


Last time:  Z test for p  (population proportion)

I.e., does our sample provide statistically significant evidence to support our claim?

For a practical example, let's look at the mpg dataset:

```{r}
head(mpg)
```


Here, let's investigate "class".  What are the classes in this dataset, ie, what are the "levels" of the class variable?

Let's use `unique`:

```{r}
unique(mpg$class)
```


According to literature distributed by GM, 20% of cars sold are midsize cars.  But, we suspect that actually fewer than 20% are midsize.

Question:  according to our data (assuming that mpg dataset is a random sample of cars sold in the US), what proportion of cars were midsize?

```{r}
mpg$class
```

Q: how can we count up the number of midsize?

A: we check equality, and sum.

```{r}
sum(mpg$class == "midsize")
```

```{r}
count(mpg)
nrow(mpg)
```

What prop are midsize?

```{r}
41/234
```

In this sample, certainly the proportion is less than 20% (our claim that we're investigating).  But how convincing is this really?  We know this is just a sample!  Not the whole population?  How strong is this evidence?

CLT saves the day, we know the probability distro!  So let's find out how likely this is!  To the hyp test!

Our claim:  the proportion of midsize is less than 20%

H0: p = .2   (needed to find p-val!  used in z-score and p-val)
Ha: p < .2   (our claim, also our notion of "extreme"; here, extreme means "small")

test stat:

```{r}
#  z = (phat - p)/sqrt(p(1-p)/n)

(41/234 - .2)/sqrt(.2*(1-.2)/234)

```

p-val:  Here our p-val is the probabity of ovbserving a sample proportion that's 41/234 or smaller, assuming that the true proportion is 0.2.

We find it with pnorm:

```{r}
pnorm(-0.94789)
```

Conclusion:  since our p-val is >.05, we fail to reject H0.  We did not find the evidence we're looking for to support our claim that the the proportion of midsize cars is less than 20%.

### 2 samp z tests for p

What if we want to compare two different groups?  Example, for mpg, what if we to compare midsize and compact cars?  Is there evidence of a difference in proportion?  Let think of midsize as population 1 and compact cars as population 2.  Is there evidence of a difference in the proprtions sold?  

Is p1 != p2 ??

(Note, this is very similar to your 2-t tests for means from different populations)

The test works the same as all hyp tests:  

1) hyps
2) test stat
3) p-val
4) conclusion

The only difference here is our test stat:  we need to compare the difference between the two proportions!

The test stat here is:

z =  (phat1 - phat2)/sqrt(phat1(1-phat1)/n1 + phat2(1-phat2)/n2)

We already know that phat1 = 41/234 (proportion of midsize).

Need: proportion of compacts.

```{r}
sum(mpg$class == "compact")
```

So, phat2 = 47/234

The test:

H0: p1=p2
Ha: p1 != p2

```{r}
phat1 <- 41/234
phat2 <- 47/234
n1 <- 234
n2 <- 234

(phat1 - phat2)/sqrt(phat1*(1-phat1)/n1 + phat2*(1-phat2)/n2)

```

p-val:  Our p-val is the probability of observing a difference in sample proportion, phat1-phat2, as big or bigger that what we observed, assuming that the difference in population proportion is zero.

```{r}
pnorm(-.71)
```

Here, there's a 23.8% chance of observing a difference as big or bigger than the one we got.  I.e., the null hyp seems quite reasonable.  We fail to reject H0, and thus we didn't find the evidence we're looking for to suggest that there's a difference in the proportion of midsize and compact cars.  