---
title: "DA 101 Lec 08"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```

##  Hyp Tests and Significance: and intro

*[Brief powerpoint in-class, available on Google Drive shared folder.]*

In our reading about Breakfast studies, we learned that p-values and hypothesis tests are often misused.  A big problem:  software like R makes it super easy to perform the tests, but that doesn't mean you know what you're doing!

A big goal of mine:  to help take the "magic" out of hypothesis tests so that students (you!) can use them appropriately and insight fully.

## Case studies for hyp tests

Let's analyze some examples in-depth, using datasets we're familiar with.

### iris dataset

Recall iris:

```{r}
iris %>% head
```

Question:  is there evidence of a difference in mean sepal length between the setosa and the virginica plants?  

First, we need to subset (filter):

```{r}
#make subsets for each species

setosaData <- iris %>% 
                filter(Species=="setosa")

virginicaData <- iris %>% 
                  filter(Species =="virginica")

versicolorData <- iris %>%
                   filter(Species == "versicolor")


```

We have the subsets, now let's test!

The command is called `t.test()`.  

```{r}

#perform a two-t test

t.test(setosaData$Sepal.Length, virginicaData$Sepal.Length)

```

Let's interpret this:

- Type of test:  2-samp t test (comparing two population means)
- Two tail test
- H_0: $ \mu_1 = \mu_2$  (I.e., we assume the means are the same for both species) (note: always true for 2-samp t test)
- H_a:  mu_1 != mu_2

- t test statistic:  summary of the data.  t = -15.386.  This is a summary of our data, more on this later.

- our p-value:  2.2e-16 = .00000000000000022  (i.e., almost zero!)  Here's what this means:  Assuming that there's no difference in population mean between setosas and virginicas, there's a .0000000000000000022 chance of observing a sample difference that's a big as the one we saw.  I.e., our sample is **almost impossible** if the null hyp is true.

- Since p is so small, it disagrees with the null hyp.  We trust our data of course, and thus we REJECT the null hypothesis!  

- Yaaay!  We found evidence to support our claim.  There's strong evidence of a difference in mean sepal length between setosa and virginica.

- The end!








### mpg dataset


### garlic mustard dataset

In-class challenge:  perform a 2-samp t test on the variables you chose for lab 3 to investigate if there's a difference between north american plants and european.

```{r}
garlicData <- read_csv("GarlicMustardData.csv")


eurData <- garlicData %>% filter(Region == "Europe")

naData <- garlicData %>% filter(Region=="NorthAm")


# test for diff in bio1

t.test(eurData$bio1, naData$bio1)


```

Interpret:

hypotheses:

h0:  mu_1 = mu_2, i.e., the mean temp in na and eur is the same.
ha:  mu_1 != mu_2, i.e., there's a difference in mean temp

t = -5.385  (more about t stats in the future)

p = .0000001239

Since p<.05, we reject the null hypothesis.  There's strong evidence of a difference in mean temp between eur and na (probably not too surprising, but here's the mathematical evidence).

