---
title: "DA 101 Lec 05"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}
###########################################################################
################# DON'T DELETE THIS PART!!!!!##############################
###########################################################################
                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans"))                                 #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #
### Below is where you read and store your data.                          #
                                                                          #
#youShouldRenameThis <- read_csv("death penalty data.csv")
                                                                          #
                                                                          #
                                                                          #
############# ^ ################################# ^ #######################
############# | ## DON'T DELETE THIS PART!!!!!### | #######################
############# | ################################# | #######################
```

## Wake-up Challenge: Tidy up the Billboard Data using gather()


```{r}
billboard <- read_csv("billboard.csv")


billboard %>% head


billboard %>% gather(x1st.week, x2nd.week, x3rd.week, x4th.week, key = "week", value="rank")


```




#### Ex: spreading table2

See text, page chapter 12.

## Data Wrangling


When loading with read_csv, the command will try to *guess* what data types.

```{r}
death <- read_csv("death penalty data.csv")

death %>% head
```



### Understanding class

```{r}
class(death$Age)
```

We see that R understands that Age is numeric.  

Note, it didn't understand the date!  We'd have to teach it what the date format is!

Whenever you load into R (read_csv), always check what data types you have!  Things can go wrong!


### Setting data types:

It's often useful for us to manually set data types in R.  Here are a few examples where it's needed.

### as.factor,

This function changes data type to "factor", which is what we use for categorical variables.

#### Example: (unclean) CIRP data , HSGPA by SEX


```{r}
cirpData <- read_csv("CIRP2014Data.csv")


cirpData %>% head
```


Example:  graph HSGPA by SEX

```{r}

ggplot(cirpData, aes(HSGPA, fill=SEX)) + geom_bar()

```

Note, the fill didn't work?  Why not?  R thinks that gender is numerical! 1 and 2!

Fix:

```{r}

# as.factor turns it into a factor!

ggplot(cirpData, aes(HSGPA, fill= as.factor(SEX) )) + geom_bar()

cirpData %>% head

# permanently change SEX type:

as.factor(cirpData$SEX) -> cirpData$SEX

## equal to this:

cirpData$SEX <- as.factor(cirpData$SEX)


class(cirpData$SEX)



```

Question:  we've got 1s and 2s for gender, but this doesn't make much sense.  How to fix?

Answer:  `recode()`.  See below!





### as.numeric, as.double, as.integer

We can change lots of data types, see above!  These are the least commonly used, since R is good at identifying numbers.


### as.Date

Dates are important!  But, R almost never guesses the right format.

#### Example: Lobbyist Data

Read the data:

```{r}
lobbyData <- read_csv("lobbyist-data-gifts.csv")


lobbyData %>% head

#question:  did R get teh date right?

class(lobbyData$PERIOD_START)

#hmm, maybe doesn't know.  let's use as.Date:

lobbyData$PERIOD_START %>% as.Date(format = "%Y-%m-%d" ) %>% head


#it worked!  But don't forget to save:

lobbyData$PERIOD_START %>% 
  as.Date(format = "%Y-%m-%d" ) -> 
  lobbyData$PERIOD_START

#below works equally well:

lobbyData$PERIOD_START <- 
  lobbyData$PERIOD_START %>% 
  as.Date(format = "%Y-%m-%d" )


lobbyData %>% head

```



### Renaming factor levels with recode

The `recode()` function in dplyr let's us adjust our factor levels.  The "levels" are the possible options for the category; for example in the CIRP data, the levels for SEX are "male" and "female".

#### Example: (unclean) CIRP data

```{r}
cirpData %>% head

#pipe it into recode

cirpData$SEX %>% recode(`1`= "Male", `2`="Female") ->
  cirpData$SEX

#Now, labels in graph are fixed:

cirpData %>% ggplot(aes(HSGPA, fill=SEX))+geom_bar()

```


## Joining Tables



### Example: Lahman baseball scores database