---
title: "DA 101 Lec 15"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```

## Lin Reg Group Practice -- House Sales

```{r}
houseSales <- read_csv("House-Sales.csv")

houseSales %>% head

houseSales %>% nrow

```


The game:  I'll give each group a variable.  You guys will make a linear model to predict PRICE based upon this model.  Based upon YOUR model:

- interpret the coefficients (slope, intercept)
- interpret r^2
- interpret the residual plot

Group 1:  Bedrooms

```{r}
plot(houseSales$bedrooms, houseSales$price)
priceMod <- lm(houseSales$price~houseSales$bedrooms)
summary(priceMod)
```

Interpret slope:  If the number of bed increases by 1, we expect the price of the house to increase by 121,790.

R^2:  About 9.5% of the variation in price is due to the linear rel with beds.  Of course, bedrooms aren't the only factor.  But, of all the variation in price, bedrooms account for 9.5%.  

```{r}
plot(houseSales$bedrooms, priceMod$residuals)
```

These resids are not homoskedastic.  Low variance at the ends, and high variance in the middle.  Another way:  hereroskedastic.  Bad.


Group 2: Bathrooms

Group 3: sqft_living

```{r}
plot(houseSales$sqft_living, houseSales$price)
priceMod <- lm(houseSales$price~houseSales$sqft_living)
summary(priceMod)
```

Slope in general:  if x increase by 1 unit, we expect y to change by the slope.

Here:  x = sqft_living, y = price.  If a house's living sqft increases by 1 square ft, we expect the price to increase by $280.807.

R^2:  About 49% of the variation in price is due to effect of the relationship with sqft.  We know sqft can't be the only thing, but it's a big one!  It accounts for about half of the price.  makes sense!  Bigger houses tend to be more expensive.

```{r}
plot(houseSales$sqft_living, priceMod$residuals)
```

This is a textbook example of non-homoskedasticity.  We clearly see variance increasing as size increases.  Takeaways:

-  We're seeing a pattern in the residuals, suggesting the linear model isn't the best or most appropriate.  
- For small houses, the model does a good job (small residuals).  For very large houses, our model has lots of error.  Makes sense -- very large houses are unusual!  Don't fit the pattern!

## Next Topic:  Multilinear models

What if we want to consider more than one variable?  Not just sqft, not just bathrooms, not just bedrooms, etc.  How can we see the effect of all of them?

```{r}
#predict price  (y = price)
multiModel <- lm(price~bedrooms+bathrooms+sqft_living+improvements+yr_built, data=houseSales)

summary(multiModel)
```

Cool!  R^2 is bigger!  Did more variables help?  To be continued...