---
title: "Math 120 Week 03"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: cerulean
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```









# Monday


## Example (Z Scores)

On average, giraffes are 18' tall with std dev 2'.  Elephants have mean height 10' with std dev 0.8'.

If a particular giraffe is 16.5', and a specific elephant is 9.3' tall, which is taller relative to their species?

Answer:  z scores!!  Recall:  a z score for a particular obs is the distance between the observation and the mean, measured in std devs.  I.e., how many std devs above/below the mean are you?  

Giraffe:

```{r}
(16.5-18)/2
```

Elephant:

```{r}
(9.3-10)/.8
```


Both are below the mean (negative z score), but giraffe had bigger z score.  Thus it's taller.

## Empirical Rule (68-95-99.7 Rule)

**IF** our data follows a roughly symmetric, bell-shaped unimodal distribution (normal distribution):

- 68% of the data lies within +/- 1 std dev of the mean (z score btwn -1 and 1).
- 95% of the data lies within +/- 2 std dev of the mean (z score btwn -2 and 2).
- 99.7 of the data lies within +/- 3 std dev of the mean (z btwn -3 and 3).

The empirical rule is a "heuristic"  for predicting "where" we expect data to be.  In what ranges?

## Empirical Example

Human height follows a normal dist.  For men, the mean is 69" and the std dev is 2.7".  About what percent of men are between 66.3" and 74.4"?  

Z scores!

```{r}
(66.3-69)/2.7
```

```{r}
(74.4-69)/2.7
```

```{r}
34+34+13.5
```

About 81.5% of men have height in that range.

About what percent of men are 77.1" tall or less?

```{r}
(77.1-69)/2.7
```


Hard way:  add all to the right of z=3

.15+2.35+13.5+34+34+13.5+2.35 = something

Smart way:  1 - .15 = 99.85% of men are 77.1" tall or below.

## Conclusions about Empirical Rule

Empirical rule lets us make predictions about where we expect data to lie (only in a normal dist).  

Empirical rule lets you have feelings about z scores!  We know how common/uncommon a particular measurement is!

If z score between -2 and 2, that's pretty common (middle 95%).  Z bigger than 2 is "unusual", and z bigger than 3 is crazy! Only 0.15% of the population that range!

## Descriptive statistics

We just spent two weeks learning "descriptive statistics".  I.e., what is data, what kinds of data, how do we view data, and what do we say about data?

Q:  What if I want to know the average systolic blood pressure of **all** 18-22 yo Americans?  

Problem:  it's impractical/impossible to track down all gazillion 18-22 year olds and administer a sphygmomanometer.  Finite time, money, resources.  

In reality, we're forced to work with samples.  We hope they're Representative, i.e., the sample statistic is the similar to the population parameter.  We have to balance these two "levels" of our data:

A **statistic** is a numerical summary of a sample.  I.e., the things we have.

Examples:  
- sample mean is xbar.
- sample std dev is "s"
- sample proportion is "phat"

For each, there's a population version.  A **parameter** is a numerical summary of a population.  We hope to know these, but never will!  Big problem, since sample data is often limited!

Examples:
- pop mean is mu
- population std dev is sigma
- population proportion is "p" (unusual Latin instead of Greek)

We'll never know these parameters, but we hope to say something smart about them based on our statistics (i.e. our sample data).  This process is called **statistical inference**:  making conclusions about population parameters based on sample statistics.

## 4 Sampling Strategies

(Note:  reading in 1.3 of your text)

Note:  the population in question is up to YOU to determine!  Ex:  all Americans, all university students, all men, all 18-22 year olds, etc.  

1) Simple Random Sample (SRS).  This is the best one!  In an SRS, all members of the population are equally likely to be chosen.  We need a list (in some order) of every member of the population, then use a random number generator to select our sample.  Usually pretty hard to do!

2) Convenience sample.  Worst kind!  Just sample whoever or whatever is convenient!  Sadly, quite common.  No reason to assume that a convenience sample is representative.  

3) Stratified sample.  
















