---
title: "Math 120 Lec 21"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```







## Hyp Tests

Hypothesis tests are just like criminal trials.

In a criminal trial, we wonder:  innocent or guilty?

In a hyp test, the NULL HYPOTHSIS (H0) is on trial.  The null hyp is a statement about the population.  The alternate hypothesis is logical opposite.  (Ha).

In a criminal trial, we assume innocent until proven guilty.  In a hypothesis test, we assume H0 is true, until we find evidence that's it's false.

The whole reason we're doing the test is becasue we susppect H0 is false!  I.e., the alt hyp (Ha) is our claim about the population.

In a criminal trial, we have witnesses.  In a hyp test, we have sample data!

In a criminal trial, there are laws and lawers and arguments.  In a hyp test, there's probability!  (p-value)

A p-value is the probability of observing a sample result as or more extreme than the one we got, assuming H0 is true.  The assumption is important: we can't compute probabilities unless we make an assumption about the population!

In a criminal trial, a judge or jury weighs the strength of the evidence.  In a hyp test, we decide if our p-value is SMALL ENOUGH.  

If p is small, it means our sample result is unlikely under H0.  In this case, we REJECT H0, we trust our data!  Small p-val -> reject H0.  Support Ha.

Every hyp tests ends in one of two ways:  if p is small, we reject H0.  If p is large, we fail to reject H0.  


Wait, how small is "small"?   Totally up to us!  It's called the "significance level", "alpha".  Generally, we use alpha = .05 by default.  In other words, if our sample has less than 5% chance of occuring, we reject H0.

Example) In Fall 2015, my stats class had 16 women out of 24 students.  Based upon this, I suspect that the gender ratio at DU was higher than 50% female.  Does this sample result support my claim?  Perform a hyp test.  Use alpha=.05.

First, need hyptheses.  Both H0 and Ha.

My claim:  the proportion of female students is >50%.  I assume that this isn't true, and look for evidence to the contrary.

### Step 1

H0: p<= .5  (really, we think p=.5)
Ha: p > .5  (p is the pop proportion of female students)

### Step 2: Data

In this sample, we saw a phat of 16/24.  

```{r}
16/24
```

Certainly, we observed a proportion greater than .5.  But, we know that sample results vary randomly; did we find strong evidence?  Let's find the probability!

### Step 3: P-value

Assuming p=.5, what's the probability that phat is >.6667?

```{r}
# compute P(phat > .66667) given p=.5

1-pnorm(.666667, .5, sqrt(.5*(1-.5)/24))
```

In other words, if the gender ratio really is 50%, there's about a 5.12% chance of observing a sample result like we one we got.

Our p-value:  p = .0512.  Close but no cigar!  Our p-val is somewhat small, but not smaller than .05.  This is NOT a small p-value!  

We fail to reject H0.  We failed to find evidence to support our claim.  H0 is getting off free/innocent, couldn't find it guilty!  There's not enough evidence to support our claim that more than 50% of students are female.

This sample result seems within random chance if H0 is true.

## Example 2

After a night of bad luck gambling, we suspect that coin is unfair and gets heads less than half the time.

To test this, we flipped the coin 50 times, and found 20 of them were heads.  Is there evidence to support out claim?  Hyp test!  Alpha = .05


### Hypotheses

H0: p = .5   (we assume that there's no effect)
Ha: p < .5   (p is the pop proportion of heads)


### Sample REsult

phat: 20/50

```{r}
20/50
```

This is less than 50%.  But, is it extreme enough to conlude about ALL POSSIBLE COIN TOSSES, not just 50 we did?  Is this sample result exteme enough for us to make a conclusion about the population?

### p-value

What's the prob that fair coin flips 20 out of 50 heads?
### Conclusion

```{r}
pnorm(.4, .5, sqrt(.5*(1-.5)/50))
```













