
---
title: "Math 120 Fall 2020 Homepage"
date: "Updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: false
    toc_depth: 3
    theme: yeti

      
  
---

```{r setup, include=FALSE}

                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans")) + 
  theme(axis.text.x = element_text(angle=-90))                            #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #

```
<br><br>

# **Final Quiz and HW Info**



To help you integrate the final quiz into your schedule, I've decided to let students take it **AT ANY TIME BETWEEN NOV 30 AT 2PM UNTIL DEC 1 AT 4PM (EST)**.


**THERE WILL BE A 1 HOUR TIME LIMIT** that begins as soon as you start the quiz.  Regardless, **the window will end at 4pm on Dec 1**.  So, don't start at 3:50 on Dec 1!

Remember, Notebowl saves your work the instant you type it.  **IF YOUR BROWSER FREEZES, JUST QUIT AND RELOAD NOTEBOWL!!!**  DON'T JUST WAIT!  I've done 100s of Notebowl quizzes, and never has a student lost any work from reloading.


The final HW is due at 10pm Monday Nov 30.



<br><br>


## Video Lectures

These are all on [our Google Drive shared folder](https://drive.google.com/drive/folders/1DUJNs5i-88NPDnLlGW04BCl0NkWm700S?usp=sharing)

The work I do in those vids is availabe on this site below.  



## Math 120 Lecture Notes
*They look great on your phone, so you can read them anywhere.  Steal my code!  Don't understand something?  Ask me!*

- [Week 1](https://millera2.github.io/Math-120/Math-120-Lec-1.html)
- [Week 2](https://millera2.github.io/Math-120/Math-120-Lec-2.html)
- [Week 3](https://millera2.github.io/Math-120/Math-120-Lec-3.html)
- [Week 4](https://millera2.github.io/Math-120/Math-120-Lec-4.html)
- [Week 4](https://millera2.github.io/Math-120/Math-120-Lec-4.html)
- [Week 5](https://millera2.github.io/Math-120/Math-120-Lec-5.html)
- [Week 6](https://millera2.github.io/Math-120/Math-120-Lec-6.html)
- [Week 7](https://millera2.github.io/Math-120/Math-120-Lec-7.html)
- [Week 8](https://millera2.github.io/Math-120/Math-120-Lec-8.html)
- [Week 9](https://millera2.github.io/Math-120/Math-120-Lec-9.html)
- [Week 10](https://millera2.github.io/Math-120/Math-120-Lec-10.html)
- [Week 11](https://millera2.github.io/Math-120/Math-120-Lec-11.html)
- [Week 12](https://millera2.github.io/Math-120/Math-120-Lec-12.html)
- [Week 13 and 14](https://millera2.github.io/Math-120/Math-120-Lec-14.html)


## Final Quiz: Week 13 and 14

Topics:  Hyp tests and CIs for two means, hyp tests and CIs for two proportions, error types in hypothesis testing.

Book practice:

- 5.25,(p 203)
- 6.21 (p 226)
- 6.23 (p 226)
- 7.25 (p 275)
- 7.27 (p 275)


## HW Assignments

[Click here for HW assignments](https://millera2.github.io/Math-120/HW.html)



## Syllabus

[Math 120 Fall 2020 Syllabus (Miller)](https://millera2.github.io/Math-120/Math-120-Fall2020-Miller.pdf)

## Office Hours 

### Prof Miller's Hours (Google Meet)

[Google Meet Room](https://meet.google.com/wtr-funk-wph)

Hours: Most weekdays before 4pm.

If you want to talk, shoot me an email (always use Denison email addresses) and propose a couple of times that work for you.

Of course, I don't check email in the evening, so we'll have to meet the next day in that case.

So, **don't wait till the night before a deadline to get help!**  

I like chatting with students -- come find me!

I think this will be better than setting aside specific times that work for everyone.

```{r}
qnorm(.1)
```



### TA Hours

*TBA*




## Shared Google Drive Folder:

[You can find all of our datasets here](https://drive.google.com/drive/folders/1tJFLMDFflBiXDHidWoTZU0SjzVi_z3zW?usp=sharing)




## Texbook Links

- [Math 120 Textbook](https://www.openintro.org/stat/textbook.php?stat_book=os)
  - *OpenIntro Statistics* is an excellent, free introduction to modern statistical analysis.
  

