---
title: "Math 120 Week 05"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: cerulean
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```










# Monday

## Probability


Textbook alert:  following Ch 3 very closely!

Goal:  answer the question "how likely?"

Answer:  probability.

Key vocab: chance event.  A chance event is any scenario in which at least two or unknown outcomes can occur.

Example:  Flipping a coin.  Two oucomes:  heads or tails.

Example:  Rolling a dice.  Six outcomes: 1, 2, 3, 4, 5, 6.  

## Two notions of probability

1)  The "classical" probability or "true" probability is the fraction of times we'd observe a particular outcome after seeing infinitlely many results of the chance event.  

Example 1: We roll a dice.  What's the probability of rolling a 6?  A:  1/6.  Six possible outcomes, only one 6,  all equally likely.  Thus, P(6) = 1/6

Ex 2:  We roll a dice.  What's the probability of seeing an even number or a prime numbrer?  Possibilities:  2, 3, 4, 5, 6.  Prob: 5/6.

Ex 3:  We roll a dice.  What's the probability of seeing a number that's even AND prime?  Only one outcome out of six, thus probability is 1/6.  

Ex 4: What's the probability of of rolling a number that's not prime?  Three out of six are not prime, thus probability is 3/6.

BIG PROBLEM:  Most of the time, we won't know all possible possible options!

SOLUTION:  Relative frequency probability.

2) The relative frequency probability of an event is the number of times we observed it (X), divided by the total number (n).   Ie, probability based on sample data.

Example:  In the mpg dataset, what's the probability that a randomly selected car is an suv?

```{r}
table(mpg$class) %>% addmargins()
```

We observed 62 suvs out of 234 cars total.  Relative frequency probability:  62/234.

BIG PROBLEM:  Relative frequency depend on chance, and don't always match up with true probaties.

I.e., if I flip a coin 10 times, I don't get 5 heads every single time.

FORTUNATE ANSWER:  THE LAW OF LARGE NUMBERS

As the number of repetitions of a chance event increases (n increases), relative frequency probability tends to get closer and closer to the true probability.  

I.e., if I flip a coin 10000 times, nearly 5000 of them should be heads.  


```{r}
sample(1:6, 10, replace = T) %>% table %>% prop.table()


```

Moral:  with small n (n=10), our observed proportions differer from 1/6

```{r}
sample(1:6, 10000, replace = T) %>% table %>% prop.table()

```


The law of large numbers is a relief!  It means we don't have to care too much, as long as we have big sample sizes!

## Probability Distributions

A probability dist is a list of all possible outcomes for X (our random variable), along with their probabilies.  Usually presented as a table.

## Example:  Dice distro

## Example:  

Flip two coins.  X = the number of heads.  Construct a probability distro for X.

## Example:  

Flip three coins.  X = the number of heads.  Construct a probability distro for X.

## Important probability distributions

Our goal:  get to know three of the most important ones.

Binomial dist, geometric dist, and poisson.

## Binomial Dist

Example:  A quiz has 5 multiple choice questions with answer a, b, c, d.


If a student guesses at random, what's the probability that she gets exactly 4 correct?

Suppose, CCCCI

(1/4)*(1/4)*(1/4)*(1/4)*(3/4)


```{r}

(1/4)*(1/4)*(1/4)*(1/4)*(3/4)
```

There are 5 such outcomes, and each has the same probability!

All together, the probability is 5*.002929688 = 

```{r}
5*.002929688
```

This is an example of a binomial probability!  We're counting the number of successes out a total number of trials!

The binomial has two parts:

1) The number of ways to get our success (above, that was 5)
2) The probability of each (they're always the same!)







# Monday

Recall, statistical inference is the process of making conclusions about populations based on sample data.  

To do this, we need to know:  how "likely" is our sample data?

Probability is the mathematical tool we'll use to answer that question.  Gives us context about our data, and what it might say about the population.

## The basics

NOTE:  Following closely from Ch 3 in the book.

Def:  A "random process" or a "chance event" is any scenario that could unfold in two or more unknown outcomes.

Example:  Flipping a coin is a random process.  There are two outcomes:  heads/tails.  No way to know in advance.  

Example:  Rolling a dice.  There are six outcomes:  1, 2, 3, 4, 5, 6.

NOTE:  The set of all possible outcomes is called the "sample space", S.

Def:  The probability of an outcome of a chance event is the ratio of times that we'd observe it if we could conduce an infinite number of trials of the chance event.

WEIRD!

Sometimes we can compute this.

Ex 1)  We roll a dice.  What's the probability we get a 4?  There are six equally likely outcomes, and only one of them is a 4.  

Ex 2)  Suppose we roll a dice.  What's the probability that the number is even or prime?  There are 5 numbers that are prime or even (2, 3, 4, 5, 6).  THus:   5/6

Ex 3) Roll a dice.  What's the probability that the number is prime AND even?  Only one number is both prime and even:  2.  Thus, 1/6.

Ex 4)  We roll two dice.  What's the probability that both of them are 1? 1/6*1/6 = 1/36.


Another definition of probability:  Relative Frequency.

Two notions of probability:

1)  "True" probability.  Defined above.
2)  Relative Frequency probability.  It's the ratio of times we observe an outcome after repeating the chance event n times.

Example:  Suppose Jack flips a coin 100 times.  He observes 55 heads.  The relative frequency probability is 55/100.

## Problems with both!

Generally, we can't know true probabilities because we can't repeat an experiment an infinite number of times.  

Also, relative frequencies don't always match the true probability!  Ex:  don't always get 50% heads when filpping a coin.

## Answer:  Law of Large numbers

The Law of Large Numbers states that as the number of repetitions (n) of our chance increases, the observed relative frequency get closer and closer to the true probability.

p_n -> p

This is great news!  It means, as long as our samples are big enough, both notions of probability are the same (or at least close).

## Example:  mpg

What's the probability that a randomly selected car is an suv?

```{r}
table(mpg$class) %>% addmargins()
```

62 suvs, 234 cars total, thus rel frequency is 62/234.  

##  Probability Distributions

A probability distribition has two things:

1) A list of all possible outcomes
2) Their associated probabilities.

Data distributions tell us about what happened.  Probablity distributions tell us what we expect.

## Example:  Dice

We roll a single dice.  X is the number shown.  Construct a distro for X.

## Example: Two coins

Two coins are flipped, and X = the number of heads.  Construct a distro for X.

## Example:  Three coins

Three coins are flipped, and X = the number of heads.  Construct a distro for X.

## Meet the Distros!  

Let's take a tour of the most important probability distros.

If you recognize a scenario that follows one, you'll be able to use it right away!

These all have built-in R commands.


## Binomial Distribution

A random variable X is "binomial" if it counts the number of "successes" out of a fixed number of trials (n).

Example:  Suppose we flip a coin 4 times.  What's the probability that we get exactly 3 heads?

To compute binomial probabilities, we need exactly two things:

1) How many ways are there to get that number of successes
2) What's the probability of each.

In the the example above, there are 4 ways to get exactly 3 heads.  Each has probability 1/2*1/2*1/2*1/2

All together:

```{r}
4*(1/2*1/2*1/2*1/2)
```





```{r}
-.766*5.955/1.292
```

```{r}
23.440+3.531*3.472
```

yhat = 35.670 - 3.531x












# Tuesday

## Binomial Dist


A random event / chance event is "binomial" if:

- Only two possible outcomes.  "Success" and "Failure"
- There are a fixed number of "trials" or repetitions of the chance event (n).
- The probability of success for each trials (p) must be constant.  Same for each.
- All trials must be independent.  (Can't affect eachother)

If the above are satisfied, and X counts the number of successes out of n trials, then we can find the probability of x:

Pieces of the formula:

n!/(x!(n-x)!) = "n choose x" = The number of ways to get x successes out of n trials.


## Example: Multiple Choice Quiz

A multiple choice exam has 10 questions, each with 4 options: a, b, c, d.  A student taking the exam hasn't studied, and must guess at random. 

What's the probability that she gets exactly 8 questions correct?  

First:  note that this is biniomial!

- Two outcomes:  correct or incorrect
- Fixed number of trials.  n=10
- Prob success is constant, p = 1/4
- Since guessing, all independent

```{r}
45*(1/4)^8*(3/4)^2
```

If guessing, there's a 0.04% chance of getting 8 right out of 10.  

##  Example:  free-throws

Prof Miller, despite being tall, is not very good at basketball.  Her free-throw average is 30% success.  If she shoots 7 free throws in a row, what's the chance she makes fewer than 3?

Note:  assume that she doesn't get tired, same 30% chance for each.  










# Tuesday

## Binomial Dist

A chance event/random event is "binomial" if:

- Only two possible outcomes.  "Success" and "Failure"
- The event is repeated a fixed number of "trials", n.
- On each trial, the probability of success (p) is constant (same for each trial)
- each trial is independent, i.e., don't affect eachother.

If X counts the number of success out of the n trials, then X has the following probability dist:

NOTE:  n!/(x!(n-x)!) = nCx = "n choose x" = the number of ways to choose x objects out of n.

Here:  the number of ways to get x successes out of n trials.

## Example:  Multiple Choice Quiz

A multiple choice exam has 10 questions, each with 4 options: a, b, c, d.  

A student hasn't studied, and must guess at random on each question.  What's the probability that the student gets exactly 8 questions correct?

- Only two outcomes for each question:  correct or incorrect.
- Fixed number of trials:  n=10 questions
- Prob of success is constant:  p = 1/4
- Since guessing, all are indpendent.

X counts the number of correct questions out of 10, thus binomial!

```{r}
45*(1/4)^8*(3/4)^2
```


There's about a 0.04% chance of guessing 8 correct out of 10.

## Example:  Free Throws

Despite being very tall, Prof Miller isn't very good at basketball.  Her free-throw success rate is 30%.

If she shoots 7 free-throws in a row, what's the chance she makes fewer than 3?

Let's assume that she doesn't get tired, 30% for each.








# Wednesday

Continuing our tour of probability distributions.  Next...


## Geometric Dist

The geo dist involves similar circumstances to the binomial:

- Only two possible outcomes, success or failure
- Constant probability of success
- All trials independent

The difference:  wheres binomial variable count the number of success in a fixed number of trials, the geometic distri


the geo dist counts the number of trials needed until the FIRST SUCCESS.

## Example:

The probability of rolling a die and getting "6" is 1/6.  What's the probability that we have to roll twice to get the fist "6"?

```{r}
(5/6)*(1/6)
```


## Example

What's the probability we have to roll the dice 4 times until we get a 6?

```{r}
(5/6)*(5/6)*(5/6)*(1/6)
```

```{r}
(5/6)^3*(1/6)
```


Geometric random variable count the number of trials until the first success.


## Example

About 21% of DU students are STEM majors.  If we're interviewing DU students, what's the prob that we have to talk to 10 students until we find the first STEM major?

```{r}
(.79)^9*.21
```

About a 2.5% chance.  

## Example

At Denison, about 55% of students are female.  What's the probability that we have to interview more than 2 students until we find the first female student?

INTERESTING NOTE!  Geometric variables can take an INFINITE NUMBER of possible options.

Here, could have X=3, 4, 5, 6, 7, ......... a zillion.

Solution:  1-P(not greater than 2)

```{r}
1-(.45)^1*.55 - (.45)^0*.55
```

There's a 20.25% chance that we have to interview more than 2 randomly selected DU students until we get the first female student.

## Poisson Dist

A Poisson random variable counts the number of occurances of an even over a unit of time or space.

Examples:

- On an average weekend night, the Chiotle in Heath, Oh serves 123 diners per night.  What's the probability that this Friday, over 150 customers come to Chipotle?

- Japan began recording earthquake data in 1884.  In the 135 years since then, the Tokyo region has experienced 5 earthquakes of the "strongest" category.  What's the probability that at least 1 will occur next year?

- At an archeological dig site of 25 square meters, 534 fragments were found at random across the site.  In 1 particular square meter, what's the probability that 20 to 22 fragments are found?  

Thes are all examples of Poisson processes.  They assume that the "rate" (lambda) of occurences is constant over time/space.  

If X is the number of occurances, then the distro is this:




# Wednesday

Continuing our tour of probability distributions...

## Geometric Dist

## Example

You know that the probability of rolling a "4" 1/6.  What's the probability that we have roll twice to get the first "4"?

```{r}
(5/6)*(1/6)
```

## Example

WHat's the probability you have to roll the dice 5 times until you get the first 4?

```{r}
(5/6)*(5/6)*(5/6)*(5/6)*(1/6)

(5/6)^4*(1/6)
```

About an 8% chance.

## Geo Dist

A geometric random variable counts the number of trials until the first success.

Assumptions/requirements:

- Only 2 outcomes:  success and fail
- Const prob of success (p)
- All trials to be independent

Biggest difference from binomial:  no fixed number of trials (n).  In this case, X has the distro:

## Example

At DU, about 21% of students are STEM majors.  What's the probability that we have to interview 10 randomly selected students until we find the first STEM major?

p = .21, X = 10

```{r}
(1-.21)^9*(.21)
```

About a 2.5% chance.

## Example

About 21% of students are STEM majors.  What's the probability we have to randomly interview more than 2 students before we find our first STEM major?

Woah, problem!  X = 3, 4, 5, 6, 7, ..... Infinitely many options!

Instead:  1 - P(not more than 2)

Woah, can't have X=0!  In geometric, need at least 1 success.  Interesting contrast from binomial!

```{r}
1-(1-.21)^(2-1)*.21-(1-.21)^(1-1)*(.21)
```


There's a 62.4% chance that we'll need more than 2 students until first STEM.

## Poisson Dist

Here are examples of Poisson problems:

- On an average weekend night at Chipotle in Heath, Oh, 123 customers are served.  What's the probability that on THIS friday, over 150 customers come to Chipotle.

- Japan began recording earthquake data in 1884.  Since then, the Tokyo area has experienced 5 of the "strongest" category earthquakes.  What's the probability that at least one such earthquake will occur next year?

- At an archeological dig site of 25 sq meters, a total of 534 fragments were fountd scattered at random.  In a particular 1 sq meter section, what's the probability that 20 to 22 fragments will be found?

A Poisson random variable counts the number of occurances of an event over a unit of time or space.  

If X is poisson, it has the following dist:




# Friday

## Solving Poisson Problems



- Japan began recording earthquake data in 1884.  In the 136 years since then, the Tokyo area has experienced 5 of the "strongest" category earthquakes.  What's the probability that at least one such earthquake will occur next year?

Find lambda:

```{r}
5/136
```


```{r}
1-(5/136)^0*exp(-5/136)/1
```

There's a 3.6% chance of an earthquake in the next year.

## Mix-up

Japan began recording earthquake data in 1884.  In the 136 years since then, the Tokyo area has experienced 5 of the "strongest" category earthquakes.  What's the probability that at least one such earthquake will occur next *two years*?

Same, but with lambda = 2*5/136 ...

```{r}
1-(2*5/136)^0*exp(-2*5/136)/1
```

A Poisson random variable counts the number of occurances of an event over a unit of time or space.  


## Probability functions in r:

- dbinom(x, n, p)
- dgeom(x, p)
- dpois(x, lambda)

## Example:  eartquakes

We saw lambda = 5/136, finding P(X=0).  So,

```{r}
1-dpois(0, 5/136)
```

## Important note:  dgeom

In book/class: x = total number of trials until first success
In r, dgeom:  x = number of failures, i.e., #trial -1

Example:  What's the probability we have to roll a dice 10 times until we find the first 6?  

Total trials = 10
Numbeber of failures = 9

```{r}
dgeom(9, 1/6)
```









# Friday

## Solving Poisson Problems



- Japan began recording earthquake data in 1884.  In the 136 years since then, the Tokyo area has experienced 5 of the "strongest" category earthquakes.  What's the probability that at least one such earthquake will occur next year?


Here, X~pois, it counts the number of earthquakes per year year.

```{r}
1-(5/136)^0*exp(-5/136)/1
```

About a 3.6% chance of strongest category quake.

## Mix-up


Japan began recording earthquake data in 1884.  In the 136 years since then, the Tokyo area has experienced 5 of the "strongest" category earthquakes.  What's the probability that at least one such earthquake will occur next five years?


Only diff:  lambda has changed:  lambda = 5*5/136 = 25/136


```{r}
1-(5*5/136)^0*exp(-5*5/136)/1
```



## Built-in R probability functions

Three:

dbinom(x, n, p)
dgeom(x, p)
dpois(x, lambda)

Example:  first quake example

1 - P(X=0),  lambda = 5/136

```{r}
1- dpois(0, 5/136)
```

WARNING:  dgeom works a little different

In class/book:  x = total number of trials
In R:           x = number of failures before the first success

Example:  What's the probability we have to roll a dice 10 times before we get the first six?

10 rolls -> 9 failures, last is success

```{r}
dgeom(9,1/6)
```





# Monday

## Probability from Two-Way (Contingency) Tables

Example:

```{r}
table(mpg$cyl, mpg$drv) %>% addmargins()
```


## First:  silly example on board

Important words:  and, or , not

## Example

Based on mpg data, what's the probability that a car has 4 cyl or rear wheel drive?

# Conditional Probability

Sometimes, knowing whether or not one event has occured affect another.

Ex:  if it's cloudy, there's a higher chance of rain.

This is called "conditional" probability.  Symbol:

P(A | B) = "probability of A, given B"

The thing after the vertical bar is the "given condition".

## Example:  dice

What's the probability of rolling a 2, given that a prime number was rolled?

## Independence

Idea:  two events are independent if the outcome of one doesn't affect the likelihood of the other.

Mathematically, there a few ways to check.

We can verify that events A and B are independent if any of the following hold:

1) P(A|B) = P(A)   (Doesn't matter if B happened, A unaffected)
2) P(B|A) = P(B)  (Other way around)
3) P(A and B) = P(A)*P(B)  (easy to check!)

## Example

Based on the animal table, are the events G = is a giraffe and R = is a republican independent?  Or, is there some kind of dependence relation?

P(G) = 30/100
P(R) = 40/100
P(G and R) = 10/100

Since .3*.4 is not the same as .1, no!  Not independent!




## Example


```{r}
table(mpg$cyl, mpg$drv) %>% addmargins()
```

Are the events 4 = 4 cyl and f = front wheel drive independent?

Q:  does P(4)*P(f) =? P(4 and f)

```{r}
81/234*106/234
```

```{r}
58/234
```



# Monday

## Imporatnt probability words:  and, or, not




## Probability from tables


"Two way" tables or "contingency" tables compare levels for TWO categorical variables.

```{r}
death <- read_csv("death penalty data.csv")

table(death$Method, death$Region) %>% addmargins()
```

## Conditional Probabaility

Idea:  sometimes, knowing that one event has occurred affect the likelihood of another.

Ex:  If it's cloudy, then there's a higher chance of rain.

We call this a "dependence" relationship.  

If no relationship, we call them "independent".

To find:  use "conditional" probability.

Symbol:  

P(A | B)  = "probability that A happens, given that we know B happened"

Ex)  Dice.  What's the probability that we roll an even number, given that we roll a number greater than 2?

## Checking for Independence

Q:  are two events "independent"?  Are A, B indep?  

Three ways to check:

1) P(A|B) = P(A)  (I.e., B doesn't affect A)
2) P(B|A) = P(B)  (other way around)
3) P(A and B) = P(A)*P(B) (usually easiest)

If any one of these three holds, then so do the others, and we say A, B are **independent**.

## Example: animals

Are the events G = giraffe and R = republican independent?

Need 3 pieces of info:

P(G)= .3
P(R)= .4
P(G and R)= .1

Since .3*.4 is NOT equal to .1, we've found that the events are NOT independent.  I.e. there's some dependence relation between being a giraffe and being republican.

## Example:  death penalty
```{r}
table(death$Method, death$Region) %>% addmargins()

```

Q:  are the events L = lethal injection and S = from the south indpendent?

P(L) = 1267/1442
P(S) = 1175/1442
P(L and S) = 1016/1442

```{r}
1267/1442*1175/1442
```

```{r}
1016/1442
```

Close, but not quite.  These are NOT independent events.











## Example 1

At a fast food restarurant, on average, 25 customers come through the drive-thru every hour.

WHat's the probability that at least 3 customers come in the next hour?

lamba = 25/1



## Example 1

At a fast food restarurant, on average, 25 customers come through the drive-thru every hour.

WHat's the probability that at least 3 customers come in the next half hour?

all the same, but 

lamba = 25/2


```{r}
#yhat

6.463+2.489*69
```
resid:  y - yhat

```{r}
145-178.204
```


Slope: 2.489

If a man's height increases by 1 inch, we'd predict his weight to increase by 2.489 lbs.


r^2:  67.4%

Of all the variation we see in men's weights, about 67.4% is due to the relationship between weight and height.  





Ex) Giraffes.  Adult male giraffe height follows a normal dist with mean 18' and std dev 1.5'.  

Approximately what percent of giraffes are between 13.5' and 21'?


Hold on.  What's a z score?

Idea:  different distros have different centers and spread.  So, we can't directly compare.  Example:  ACT and SAT.  Totally different scales!  

Z scores account for both center and spread:  for a measurement x, the z score is the distance from mean measured in std deviations.

Z = # of std devs away from the mean

(Note:  std dev ~ avd distance from the mean.  That's how it measures spread!)

Percent of giraffes:

```{r}
2.35+13.5+34+34+13.5
```


## Distro examples

About 82% of DU students are from out-of-state.  

1)  What's the the probability we have to sample 10 students before we find the first out-of-state student?

2)  What's the probability that, if we take a sample of 10 students, at least 7 of them are out-of-state?

 What's the probability that, if we take a sample of 10 students, less than 7 of them are out-of-state?
 
 P(X<7) = 1-P(X=7)-P(X=8)-P(x=9)+P(X=10)

3)  If we keep sampling students until we find the first out-of-state student, what's the prob that we have to sample at least 4 students?





















```{r}
.821^2
```

In general:  r^2 is the percent of variation in y variable that's becauase of/attributable to the lin relationship between x and y.

About 67.4% of the variation in weight is due to the relationship between height and weight.





## Exam review:

### Question:  requirements for poisson

If X is poisson, x counts the number of occurrances (successes) over a fixed interval of space/time  (a day, a year, two hours, sq meter).  
Requirements:

- Lambda is constant, i.e., fixed rate of occurences over time
- The probability of occurrences is proportional to the length of time/space.  I.e., longer time should generally mean more occurances.  Or, more space should mean more occurrences.  

### Question:  calculate R?

You never will.  Do it in software

### Meaning of "least squares" line

Zillion names:

least squares
regression
best fit
trend line 

All the same, there really is one unique "best" line.  THe least squares line is the one that minimizes the square residuals.

Q:  why minimize square resids? why not just resids?

A:  some are pos, some are neg.  The avg residual is always zero!  Boring!  Square to make all positive, i.e. minimize "total" error.

Q:  What's the meaning of standard deviation?

A: Std dev is "average distance" from the mean.




## Example:  8.23

a)  slope

```{r}
.636*113/99
```

intercept:

```{r}
129-.726*108
```


```{r}
0.636^2
```


Example)

About 82% are out of state.

If we take a sample of 10 students, what's the probability that at least 8 of them are out-of-state?









Example

At a fast food restaurant, customers come to the drive through at an average of 15 customers per hour.

What's the probability that at least 4 come in the next hour?

