---
title: "Math 120 Lec 23"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: cerulean
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```







## Hyp Tests

Last week, we focused on Confidence Intervals.  CIs were the first tool for statistical inference we've learned in this class:  they let us **estimate** a population parameter based upon a sample result.

This week, we'll be learning the other major tool for statistical inference:  **hypothesis tests**, or "hyp tests" for short.  Hyp tests are formalized procedures for **asking questions** about a population parameter.

**There's a lot to take in at first**, but just like CIs, **all hyp tests have the same structure.**  Once you get that down, you understand how all hyp tests work.

These might seem intimidating at first, but hang in there!  With a little practice, they're really not too tough.

I think the easisest way to understand hyp tests is to compare them to modern **criminal trials**.  Let's dive in!







## Hyp tests are criminal trials


Hypothesis tests are just like criminal trials.  The two are philosophically identical procedures.  We'll break the process down into four parts:

1) In a criminal trial, we wonder:  innocent or guilty?

In a hyp test, the NULL HYPOTHSIS (H0) is on trial.  The null hyp is a statement about the population.  The alternate hypothesis is logical opposite.  (Ha).

In a criminal trial, we assume innocent until proven guilty.  In a hypothesis test, we assume H0 is true, until we find evidence that's it's false.

The whole reason we're doing the test is becasue we suspect H0 is false!  I.e., the alt hyp (Ha) is our claim about the population.













2) In a criminal trial, we have witnesses.  In a hyp test, we have sample data!  We'll use special statistics, called "test statistics", to summarize the evidence.


3)  In a criminal trial, there are laws and lawers and arguments.  In a hyp test, there's probability!  (p-value)

A p-value is the probability of observing a sample result as or more extreme than the one we got, assuming H0 is true.  The assumption is important: we can't compute probabilities unless we make an assumption about the population!

Finding p-vals are identical to the **forwards problems** for sampling distributions that we practiced in Week 8 (the first post-apocalypse week).












4) In a criminal trial, a judge or jury weighs the strength of the evidence.  In a hyp test, we decide if our p-value is SMALL ENOUGH.  

If p is small, it means our sample result is unlikely under H0.  In this case, we REJECT H0, we trust our data!  Small p-val -> reject H0.  Support Ha.

Every hyp tests ends in one of two ways:  if p is small, we reject H0.  If p is large, we fail to reject H0.  

Wait, how small is "small"?   Totally up to us!  It's called the "significance level", "alpha".  Generally, we use alpha = .05 by default.  In other words, if our sample has less than 5% chance of occuring, we reject H0.

















## Example 1:  Hyp test for proportion

In Fall 2015, my stats class had 16 women out of 24 students.  Based upon this, I suspect that the gender ratio at DU was higher than 50% female.  Does this sample result support my claim?  Perform a hyp test.  Use alpha=.05.


Note:  phat = 16/24


1)  Hypotheses.  

H0: p = .50  (Assume by default that there's not an imbalance)
Ha: p > .50  (Think:  proportion of female students is greater than 50%)

2)  Test stat.  Think:  z score!

z = (obs-exp)/stdev

obs:  phat = 16/24
exp:  p = .5
stdev:  sqrt(.5*(1-.5)/24)

```{r}
(16/24 - .5)/sqrt(.5*(1-.5)/24)
```

3)  p-val:  Probability of getting a sample result as or more extreme, assuming that p = .5.

P(Z>1.63) = 1-.9484

```{r}
1-.9484
```

Here, p = .0516.  

4)  Remember, if p is small (p<.05), then reject H0.  Otherwise, if p>.05, we fail to reject H0.

Here it's close, but still .0516>.05, so we fail to reject the H0. 

We didn't find strong evidence to suggest that the proportion of female students at DU is greater than 50%.














## Example 2:  Hyp test for proportion

After a night of bad luck gambling, we suspect that dice is unfair.  Specifically, we think it rolls a 1 more frequently than it should (1/6 of the time).

To test this, we rolled the dice 100 times, and found that 25 of them were 1s.

Does this provide evidence that the proportion of "1" rolls is higher than expected?


1)  H0:  p = 1/6, Ha: p > 1/6

2) test stat

```{r}
(25/100-1/6)/sqrt(1/6*(1-1/6)/100)
```

3) p-val:

Look up on table,
```{r}
pnorm(2.24)
```

Thus p = 1-.9875

```{r}
1-.9875
```

4)  Since p<.05, reject H0.  There's strong evidence that the proportion of "1"s rolled with the dice is greater than 1/6.