---
title: "Math 120 Exam Review"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```

## Exam Review

Halle asks:  cdf vs pdf?

Answer:

- Key question:  interested in individual outcomes (X=2) or a range of outcomes (X>2).
-pdfs describe individuals.  Example, if we're flipping a coin 10 times, what's the prob that 4 of them are heads?
Think:  discrete or cts?  Discrete, since countable! Thus, either geom or binom or pois.  Fixed number of trials?  Yes!  n=10.  Thus, binom.
```{r}
dbinom(4, 10, 1/2)
```

Here, pdf was used since only one outcome.  

What if we wanted the prob of more than 2, less than 6 heads?

For discrete, very helpful to write out X vals.  

X = 3, 4, 5

```{r}
pbinom(5,10,1/2) - pbinom(2, 10, 1/2)
```

### List of commands to know

Note, different software uses different commands for the distros.  BUT, the concept is the same in all!

- dbinom(x,n,p), pbinom
- dpois(x, lambda), ppois
- dunif, punif(x, low, high)
- dnorm, pnorm(x, mean, stdev)

In r, pdfs always start with d, cdfs always start with p.

### Uniform distro example

Let's say that the amount of time a patients waits in the doctor's waiting office follows a uniform dist between 20 and 80 minutes.

a) What's the prob you have to wait for more than an hour?

P(X>60)= area of shaded = (20)*(1/60) =.333333

How to do in R?  Cts -> use cdf always -> punif

```{r}
  punif(80, 20, 80)-punif(60,20,80)
```

Note:  punif(80,20,80) = 1 !!! it's all the area!

```{r}
1-punif(60,20,80)
```

Warning!  The height of the unif is determined by its BOUNDS, not the X value in question.  It will always be the same!

### Clarification about geometric

DO understand geometric and be able to identify geom distros.

DO NOT worry about dgeom and pgeom.

### Normal dist problem

The daily coffee consumption of math professors at DU follows a normal dist with mean 16oz, std dev 3oz.

What's the probability that Prof Miller drinks more than 20z of coffee?

NOTE ABOUT MEAN AND STD DEV:  For normal dist, they must be given.

```{r}
1-pnorm(20,16,3)
```

What if we'd wanted to compare?  Z-score!

Here:

```{r}
#z score

(20-16)/3
```


```{r}
1-pnorm(1.33333)
```

Note:  if you don't supply mean and stdev, R will assume that you mean std normal with mu=0 and sigma=1.

About that z score:  what does z=1.3333 mean here?

In general, z scores measure the distance between an observation and the mean, measured in std dev.  Think:  std devs are the "Units" of the z score.

Here:  Prof Miller coffee consumption is 1.333 std dev above the mean.

### Graphs to know
 
This list is improvised and not comprehensive, but here are the imprtant graphs I'm thinking of now:

- histograms (numerical)
- bar graph (categorical data)
- box plot (interepte all the things!)
- scatter plots (correlation)
- residual plots (analyze error)

### Boxplot ex (see board for plot)

Every boxplot shows 6 things:

- low = 1
- q1 = 2
- med 4 (2 is possible, but less likely)
- q3 = 4
- high 6
- outliers (if any) 1, 6



### Robust

2) What does it mean for a statistic to be "robust"?  What are some examples of robust statistics?


A statistic is robust if it's not greatly affected by outliers and skew.

For center, the robust stat  is the median.  It's in the middle, so not affected by the extremes of the dist.

Ex:  1, 1, 1, 1, 1, 1, 1, 1, 1, 92345345645

```{r}
x <- c(1, 1, 1, 1, 1, 1, 1, 1, 1, 92345345645)

median(x)
```
```{r}
mean(x)
```

For spread, the IQR is our robust statistic.  Think:  IQR is the range of the middle 50% of the data (btw, that's the story for the equation).