---
title: "Math 120 Lec 15"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```


## Last time
Exam example:  What's the prob that it takes a student **exactly** 50.000000000000-... minutes?

It's zero!  Wuuuuut?  Weird!

P(X == 50) = 0

Why does this happen?  Continuous probabilities are all about *area*.  If we specify exactly one specific number, there's no area left!!!

What's the P(X==51)?  Still zero!  Wooooah!

This happens for ANY specific outcome of a continuous RV:

P(X==a) = 0 for any real number a.  If we get too specific, we exclude all outcomes!

I.e., for continuous RVs, it only makes sense to talk about RANGES of probabilities.

One more question on this issue:  same exam example, X has uniform dist ranging from 40 to 60.  Compute:

- P( 45  <= X <= 50) = (5)(1/20) = .25

- P( 45  <  X <  50) = same area, same prob!

#### Moral of the story

It doesn't matter whether or not we include endpoints for cts distros.

Big differnce from discrete distros!

## The normal distribution

Remember, our goal for the past week or so has been to meet the most important distros that are used in the real world.  We've seen:

- binom (discrete)
- geom (discrete)
- poisson ( discrete)
- uniform (cts)

Today, we're going meet another.  It's, **by far** the most useful and famous distro.  The bell curve, or **normal** distro.

(see picture on board, also, in your own personal notes because taking notes from the board is important).

A few key properties:

- Tails never touch.  So there's probability all along the nubmer line.
- Symmetric about the center, mu, it's mean/expected value
- The pdf and cdf are nasty, it's impossible to find these areas by hand.  Instead:
  - Use a statistical table, but that's sooo 2018.
  - Software.  It's easy in r:  `dnorm` and `pnorm`.
  - Note, since cts RVs are all about ranges, we'll almost never use `dnorm`.  
  - Note II:  pnorm gives you ALL of the area to the LEFT, i.e., all the way to neg infinity.
  
### Using the normal in R

Even though we say "the" normal dist, it's actually a family of distributions that depend on **center** and **spread**, i.e., expected value and variance.

Example:  human height follows a normal dist, but it's different for men and women.  For men, the center (exp val) is 69" (i.e. 5'9"), std dev 2.7.

For women, the center is 64" (i.e 5'4") with std dev 2.4.  Let's draw (see board)

Sophie's observation:  the curve for female heights is narrower and taller.  That's becasue of the smaller std dev.

Recall:  std dev = sqrt(variance)
          variance = (std dev)^2
          
They both measure variability.  We often discuss std dev because the units are the same as the data.

Let's compute:

#### Normal Examples

1) Using the distro info above, find the probability that a randomly selected woman is less than 66" tall.

```{r}
# X is normal, center is 64, std dev is 2.4

pnorm(66, 64, 2.4)




```

There's a 79.8% chance that a rando woman is 5'6" or shorter.  

2)  If we select a man at random, what's the prob that he's between 68" and 71"?

```{r}
#  X is norm, mean = 69, std dev = 2.7

pnorm(71, 69, 2.7) - pnorm(68, 69, 2.7)

```



  