---
title: "Math 120 Week 07"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    theme: cerulean
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```






# Monday

## Warm-up:  Normal dist in real life

1)  Height for adult men in the US follows a normal dist with mean 69" and stdev 2.7".  What's the probability that a randomly selected man is:

a) Taller than 66"?

```{r}
(66-69)/2.7
```

```{r}
1-.1335
```
There's about an 86.65% chance of this.  

b)  Between 68" and 72"?

z-scores:

```{r}
(68-69)/2.7
```
```{r}
(72-69)/2.7
```

P( -.37 < Z < 1.11) = P(Z<1.11) - P(Z< -0.37)

```{r}
.8665 - .3557
```


2)  Height for adult women in the US follows a normal dist with mean 64" and stdev 2.4".  What's the probability that a rando woman is 75" or taller?

```{r}
(75-64)/2.4
```

Off the table on the right, P(X>75) is approx zero!!

3) IQ follows a normal dist with mean 100, std dev 15.  How high must one's IQ be in order to be in the top 10% of all IQs?

From table:  z = 1.28 (90% below)

```{r}
1.28*15+100
```

One's IQ would have to be at least 119.2 in order to be in the top 10%.  

## Sampling Dist for xbar

Recall: a sampling distribution is probability distro for a sample statistic (like xbar or phat).  

First, we'll talk about the samp distro for xbar, sample means.  I.e., answer the question:  what's the probability of obtaining a sample mean like the one we got?

Key properties:

1)  The mean of the mean is the mean.  The expected outcome for our sample mean is the same as the population mean we're sampling from.  Taking samples doesn't change the center of the distro!

2)  sigma_xbar = sigma_x / sqrt(n).  What does this mean?  Increasing sample size (n) will DECREASE the spread of the sampling distro.  

3)  If the original pop is normal (think: human height), then xbar has a normal dist also, no matter what sample size.  Even n=2, perfectly normal.  If X is normnal, so is Xbar.

4)  (CLT)  If n is "large" (n>=30), then xbar has a normal dist NO MATTER WHAT the original distro for X is.  Wooooah!


Since #1 and #2 tell us center and spread, we can find z-scores!  Ie, can use the z-table!!

Example)  Suppose we take a sample of 50 adult men (recall that mean height for men is 69" and std dev is 2.7").  What's the probability that their mean height is at least 69.4"?  

Note:  we're not asking about a partiular man.  We're asking a sample!!

We know: 

mean for xbar:  69", same as mean for X

stdev for xbar:  2.7/sqrt(50)

```{r}
2.7/sqrt(50)
```

Need P(xbar > 69.4)

z-score:

```{r}
(69.4-69)/(2.7/sqrt(50))
```

```{r}
1-.8351
```

There's about a 16.49" chance that the sample mean height is at least 69.4".  

Example)  IQ follows a normal dist with mean 100, stdev 15.  If we take a sample of 10 people, what's the probability that their mean IQ is less than 90?

P(xbar < 90)

mean for xbar:  100 (doesn't change)
stdev for xbar:  15/sqrt(10)    (does change!)

z:

```{r}
(90-100)/(15/sqrt(10))
```



There's about a 1.46% chance of such a small sample mean!





# Monday

## Warm-up: Normal dist in real scenarios

1)  Height for adult men follows a normal dist with mean 69" and std dev 2.7".  What's the probability that a randomly selected man is...

a) taller than 66"?

```{r}
(66 - 69)/2.7
```
```{r}
1-.1335
```

There's about an 86.65% chance he's taller than 66".

b)  between 68 and 72"?

```{r}
(68-69)/2.7
```

```{r}
(72-69)/2.7
```

Big - small:

```{r}
.8665 - .3557
```

There's about a 51.08% chance he's in that range.

2) Height for women:  normal, mean = 64", stdev = 2.4.  What's the probability that a rando woman is 75" or taller?

```{r}
(75-64)/2.4
```

Off the table!  Prob(z>4.58) is approx zero!

3) IQ follows a normal dist with mean 100, stdev 15.  How high would your iq have to be in order to be in the top 1% of IQ?

From table:  z = 2.33

```{r}
2.33*15 + 100
```

One's iq would have to be at least 134.95.

## Sampling Distribution for xbar

Recall, a **sampling distribution**  is a probability distribution for a sample statistic.

Instead of describing individuals (X), samp dists describe sample results!

Two main ones:  samp dist for xbar for phat.  WE'll do xbar (sample mean) first!

Key properties of samp dist for xbar:

1)  The mean of the mean is the mean.  Our expectation for sample mean xbar is the same as our expectation of the population.  Center doesn't change!  

2)  sigma_xbar = sigma_x/sqrt(n)  The spread of the sampling dist gets smaller as n increases!

3)  If X is normal, so is xbar.  I.e., if we're sampling from a normal pop (think:  human height), then xbar will follow a normal dist for any sample size (even small n).  

4)  Central Limit Theorem:  If n is large enough (n>=30), then xbar has a normal dist NO MATTER WHAT the original pop distro is!!!

Fortunately, using this is easy.

Example)  Suppose we take a sample of 50 men.  What's the probability that their mean height is at least 69.4"?

(Remember, mean height = 69, stdev = 2.7)

Need a z score.  We know xbar has a normal dist (since X is normal and n is big).

Mean xbar = same as original pop, 69"

Stdev xbar = original sigma/sqrt(n).  Here:  2.7/sqrt(50)

```{r}
(69.4-69)/(2.7/sqrt(50))
```

P(xbar > 69.4) = P(Z > 1.05) = 1-.8531

```{r}
1-.8531
```

There's about a 14.69% chance of such a sample.

Example)  IQ:  normal, mean IQ = 100, stdev IQ = 15.

If we take a sample of 10 people, what's the prob that their mean IQ is less than 90?

n=10

mu_xbar = 100

sigma_xbar = 15/sqrt(10)

z score:

```{r}
(90 - 100)/(15/sqrt(10))
```

z=-2.11

1.74% chance of such a sample result.  

QUiz:  normal dist, CLT




















