---
title: "Math 120 Lab 03 - Basic Probability"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: false
    toc_depth: 3
    code_folding: hide
    theme: yeti
    df_print: paged
  
---

```{r setup, include=FALSE}
###########################################################################
################# DON'T DELETE THIS PART!!!!!##############################
###########################################################################
                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
library(kableExtra)                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans"))                                 #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #
### Below is where you read and store your data.                          #
                                                                          #
deathData <- read_csv("death penalty data.csv")
                                                                          #
                                                                          #
                                                                          #
############# ^ ################################# ^ #######################
############# | ## DON'T DELETE THIS PART!!!!!### | #######################
############# | ################################# | #######################
```



## Lab 3 Abbreviated Solutions

[Note: I used the `kable` package to make my tables prettier.  That wasn't required in this lab.]

1) If an executed prisoner is randomly selected from the data, what's the probability that they died by lethal injection?  

```{r}
table(deathData$Method) %>% addmargins() %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = F)
```

The probability is 1267/1442 = **`r 1267/1442*100`%**


2) If three executed prisoners are randomly selected from the data, what's the probability that all three died by lethal injection?  What assumption must you make in order to compute this probability?  Is that a reasonable assumption?

We may assume independence since large population.  Thus:  P(all three injection) = (.87864)^3 = **`r (.87864)^3`**

3) If an executed prisoner is randomly selected from the data, what's the probability that they died by lethal injection given that they were executed in the south?  

```{r}
table(deathData$Method, deathData$Region) %>% addmargins() %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = F)
```


The probability is 1016/1175 = **`r 1016/1175`.**

4) If an executed prisoner is randomly selected from the data, what's the probability that either they're black or that they died in the south?

```{r}
table(deathData$Race, deathData$Region) %>% addmargins() %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = F)
```

P(black or south) = `r 495+1175-426`/1442=**`r 495/1442 + 1175/1422 - 426/1442`**


5) If an executed prisoner is randomly selected from the data, what's the probability that they're both black and they died by electrocution?

```{r}
table(deathData$Method, deathData$Race) %>% addmargins() %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = F)
```

70/1442 = `r 70/1442`

6) Consider the events A and B, where
    - A = the event that a randomly selected executed prisoner is from the Midwest.
    - B = the event that a randomly selected executed prisoner died by firing squad.
  - Are events A and B mutually exclusive?  Why or why not?
  
```{r}
table(deathData$Region, deathData$Method) %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = F)
```
  
Yes, A and B are mutually exclusive since there are no outcomes in both.
  
7) Consider the events A and B, where
    - A = the event that a randomly selected executed prisoner is from the West.
    - B = the event that a randomly selected executed prisoner died by firing squad.
  - Using the definition of independence established in class, are events A and B independent?  Why or why not?  Is that the end of the story?  Are we convinced one way or the other?
  
```{r}
table(deathData$Region, deathData$Method) %>% addmargins() %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = F)
```

We see that P(West) = 85/1442, P(Firing Squad) = 3/1442, and P(West and Firing Squad) = 3/1442.

Since 85/1442 * 3/1442 != 3/1442, these events are not independent.  There's quite a difference on both sides, so we'd feel pretty sure that the events aren't independent.

  
8) Suppose X = the number of victims killed by a prisoner who's received the death penalty.  Construct a probability distribution for X.  Find the mean and variance of X.

```{r}
table(deathData$`Victim Count`) %>% prop.table() -> P

X <- P %>% names %>% as.numeric()


#mean/exp val:

sum(X*P)

#var:

sum( (X-1.537448)^2*P)

```


9) Referring to all the probabilities you've computed above: are these theoretical probabilities or relative frequency probabilities?  How does that contextualize the conclusions you've made?  What does the Law of Large Numbers have to say about it all?

Answers vary.  Key points:

- These are all relative frequency probabilities.  It's impossible to compute theoretical probabilities here, since there's no way to know the outcomes of all executions ever.  Generally speaking: any time we obtain probabilities from data, we're always working with relative frequency (also called "experimental") probabilities.  From our perspective as statisticians, this is an inconvenient truth!

- The Law of Large Numbers states that, as sample sizes increase, relative frequencies tend to get closer to theoretical probabilities.  From our perspective as statisticians, this is a relief!  Even though the truth is unknowable, large sample sizes give us a better and better approximation of the truth.

- The observations that we've made are very coarse and simplistic.  That's ok; this is just a starting point!  In just a couple of weeks, we'll have the additional tools needed to make smart conclusions about our populations based on these simple measures here.  

- Remember in general that quantitative analysis (i.e, what we do here in Math 120) should **accompany** qualitative analysis (i.e., what you do in a humanities course), but it should never **replace** it.  Crunching numbers is just one more tool we humans have for navigating the dim and murky depths of reality as we perceive it.  Understanding who these people were, what happened, and the nuances of their crimes and trials is just as (or much more) important as understanding conditional probabilities, variances, and two-way tables.  

