---
title: "Math 120 Week 09"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    theme: cerulean
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```








# Week 9 - Vid 1


Remember, you can pause the video at any time!

Pause frequently to think, take notes, and **practice**.

Pre-Reading Assignment:  Week 8 Lecture Notes


Secret Video Question Answer:  Purple














## Review of Sampling Distributions

Remember, sampling distributions are probability distributions for sample statistics.  We focus on two samp dists:

- Samp dist for xbar (sample mean for quantitative data)
- Samp dist for phat (samp proportion for categorical data)

Samp dists let us answer a key question:  

- **How likely is the sample result that we observed?**








## Center and Spread for Samp Dists

For xbar:

- mu_xbar = mu  (expect sample mean to be close to population mean)
- sigma_xbar = sigma/sqrt(n)  (spread decreases as n increases)

For phat:

- mu_phat = p  (expect sample proportion to be close to population proportion)
- sigma_phat = sqrt[  p*(1-p)/n  ]  (spread decreases as n increases)








## Shape for Samp Dists

Fortunately, the Central Limit Theorem tells us that samp dists always follow a **normal** distribution, as long as our sample size is **big enough**.

- For xbar, we need n at least 30
- For phat, we need both np and n(1-p) at least 10










## Example for xbar (forwards)

Suppose that the mean resting heart rate of DU students is 66.1 bpm, with standard deviaiton 6.2.

If we take a sample of 45 students, what's the probability that their mean blood pressure is more than 68 bpm?

mu = 66.1,   sigma = 6.2  (population info)

Q: P(xbar > 68)

Think:  xbar = 68, n=45

Note:  since n>=30, we know that xbar follows a normal dist (CLT).  Can use the z table!

zscore = (obs - exp)/stdev

obs = xbar (sample result) = 68
exp = mu (population mean) = 66.1
stdev = 6.2/sqrt(45)

```{r}
(68 - 66.1)/(6.2/sqrt(45))
```

Here, z = 2.05574 = 2.06

Remember:  Q:  P(xbar > 68) = P( z > 2.06)

Note:  pnorm is same as Z table
```{r}
pnorm(2.06)
```

Table:  .9803

Remember, "greater than", so "1-prob"
```{r}
1-.9803
```

There's about a 1.97% chance of observing a sample where mean blood pressure is greater than 68bpm.  











## Example for xbar (backwards)

Suppose that the mean resting heart rate of DU students is 66.1 bpm, with standard deviaiton 6.2.

If we take a sample of size 100, in what range of values would we expect the middle 95% of such sample means to lie?

mu = 66.1, sigma = 6.2

n = 100

First:  use area to find z-score.  

If middle area is 95%, each tail area is 2.5% (.025).  Half of 5%!

In the z table, find the closest value to .025.  This is exactly in the table, 

z = -1.96.  Use both:  +-1.96.  

Remember z score formula for xbar:


z = (xbar - mu)/(sigma/sqrt(n))

Solve for xbar:  

xbar = z*(sigma/sqrt(n)) + mu

z = +-1.96 (use both)
sigma = 6.2
n = 100
mu = 66.1

Now, find xbar:

```{r}
-1.96*(6.2/sqrt(100))+66.1
```
```{r}
1.96*(6.2/sqrt(100))+66.1
```

95% of the time we take a sample of size n=100, we'll find a sample mean between 64.8848 bpm and 67.3512 bpm.  






# Vid 2




## Example for phat (forwards)

About 51% of all DU student are female.  

If we take a sample of 100 DU students, what's the probability that at least half of them are female?

Note:  about probability of a sample result.  Use samp dist!

p = .51 (population proportion female)
n = 100

P( phat  > .5 )   [i.e. sample prop is at least half]

Q:  does phat have a normal dist?  Is our sample big enough?  

A:  Check, need np > 10 and n(1-p) > 10

```{r}
100*.51
100*(1-.51)
```
Yes!  Both are bigger than 10.  CLT says phat is normal!

z = (obs - exp)/stdev

obs:  phat = .5
exp:  p = .51
stdev: sqrt(.5*(1-.5)/100)

Note:  use p (pop prop) in stdev, not phat!
```{r}
(.5 - .51)/sqrt(.51*(1-.51)/100)
```

z = -.20 

In table, P(Z< -.20) = .4207

```{r}
pnorm(-.2)
```

"At least" means "bigger than", so do 1-prob

```{r}
1-.4207
```

There's about a 57.93% chance that a sample of 100 DU students is comprised of at least half female students.  Feelings:  this is a very common result.









## Example for phat (backwards)

About 51% of DU students are female.

If we take samples of size 100, in what range would the middle 99% of sample proportions lie?  

Note:  given area (middle 99%), need to find value (sample proportion).  Backwards!

If middle 99%, each of the two tails has 0.5% , .005.  Find the closest to .005 in table, get z score.  

Table tells us: z = -2.575.  Use both +- for the two bounds.  

For phat,

z = (phat - p)/sqrt(p*(1-p)/n)

phat = z*sqrt(p(1-p)/n)+p  (backwards z score for phat)

Lower bound:
```{r}
-2.575*sqrt(.51*(1-.51)/100)+.51
```

Upper bound:

```{r}
2.575*sqrt(.51*(1-.51)/100)+.51
```

99% of the time, the sample proportion of female students would be between 38.12% and 63.87%.  


