---
title: "Math 120 Week 12"
author: "Prof. Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: cerulean
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```












# Week 13 Vid 1

## Final topic: Hyp test and CIs for two means

We're ready to cover the two **most important tools in this course**: hyp tests and confidence intervals for **two means**.

This will be very similar to the work we did already for 1 mean (i.e., tests and CIs for mu).  The new twist:  now, instead of estimating (confidence interval) or asking questions about (hypothesis test) one mean from one population, we'll be **comparing two different populations**.

If you only take one thing with you from this course, **this is it!**.  We're ending on a bang!  Woohoo!













## The basics

First, let's remember what we already know.  The new techniques are very similar.

If we're talking about a quantitative variable for a population, then

    point estimate for mu = xbar

Remember, point estimates are a single numerical approximation of a parameter.  Above, we're restating the familiar:  we use xbar (the sample average) as our point estimate for mu (the population mean).  

We need to know the standard error (variability/spread/width) for this estimate too.

    standard error for xbar = s/sqrt(n)
    
Where "s" is the standard deviation of the sample, and n is the sample size.  This is the what we used in our CIs and hyp tests for mu.

Ok, what if we have **two** samples from two different populations....?















If we have two populations, then we want to talk about the **difference** between them.  

For all of these examples, you can imagine a medical trial:  if there's a control group (who didn't receive the new drug) and a test group (who did receive the new drug), then we want to see if there's a meaninfgul difference.  Think of these as population 1 and population 2.

Each population must have its own population mean (we call these mu1 and m2), and thus the thing we care about is

    mu1 - mu2
    
In order to do CIs and hyp tests, we need a point estimate and a standard error, just like before.  Here:

    point estimate for mu1 - mu2 = xbar1 - xbar2

Makes sense!  To estimate the difference between the populations, we'll use the differnce between the **sample means** xbar1 and xbar2.

The standard error is shown below: (check page 267 of book for pretty typeset)

    std error for xbar1-xbar2 = sqrt(s1^2/n1 + s2^2/n2)
    
These simple building blocks are all we need to do Hyp Tests and CIs.  Let's take a look!




















## Hyp Tests for 2 means

This hyp tests answers a very important (and frequently-asked) question:  **is there evidence of a difference** between the two groups??

These hyp tests have the same 4 steps as any others:

1) Hypotheses
2) Test Stat
3) P-val
4) Conclusion









### Hypotheses for 2 means

In -all- hyp tests for two means, the **null hyp is always the same**:

H0:  mu1 - mu2 = 0

In other words, we **always** assume that there's **no difference** between the two populations.This part of the test is easier!

For the alternate hypothesis, there are three possibilities:

- Ha:  mu1-mu2  < 0    (left tail)
- Ha:  mu1-mu2  > 0    (right tail)
- Ha:  mu1-mu2 != 0    (two tail)

This depends on our "claim" or "suspicion" about the population.
















### Test Statistic for 2 means

Our test stat has the same form as all the others:

    test stat = (observed - expected)/(standard error)

Here, 

observed = xbar1 - xbar 2 (the actual difference in sample means)
expected = mu1 - mu2 = 0 (we always assume no difference for the population means)
std error = sqrt(s1^2/n1 + s2^2/n2)

All together, that's:

t = (xbar1 - xbar2)/sqrt(s1^2/n1 + s2^2/n2)

Looks nasty, but it's really not too bad to type out.














###  p-val for 2 means

In short:  use the t-table **exactly as before**.  

There is a problem, though:  what's our degrees of freedom?  After all, there are -two- samples, with -two- sample sizes.  Which one do we use?

There's a fairly complicated formula for finding the exact df.  Here, we'll take a simpler approach:  **use the smaller of the two** to find df.  

For example, if sample 1 has n1 = 34, while sample 2 has n2 = 43, then use df = 34-1 = 33.


## Conclusion

Same as always!  If p<.05, we reject H0 and find evidence to support our claim.  If p>.05, we fail to reject H0 and don't have enough evidence to support our claim.

Ok, let's try one out!















## Example 1: Hyp Test for 2 means

A hungry math professor wonders:  does Pizza Hut have a different average delivery time than Donatos?

To test this, she records data about her many pizza orders.  

Over the 35 times she ordered at Pizza Hut, the average delivery time was 24.3 minutes, std dev 4.1 minutes.

Over the 28 times she ordered at Donatos, the average delivery time was 26.3 minutes, std dev of 5.8 minutes.

Is there statistically significant evidence to suggest that Pizza Hut delivery times differ from Donatos?

Let's do the test!

### Collecting info

Since there's a lot of info in the problem, I reccommend that studetns take a moment to collect it all.  Here, let's have Pizza Hut be sample 1, and Donatos is sample 2.  That means:

n1 = 35, xbar1 = 24.3, and s1 = 4.1

n2 = 28, xbar2 = 26.3, and s2 = 5.8

This makes it easier to run the test.

### The test

1) H0: mu1-mu2 = 0,    Ha: mu1-mu2 != 0 (two tail!)

2) Using the formula for the test statistic (see above):

```{r}
(24.3 - 26.3)/sqrt(4.1^2/35 + 5.8^2/28)
```

Look up 1.54 in the t table.  

3)  P-val

Use df = 28 - 1 = 27.  (Used the smaller sample size.)

Thus:  .1 < p-val <.2


4) Conclusion:  

Here, p>.05 (remember, alpha = .05 is our "default" significance level), so we fail to reject H0.
There's not enough evidence to conclude that there's a differnce in mean delivery times between the two pizza chains.













## Confidence intervals for 2 means

Ok, let's finish off with a confidence interval.  CIs for two means **estimate the difference in mean** between the two populations.  How different are they?  What are the possibilities?

We've seen that all CIs have the following format:

(point estimate) +/- (critical value)*(standard error)

Using the info for 2 means, that gives us:

(xbar1 - xbar2) +/- t*sqrt(s1^2/n1 + s2^2/n2)

Here, the t critical value is exactly the same as it was before.  No problem!

Let's do an example with the same data from the pizza delivery test.


















## Example 2:  CI for two means

Over the 35 times she ordered at Pizza Hut, the average delivery time was 24.3 minutes, std dev 4.1 minutes.

Over the 28 times she ordered at Donatos, the average delivery time was 26.3 minutes, std dev of 5.8 minutes.

Let's **estimate the difference in mean delivery time** between Pizza Hut and Donatos with a 95% confidence interval.

The only piece of info we don't have is the **critical value**, t.  Since the CI continas the middle 95%, that means our tail areas are 2.5% each.  Thus, look up .025 (one tail) or .05 (two tail) in the t table.  

Our df here is 27 (just like for the test).  Thus,

t = 2.05

n1 = 35, xbar1 = 24.3, and s1 = 4.1

n2 = 28, xbar2 = 26.3, and s2 = 5.8

Plug and chug:

Lower:
```{r}
(24.3 - 26.3) - 2.05*sqrt(4.1^2/35 + 5.8^2/28)
```

Upper:
```{r}
(24.3 - 26.3) + 2.05*sqrt(4.1^2/35 + 5.8^2/28)
```

We're 95% confident that the true difference in mean delivery time is between -4.7 minutes and 0.7 minutes.

Think about it this way:  notice that it's possible that the difference in mean delivery time is zero!

         (-----------------------------)
  mu1-mu2 = -4.7                  mu1-mu2 = 0.7

I.e., it's possible that there's -NO- difference!  This agrees with our hyp test, where we failed to reject the H0: mu1-mu2 = 0.  The confidence interval says the same thing:  it's possible that the difference is zero!


## Quiz 10 Solutions

According to studies performed in the 1950s, about 10.5% (i.e., 0.105) of all Americans are diabetic.

A medical researcher suspects that modern proportions differ from this historical value.  To investigate, she takes a random sample of 320 Americans and finds that 45 of the 320 participants are diabetic.

Does this sample data support her suspicion?  Perform a complete hypothesis test.  **Show all 4 steps for full credit**

### Hypotheses

H0:  p = .105
Ha:  p != .105

### Test stat

```{r}
(45/320 - .105)/sqrt(.105*(1-.105)/320)
```

### p-val

Look up in z-table, tail area = .0188

Since 2-tailed test,p-val = 2*.0188 

```{r}
2*.0188
```

### Conclusion

Since p<.05, we reject H0.  There's no evidence that the modern proportion of diabetics differs from the 1950s proportion of 10.5%.