---
title: "Math 120 Week 10"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

studentSurvey <- read_csv("Student Data.csv")
deathPenaltyData <- read_csv("death penalty data.csv")
housingData <- read_csv("../Data/House-Sales.csv")
#------------------------------------------------#
```


Quiz #3:

x = 60, mean = 64, stdev = 2.4

z-score:

```{r}
(60-64)/2.4
```

z = -1.67

From table, left area is 0.0475.  We want "at least":

```{r}
1-0.0475
```

Quiz #4

Find z-scores:

x = 20:

```{r}
(20-21)/5.8
```

x = 25:

```{r}
(25-21)/5.8
```

Look em up!

```{r}
0.7549-0.4325
```

Quiz #5:

```{r}
(100-75)/6.1
```

Almost no area above z=4, so P(X>100) ~ 0.


Quiz #6

If he's in top 60%, he's ABOVE 40%.

z = -.25

z = (x-mean)/stdev

-.25 = (x - 69)/2.7

```{r}
-.25*2.7+69
```

Quiz #7:

For middle 50%, tail cutoffs are z = +/- .67

```{r}
-.67*7.3+76
```

```{r}
+.67*7.3+76
```

IQR:  

```{r}
80.891-72.109
```


# Monday Oct 26

Section 5.2:  Confidence intervals.

We're about to do real, honest statistical inference! Yaay!


A confidence interval is a range of reasonable values for a population parameter based upon sample data.


We never know exactly what the true parameter is, but we can provide a likely range of values.


First, we're going estimate population proportion, p.  Our first example:  estimate the proportion of DU students who have brown hair.  

TO find this, simply find the "middle %" of the sampling distribution based on our sample result.

Common values:  95% confidence interval, 90% confidence interval, 99% confidence interval.

Here, the distribution we're looking at is the sampling distribution for phat.  CATEGORICAL DATA!!!

This distribution is normal, has mean p, stdev p(1-p)/n.  


If we're trying to estimate p (pop prop), so we'll have to use sample prop phat instead.


    z-score:  (phat - p)/stdev(phat*(1-phat)/n)
    
Goal:  solve this for p!  Thre's always two:  one for the negative z-score, one for the positive.  

After algebra, here's the forumula for CI for p:


           phat - z*sqrt(phat*(1-phat)/n)  <  p <  phat + z*sqrt(phat*(1-phat)/n)

$$ \hat{p} \pm z^* \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

The z-score here has a special name, "critical value".  It's the z-score that corresponds to the 'middle %'.

CAUTION:  THe interval is only valid if our sample size is LARGE ENOUGH!

    phat*n >= 10    (1-phat)*n >= 10 
    
**Example**  Using our class data (assume that it's a random sample), estmiate the proportion of all DU students with brown hair with 95% confidence.

From our data:  n = 42, phat = 28/42

CHECK:  Is n large enough?

```{r}
42*28/42
```

```{r}
42*(1-28/42)
```

Both larger than 10, so yes!  Use normal dist.

Need to find critical value z*.  Since 95% is in the middle, that means there's 2.5% left over for the tails.  Let's find the z-score that matches area=.025.  Use critical value zstar = 1.96.


Plug into formula for CI:

    phat - z*sqrt(phat(1-phat)/n)

```{r}
(28/42) - 1.96*sqrt((28/42)*(1-28/42)/42)
```

```{r}
(28/42) + 1.96*sqrt((28/42)*(1-28/42)/42)
```

Based upon our sample data, we're 95% confident that the true proportion of all DU students is between 52.4% and 80.9%.


### Margin of Error

The MOE is the stuff after the +/- in the CI formula.  It determines the width of the interval!


    ( ---------------------   phat   ---------------------)
    
Visually, the MOE is the width of the arms.

In the hair example, our MOE was

```{r}
1.96*sqrt((28/42)*(1-28/42)/42)
```

In general, we hope that our MOE is small!  Maybe 14.2% doesn't seem so great....














































# Monday 10-19

## Ch 5:  Inference

Recall, **statistical inference** is the process of making mathematically supported conclusions about populations based upon sample data.


Statistical inference always starts with a "point estimate", ie a single number that represents our best guess about the population paramater.

  (Rember:

  - Parameters are summaries of populations
  - Statistics are summaries of samples

  )

You already know a couple of point estimates:

-xbar is a point estimate for a population mean, mu
-phat is a point estimate for a population proportion, p

Two problems:

-No guarantee that these are correct for populations!  Sample are random!
- (So far) we don't know how much variability there is in these point estimates.  

Let's study both of these with an example from the text (p170):

- Given the parameter:  p = .88  (in real life, we wouldn't know!)

From some of our example random samples of size 1000, we see the resulting proportion varies.  We saw:  86.1%, 87.8%, 90.1%, 89.4%, etc etc.

Idea:  let's perform this sample 5000 times and look at the results.  We are going to make a histogram of 5000 sample proportions from 5000 samples.  

Observations from the histogram:

- Looks like sample proportions follow a normal dist (ooh!)
- The center of the distribution is .88 (most are close to the true value).
- We have an idea for the variability of the sample results.  Here, they all seem to be within +/- 4% of the true value.

## Sampling Distributions

The histogram you saw above is an example of a "sampling dist", ie the distribution for possible SAMPLE RESULTS.

Restate:  a sampling distribution is a probability distribution that describes the likelihood of observing a sample result of fixed size, n (i.e. it DOESN'T tell you about individuals!)

As you've seen, these often look like normal dists!  This amazing fact is called the **Central Limit Theorem**.  

According to the CLT, sample statistics (like xbar and phat) will always follow a normal distribution as long as the sample size is "large enough".

For **sample proportions**, phat,  they follow a normal dist with:

1) mean = p

$$\mu_{\hat{p}}=p$$

In other words, the expected value for sample proportions phat is the true population proportion, p.

2) stdev = sqrt(p*(1-p)/n)

$$\sigma_\hat{p}=\sqrt{p(1-p)/n}$$

Thus, for any particular sample proportion, phat, we can find z-scores!

For phat:

     z = (phat - p)/sqrt(p*(1-p)/n)
       = (obs - exp)/stdev

























# Friday Oct 23

Recall: sampling distributions are probability distributions for entire sample results.

The goal:  answer the question, "how likely (or unlikely) is our sample result?

Remember, the only reason this is possible is because of the CENTRAL LIMIT THEOREM: ie, if our sample size is "large enough", we know that our sample results (xbar and phat) follow a normal distribution.

Wait, what's "large enough?"  Answer: it depends!

First, we'll think about categorical variables (phat).  
Later, we'll think about quantitative variables (xbar).  

For now, "large enough" for categorical variables means the following:

    n*phat >= 10   n*(1-phat) >= 10
    
Equivalently, we need at least 10 individuals that ARE IN the category, and at least 10 indivuals that ARE NOT IN the category.  


**Ex**  In our class, 65% of students have brown hair.  There are 50 students (n) in our class.  Could we use a normal distribution to describe the likelihood of such a scenario?

Check:

```{r}
50*.65
```

```{r}
50*.35
```

Yes!  n is "large enough"!

**Ex**  Suppose that 60% of all DU students have brown hair.   How likely would it be to take a sample of 50 students, and find that 65% (or more) of those students have brown hair?


We saw above that n is "large enough"

mu_phat = mean = .6 (population proportion)
sigma_phat = stdev = sqrt(.6*(1-.6)/50)

```{r}
sqrt(.6*(1-.6)/50)
```

obs = .65
exp = .60
stdev = .0693

    z = (obs - exp)/stdev
      = (.65 - .6)/.0693
      

```{r}
(.65 - .6)/.0693
```

z = .72.  Look it up!


```{r}
1-0.7642
```

There's about a 23.58% chance that in a samle of 50 students, 65% or more would have brown hair.

Such an occurrance is not unlikely.


**Ex**  According to the internet, 82% of DU students are from "out-of-state" (not from Ohio).  

If we took a sample of 100 students, what's the probability that at least 90% of them are from out-of-state?

Check to see if we can use normal dist:


```{r}
100*.9
```

```{r}
100*(1-.9)
```

Close!  But, both are at least 10, so we CAN use the normal dist!

obs = .9
exp = .82
stdev = sqrt(.82*(1-.82)/100)


```{r}
sqrt(.82*(1-.82)/100)
```

Find zscore!


```{r}
(.9 - .82)/.03842
```

z= 2.08



```{r}
1-0.9812
```

There's about a 1.88% chance of obtaining such a sample result.  Pretty unlikely!















