---
title: "Math 120 Lec 24"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```





## Hyp Tests (Continued)


CIs and Hyp Tests are our two big tools for "inference"

CIs are used for estimates.  What is it?

Hyp Tests are used for answering questions.  Is our support for our claim?


### Four key steps of hyp tests

#### 1) Hypotheses

Two competing hyps:

H0:  the null hyp.  This is the assumption we make about the population.  It's ALWAYS a statement of equality about a pop parameter.

Ex:  H0:  p = .5  (note that p is the pop proportion)

It's needed to compute p-val!  We can't find probabilities unless we make an assumption about pop.  That's H0.

Ha: the alt hyp.  The alt hyp is our claim about the population.  It's what we think is "up".  It's the reason we're doing the test.  We hope to support it!  

The alt hyp is always a statement of INEQUALITY.  <  >  !=

Example:  Ha:  p < .5

The alt hyp is the logical opposite of the null hyp.  Only one can be true!

#### 2) Test statistic

In short: a summary of our data.  

Example:  phat

Almost always, expressed as a z score.  How about for phat?

#####

z = (phat - p)/sqrt(p(1-p)/n)

#####

#### 3) p-val

The p-val is the probability of observing a sample that's as (or more) extreme than the one we got, assuming H0 is true.

In short, to find:  pnorm.  Find tail area past the test stat.

#### 4) Conclusion

If our p-val is small, then H0 is saying that our data is unlikely.  But of course we trust our data, not the hypothesis!

So, if p is small, we reject H0.  We have evidence to support Ha.

(Remember, "small" means "alpha", alpha = .05 by default)

If p is large (p>alpha), then we FAIL TO REJECT H0.  We didn't find the evidence we're looking for to support Ha.  We have sad feelings, our paper doesn't get published, try again!


#### Example: 5.32 p205

Do we have evidence that more than 8% of children are nearsighted.

phat = 21/194

```{r}
21/194
```

The sample result IS BIGGER than 8%.  Is this result extreme enough to convince us that the population proportion really is greater than 8%?  How likely is this sample? What human feeligns do we have?

1) **Hypotheses**

    H0:  p = .08
    
    Ha:  p > .08

2) **Test stat**

    phat = 21/194

```{r}
# find zscore.  note: center = p, se = sqrt(blaahblahs;as)


# idea:  (phat - p)/sqrt(p*(1-p)/n)

(21/194 - .08)/sqrt(.08*(1-.08)/194)


```

  (Human feelings:  Not an unusual score!)

3) **p-val**

```{r}
1-pnorm(1.450243)
```

    Note:  this is a "right-tail" test.  You can tell from Ha.

4) **Conclusion**

    Is p small?  Is p<alpha?  Is p < .05?

    No!  p is large!  We didn't find strong evidence against H0.  The sample results doesn't seem unlikely under H0.  We fail to reject H0!

    We didn't find the evidence we're looking for to suggest that more than 8% of children are nearsighted.  Nobody publishes our result.  We are sad and dejected.  