---
title: "Math 120 Lec 04"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}
###########################################################################
################# DON'T DELETE THIS PART!!!!!##############################
###########################################################################
                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans"))                                 #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #
### Below is where you read and store your data.                          #
                                                                          #
cirpData <- read_csv("CIRP_DATA_2014_CLEAN.csv")
                                                                          #
                                                                          #
                                                                          #
############# ^ ################################# ^ #######################
############# | ## DON'T DELETE THIS PART!!!!!### | #######################
############# | ################################# | #######################
```

## Reading discussion: How did lab go? 

Remember: the other humans around you are very, very helpful!  ;D

## Box Plots and Scatterplots In-Depth

### Parts of a Boxplot

See boardwork.

Example:  iris Sepal.Length

```{r}

iris %>% head

# make boxplot of sepal length

#qplot note:  you need both x and y here

#  x = facet category
#  y = the actual data you're interested in (sepal length)


qplot(x=Species,
      y=Sepal.Length,
      data=iris,
      geom="boxplot",
      fill=Species)


```


### Weird-looking boxplot examples!

The cirp data has some weird ones!

```{r}
# let's boxplot YRSTUDY1

cirpData <- read_csv("CIRP_DATA_2014_CLEAN.csv")


cirpData %>% head

qplot(x=SEX,
      y=YRSTUDY1,
      data=cirpData,
      geom="boxplot")



```

Woah!  What happened?  Almost everyon chose "6", i.e., 4 years of english. 

Think:    6, 6, 6, 6, 6, 6, 6, 6, 6, 6,6 ,6, 6, 6, 6,6, 4, 5, 7

### When scatterplots make sense (and when they don’t)

First, a good example: mpg dataset

```{r}

mpg %>% head

#idea:  compare city and hwy fuel efficiency!

qplot(cty, 
      hwy, 
      data=mpg, 
      geom="point")



```

This is a clear positive relationship.  That makes perfect sense:  if a car is efficient in the city, of course it's efficient on the hwy.


```{r}
# try disp and hwy


qplot(displ,
      hwy,
      data=mpg)
```

This is a negative relationship.  Makes sense!  Bigger engines are less efficient.

#### Good Bad Example

```{r}

# scatterplot of RATE01 and RATE03 (academic ability and compet)

qplot(x=RATE01,
      y=RATE03,
      data = cirpData,
      geom="point")

```

Woah!  What happened?  Scatterplots don't make sense for **categorical data**!!  Can't tell how many points!

A possible fix:  jitter.

```{r}

qplot(x=RATE01,
      y=RATE03,
      data = cirpData,
      geom="point")+geom_jitter(width=1/20, alpha = 1/2)

```




## Sampling

### Population

The population is everyone (or everythign) we're interested in.

Ex:  All Americans
Ex:  All University studetns

Whoever you care about as an investigator.

### Sample

Problem:  populations are too darned big!  We can't study everyone, since we have limited money and life spans!

Solution:  Let's pick a subset of the population, and study that subset.

Another problem:  is our sample **representative**?

### Census

*To be continued!  See lec 5.*
