---
title: "Math 120 Lec 01"
author: "Ali Miller"
date: "8/30/2019"
output: 
  html_document:
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

## Intro

Warning!  I'm a fast typist.  You don't have to keep up with this!  I'll share it with you after class.

**Do** keep up with the code.

It's ok if you don't understand all of it.  In fact, there will be lots you don't understand!  That's ok!

Let's just play the game together.  With practice, you'll understand the logic.

Before we get going, we'll need our tools!  Here, we'll use the "tidverse"

```{r}

#load the library:

library(tidyverse)



```

The tidyverse has all of our most important data tools.

Ok, now we need data:

```{r}

deathPenaltyData <- read_csv("death penalty data.csv")


```
We've got our data.  Now what?

It's up to us!  What variables are we interested in?

Let's plot age:

```{r}

qplot(Age, data=deathPenaltyData)


```


Ok, let's look at Sex:

```{r}

qplot(Sex, data=deathPenaltyData)

```



