---
title: "Math 120 Lec 24"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```





## Hyp Tests (Continued)


CIs and Hyp Tests are our two big tools for "inference"

CIs are used for estimates.  What is it?

Hyp Tests are used for answering questions.  Is our support for our claim?


### Four key steps of hyp tests

#### 1) Hypotheses

Two competing hyps:

H0:  the null hyp.  This is the assumption we make about the population.  It's ALWAYS a statement of equality about a pop parameter.

Ex:  H0:  p = .5  (note that p is the pop proportion)

It's needed to compute p-val!  We can't find probabilities unless we make an assumption about pop.  That's H0.

Ha: the alt hyp.  The alt hyp is our claim about the population.  It's what we think is "up".  It's the reason we're doing the test.  We hope to support it!  

The alt hyp is always a statement of INEQUALITY.  <  >  !=

Example:  Ha:  p < .5

The alt hyp is the logical opposite of the null hyp.  Only one can be true!

#### 2) Test statistic

In short: a summary of our data.  

Example:  phat

Almost always, expressed as a z score.  How about for phat?

#####

z = (phat - p)/sqrt(p(1-p)/n)

#####

#### 3) p-val

The p-val is the probability of observing a sample that's as (or more) extreme than the one we got, assuming H0 is true.

In short, to find:  pnorm.  Find tail area past the test stat.

#### 4) Conclusion

If our p-val is small, then H0 is saying that our data is unlikely.  But of course we trust our data, not the hypothesis!

So, if p is small, we reject H0.  We have evidence to support Ha.

(Remember, "small" means "alpha", alpha = .05 by default)

If p is large (p>alpha), then we FAIL TO REJECT H0.  We didn't find the evidence we're looking for to support Ha.  We have sad feelings, our paper doesn't get published, try again!


#### Example: 5.32 p205

Do we have evidence that more than 8% of children are nearsighted.

phat = 21/194

```{r}
21/194
```

The sample result IS BIGGER than 8%.  Is this result extreme enough to convince us that the population proportion really is greater than 8%?  How likely is this sample? What human feeligns do we have?

1) **Hypotheses**

    H0:  p = .08
    
    Ha:  p > .08

2) **Test stat**

    phat = 21/194

```{r}
# find zscore.  note: center = p, se = sqrt(blaahblahs;as)


# idea:  (phat - p)/sqrt(p*(1-p)/n)

(21/194 - .08)/sqrt(.08*(1-.08)/194)


```

  (Human feelings:  Not an unusual score!)

3) **p-val**

```{r}
1-pnorm(1.450243)
```

    Note:  this is a "right-tail" test.  You can tell from Ha.

4) **Conclusion**

    Is p small?  Is p<alpha?  Is p < .05?

    No!  p is large!  We didn't find strong evidence against H0.  The sample results doesn't seem unlikely under H0.  We fail to reject H0!

    We didn't find the evidence we're looking for to suggest that more than 8% of children are nearsighted.  Nobody publishes our result.  We are sad and dejected.  
    
    
    
    
    
    
    
| var 1 |  var 2 | var 2 |
|-----  |  ----- | ---   |
| thing1 | thing2 | thing 3|







## Hyp Tests

If DU faculty want to change the faculty handbook, they need a "supermajority", 2/3 vote or more.

Prof Miller wants to gauge support for her proposal. She takes a sample of 50 faculty members, and finds 38 of them support her.

```{r}
38/50
```

Of course, that doesn't mean that 76% of all faculty support her.  Is there statistically significant evidence that a supermajority (>2/3) support her?  Conduct a hyp test.

### 1 Hypotheses

H0:   p=2/3  (we'll use this for p-val)
Ha:   p>2/3  (right tail test)

### 2 test stat

phat = 38/50

```{r}

(38/50-2/3)/sqrt(2/3*(1-2/3)/50)


phat <- 38/50
p <- 2/3
n <- 50


(phat-p)/sqrt(p*(1-p)/n)


```

### 3 p-val

```{r}
1-pnorm(1.4)
```

### 4 conclusion

Since p>.05, we fail to reject H0.

We didn't find strong evidence that the proportion who support Prof Miller is greater than 2/3.

Follow-up question:  based on the sample result phat = 38/50, construct a 95% CI for p.

```{r}
phat <- 38/50
n <- 50
zstar <- 1.96

#lower bound

phat - zstar*sqrt(phat*(1-phat)/n)
phat + zstar*sqrt(phat*(1-phat)/n)



```

We estimate the proportion of faculty who support Prof Miller between (64.2%, 87.8%).

Based upon this, is there evidence against H0?  Is there evidence to support Ha?

H0:  p = .667,   Ha:  p > .667

Based upon the CI, the H0 seems totally reasonable!  Certainly, we'd fail to reject, same conclusion from our p-val.  

Hyp Tests and CIs are connected in this way:  if the H0 value is contained in the CI, we fail to reject the H0.  Note, that alpha and confidence level must match!!  C=95%, alpha = .05

(Btw, that means our default conf level and default significance level are actually the same!)  As long as they match, CIs and p-vals will -always- give the same result!  You can conduct every hyp test with a CI!!










## Left Tail, Right Tail, and Two Tail

Left tail tests:  test for "less than", so Ha:  <

Right tail tests: test for "greater than", so Ha:  >

Two tail test:  test for "not equals", so Ha:  !=

Problem:  in two tail test, p-val is area of -both- tails.  BUT, we only have one test statistic!  

Answer:  fortunately, both tails are the same since normal is symmetric!  So, don't have to worry about which side our test stat is on.  Just find the tail area, and double it!  (Btw, this means 2-tail tests always have bigger p-vals.)

When to do two-tail?  Testing for "different" or "change".  

Example 1)  Historically, 82% of DU students have been from out-of-state.    A professor suspects/wonders that this proportion has changed.

To investigate, she takes a sample of 75 studnets, and finds that 60 of them are out-of-state.  Is there evidence to support her claim?  Test with alpha=.05

Note:

```{r}
60/75
```

1) Hypos

H0:  p = .82,  Ha:  p != .82  (two tail!!!)

2) test stat

```{r}
# z = (phat - p)/sqrt(p(1-p)/n)

(60/75 - .82)/sqrt(.82*(1-.82)/75)

```

3)  P-val

```{r}
pnorm(-.451)*2    #since 2-tail
```



4)  Conclusion

Since p>>.05, fail to reject H0.  We did not find strong evidence that the proportion of out-of-state students differs from 82%



Example 2)  In 2015, the gender ratio at DU was 57% female.  Prof Miller wonders if that ratio has changed.  She takes a sample of 80 DU students, and finds that 41 of them are female.  Did she find evidence to support her claim?  Test at alpha=.05 sig level.

1)  H0: p = .57, 
Ha: p != .57  (two tail!!!)

2)  test stat

```{r}
(41/80-.57)/sqrt(.57*(1-.57)/80)
```

3) p- val

```{r}
pnorm(-1.038822)*2    #2-tail
```

4)  Conc.  Since p>.05, we fail to reject H0.  We didn't find strong evidence that the gender ratio differs from .57.



Example 3)  Accordring to Prof Miller's terrible memory, about 18% of DU students are international students.  But, she wonders if that figure has changed from last year.  

To test, she takes a sample of 150 DU students, and finds  49 of them are international.  Test with alpha=.05

1)  Hypotheses

H0: p = .18
Ha: p != .18

2)  test stat

```{r}
(49/150 - .18)/sqrt(.18*(1-.18)/150)
```

3) p-val

```{r}
(1-pnorm(4.675563))*2
```

p = .00000293

4)  Conclusion

Since p << .05, we reject H0.  We've found strong evidence that the proportion of international students differs from .18.