---
title: "Math 120 Lec 02"
author: "[Your Name Here!]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: journal
  
---

```{r setup, include=FALSE}
###########################################################################
################# DON'T DELETE THIS PART!!!!!##############################
###########################################################################
                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans"))                                 #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #
### Below is where you read and store your data.                          #
                                                                          #
deathPenaltyData <- read_csv("death penalty data.csv")
                                                                          #
                                                                          #
                                                                          #
############# ^ ################################# ^ #######################
############# | ## DON'T DELETE THIS PART!!!!!### | #######################
############# | ################################# | #######################
```

## Let's do some basic plots

Basic qplot

```{r}

#remember, all this stuff is code!  in grey!

qplot(data=deathPenaltyData, 
      Age,
      xlab = "Age",
      ylab = "Count",
      main = "Age of executed prisoners"
      )

```

Ok, now let's look at sex:

```{r}

qplot(Sex, 
      data=deathPenaltyData, 
      xlab="Sex", 
      ylab="Count", 
      main="Sex of executed prisoners",
      fill=Race)    ### <------- this is called "faceting", super important!


```


Lastly, let's do race:

```{r}

qplot(Race, data=deathPenaltyData,
      xlab="Race",
      ylab="Count",
      main="Race of Executed Prisoners")

```


Challenge:  make a graph of State, fill by race.

```{r}

qplot(State, 
      data=deathPenaltyData,
      fill=Race)

```



Question:  can you facet by Age?

Answer: No, since Age is numerical.   It only makes sense to facet by categorical variables!  Hopefully, not too many categories.


## filters

Quick demo:  how many women are in the dataset?

```{r}

deathPenaltyData %>% filter(Sex == "Female") %>% nrow


deathPenaltyData %>% nrow


#So, what percent were women?

16/1442




```


Wow!  Only 1.1% of executed prisoners are women!

Filter challenge!  How many executions in texas?

```{r}

#problem:  how is texas spelled?

deathPenaltyData %>% head              #first few entries


deathPenaltyData %>% filter(State=="TX") %>% nrow


538/1442

```

Wow!  Over 37% of all executions have been performed in Texas!

