---
title: "Math 120 Lec 13"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```

## Special Discrete Distribution:  **Binomial Dist**

Goal for next week or so:  let's get to know the most famous probability distros.

Our first new distro:  binomial!

## Example: tossing three coins

Suppose we flip a coin three times in a row.

a)  What are all possible outcomes?  How many?

There are 8 outcomes.

b) What's the probability that we observe exactly 1 head?

Since exactly three of the outcomes have exactly one head, that's 3/8 = .375.

Why this example?  Notice that each one has equal probability.

P(HHH) = P(HTH) = P(HHT) = ..... = P(TTT) = 1/8.  So, doesn't it seem kinda wasteful of us to count out all options, when each is the same?  Must we do each seperately?

What if we tossed 5 coins?  There's 32 ways.  What if 10 coins?  There's 1024 ways!  I don't wanna list those!

Fortunately, there's a shortcut:  The binomial dist.



## Binomial Random Variables

A discrete RV, X, is "binomial" if it counts the number of success out of a fixed number of trials.

Think:  success is "heads", # of trials is 3 (in last example).

Binomial variables must satisfy 4 properties:

- Only two possible outcomes:  success and failure.
- Fixed number of trials, n.  Decided in advance.  
- The probability of success is constant, i.e never changes.  We call this probability "p".
- Each trial must be indepndent.  

For our coin example:

- H or T (only two outcomes)
- n = 3
- p = 1/2  (prob of heads is constant)
- coin tosses are independent


## The distribution

If X is binomial, there's an easy way to find probailities.  (i.e., a special formula).

Every distribution has two versions:

1) The "pdf" or "probability density function" is what you guys have learned so far:  a list of all possible options, and their probs.
2) The "cdf" shows "cumulative" probability.  I.e., prob of that value or smaller.  Everything the left.


###  binom pdf
If X is binomial, n trials, chance p of success:

$$ f(x) = \sum {n \choose x} p^x(1-p)^{(n-x)} $$

where $$ {n \choose x} = \frac{n!}{(n-x)!x!} $$

This is pronounced out loud as "n choose x".

Let's digest all the pieces:

- ${n \choose x}$ gives us the total number of ways to get x successes in n trials
- $p^x$ tell the probability of having x successes. ( $\overbrace{p\cdot p\cdot p \cdots p}^n$)
- $(1-p)^{(n-x)}$ tell the probability of having n-x failures.  $\overbrace{(1-p)\cdot (1-p)\cdot (1-p) \cdots (1-p)}^{n-x}$

For the first time ever in my class, **you won't need to use this by hand**!!  Instead, know the "story" of the formula:  what are the pieces, and what do they mean?

We don't do it by hand because it's 2019.  Let's let non-humans do it, i.e. R!  It's easy:

```{r}
#Find P(X=1) with n=3 tosses, p=1/2 chance heads

#use the dbinom(x,n,p) function

dbinom(1, 3, 1/2)


```


Yaaaay! It gave the same answer, but so much easier.

In-class challenge:  If we toss a coin 50 times, what's the chance that exactly 12 of them are heads?

```{r}
#x=12, n=50, p=1/2
dbinom(12, 50, 1/2)
```

In-class challenge 2:

If we roll a dice 100 times, what's the chance we get a "4" exactly 18 times?

```{r}
#n=100     x=18     p=1/6

dbinom(18, 100, 1/6)
```

There's about a 9.7% chance.

###  cdf

Question:  what if we're interested in a range of outcomes?

Example:  If we toss a coin 100 times, what's the probability that we get less than 60 heads?

The hard way:

P(X<60) = P(X=0) + P(X=1) + P(X=2) + P(X=3) + ... + P(X=58) + P(X=59)

Oh no!  I don't wanna do all that!

Cdfs exist to solve this problem.

F(x) = the probability of observing x **successes or fewer**.  Cdfs always tell the "left-side" probabiliy.

In R:  use `pbinom()`.  This is a common pattern:  "d" means "density", and "p" means cumulative probability.

For individual outcomes, use "d".  For ranges, use "p".

Let's find our prob of less than 60 heads:

```{r}

#n=100, x<60, p = 1/2

pbinom(59, 100, 1/2)

```

There's about a 97.1% chance of getting fewer than 60 heads.

In-class challenge 3:

What's the probabilty of getting between 50 and 60 heads (not including 50 and 60)?  Still tossing the coin 100 times.  Use the cdf, `pbinom`

```{r}
#use cdf for x<60

pbinom(59, 100, 1/2) - pbinom(50 , 100, 1/2)
```



## Examples


###  Sampling DU students


###  Multiple Choice Exam


###  Baby Tiger Snakes


## Mean and variance of binomial


###  Interpreting examples above

