---
title: "Math 120 Week 6"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

studentSurvey <- read_csv("Student Data.csv")
deathPenaltyData <- read_csv("death penalty data.csv")
housingData <- read_csv("../Data/House-Sales.csv")
#------------------------------------------------#
```

# Monday Sept 21

## Intro To Probability - Basic Examples

Recall, our goal:  statistical inference is the process of making conclusions about populations based on sample data.

Main tool:  probability!!! Ie, given our assumptions, how likely is the sample we obtained!

Start simple:  basic concepts of probability.


Ex:  suppose we toss a fair coin.  What's the probability that "heads" shows up?

Answer:  1/2.  Why?  Know:  two equally likely options.  Heads and tail.  Our outcome is exactly one of them:  heads.  So, our probability is 1 out of 2.

Ex:  suppose we roll a fair 6-sided dice.  What's the probability of rolling a 2?

Answer:  1/6.  Six equally-likely options, only one is a "2".

Ex:  suppose we roll a fair 6-sided dice.  What's the probability of getting an "odd" number show up?

All possibilities:  1, 2, 3, 4, 5, 6
Odd:  1, 3, 5

Answer:  3/6 = 50% = .5 = 1/2





## Defs of Probability

Book:  The probability of an event/outcome is the proportion of times we'd observe that event/outcome if we made an infinite number of observations.

Ex:  if we tossed a coin infinity times, we'd expect half of them to be heads.

This notion is called "theoretical" or "classical" probability.

Another way:  experimental or relative frequency probability.  Here, the probability of an event/outcome is the proportion of times that we observe it after n trials.  

Classical probability:  p (population proportion)
Experimental/Relative Frequency probability:  phat (sample proportion)


We always hope to know true/theoretical probability.  BUT, we're usually stuck with experimental probability.

BUMMER!

## Law of Large Numbers

BUT, GOOD NEWS!  "Law of large numbers":  as n (number of trials/repetitions) increases, experimental probabilities (phat) tend to get closer and closer to the true theoretical probability (p).


Deep math is required to prove this.  Easy to see it in action.

We'll run a simulation with coin tosses.

For us, we'll say "heads" = 1, "tails" = 0.

Since our outcomes are only 0s and 1s, then the average of the outcomes shows the percentage of 1s.  This is phat!
















## Addition Rule for disjoint outcomes

The word "disjoint" means the two events cannot happen at the same time.

Ie, they have no outcomes in common.  If A, B are disjoint, then there's no scenario where both A happens and B happens.

Note:  "disjoint" == "mutually exclusive".  

Ex:  The lights can be on OR off, but not both.

Ex:  A coin can be heads OR tails, but not both.

Ex: Roll a dice, get an even number vs roll a dice, get an odd number.

If A, B are mutually exclusive events, then it's easy to find the P of A OR B

If so:

      P(A OR B) = P(A) + P(B)
      
(this is the "addition rule for disjoint events")


Ex:  If we flip a coin, what's the prob that it shows heads or tails?

P(H or T)

These outcomes are disjoint, can't both happen, and each one has probability 1/2 = 50% = .5

P(H or T) = .5 + .5 = 1

Ex: Roll a dice, what's the prob that we get either "1" or "5"?

Disjoint, so:

    P(1 or 5) = 1/6 + 1/6 = 2/6 = 33.33333%
    
WARNING:  Must make sure that they really are disjoint!!!!











## General Addition Rule

Q:  What if A, B are NOT dijoint?  What's P(A or B)?

Idea:  Venn Diagram

For or, don't want to double count!  Subtract one of them!

    P(A or B) = P(A)+P(B) - P(A and B)

Ex:  Roll a dice.  What's the probability that the number shown is either even or prime?

Evens:  3 of them (2, 4, and 6)
Primes:  2, 3, 5.  3 again.

    P( even or prime) = 3/6 + 3/6 - 1/6  = 5/6

EX (in-class only)  A standard deck of 52 cards has 4 "suits"

Hearts, Diamonds, Spades, and Clubs

There are two "colors":

Hears and Diamonds are red, Spades and Clubs are black

In each suit, the cards are:

2, 3, 4, 5, 6, 7, 8, 9, 10, J, Q, K, A

The J, Q, and K are "face cards"

Ex:  If we select a card at random, what's the probability that the card is either a red card or a face card?

To answer, we need to count:

- How many reds? 26 red cards (half the deck)

- How many face cards? 12 face cards (JQK of all 4 suit)

- How many are both red and face card?  There are 6: JKQ of diamionds, and JQK of hearts.

    P(Red or Face) = 26/52 + 12/52 - 6/52 = 32/52
    
```{r}
32/52
```






## Probability Distribution

For samples, the "distribution" is just the list of all values observed.

For probability experiments, the "distribution" is a list of 

1) all possible outcomes, along with 
2) the probability of each.

Important difference:  sample dists describe what has already occurred, while probability dists describe what *may* occur.

Generally, probability distributions are represented a **table**.

Ex:  Roll a dice, record the face showing up.  Make a probability distro.

1st:  what are all possibilities?  Here:  1, 2, 3, 4, 5, 6

2nd:  Each prob is 1/6.  

| X  | P(X) |
|--- | ---  |
| 1  |  1/6 |
| 2  |  1/6 |
| 3  |  1/6 |
| 4  |  1/6 |
| 5  |  1/6 |
| 6  |  1/6 |


**Rules/Properties for Probability Dists**

1) Every probability must be greater/equal to zere, less/equal to 1.
    0 <= P(X) <= 1   for all x
    
Intuitively:  can't have "negative" probability, can have more than 100% probability.

2) sum( P(X) ) = 1  (summing over all outcomes X).  If you add all probabilities, you MUST get 1.


EX:  Consider the follwing dist:

| x  | 1 | 2 | 3 |
|P(X)| .1| .7| ? |

If P(x) is a valid probablity dist, then P(X=3) must be .2!

In the above distro, What's the P(1 or 2)?

By the addition rule:  P(1 or 2) = .1 + .7 = .8











## Complement of an Event

## Independence and the Multiplication Rule

Examples p90

## Contingency Tables - Marginal and Joint Probabilities


## Conditional Probability






