---
title: "Math 120 Week 2"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

studentSurvey <- read_csv("Student Data.csv")
deathPenaltyData <- read_csv("death penalty data.csv")

#------------------------------------------------#
```

# Math 120 Week 3

## Monday Aug 31

### Describing Two Quantitative Variables (Bivariate Data)

Bivariate data is a pair of two quantitative varibles.  Each pair, (x,y), describes the same case/individual.  Paired quarntitative variables.

Ex:  Heights and weights for students in the class.  

Goal:  investigate the relationship (if any) between the two variables.













To visualize bivariate data, we use a **scatterplot**, i.e., a point-by-point (x,y) coordinate representation of the data.


Let's look at some examples:

- mpg hwy vs cty

```{r}
mpg %>% ggplot(aes(hwy, cty))+geom_point()
```

Human feelings:  both hwy and city get bigger together!  Ie, if a car has higher hwy mpg, we expect in general it's cty mpg to be high also!


- mpg displ vs cty

```{r}
mpg %>% ggplot(aes(displ, cty))+geom_point()
```

Human feelings:  as engine size (displ) gets bigger, fuel efficiency decreases.  Makes sense!  Bigger cars need more gas!

- iris sepal length vs sepal width

```{r}
iris %>% ggplot(aes(Sepal.Length, Sepal.Width))+geom_point()
```

Human feelings: looks a blob!  It doesn't seem like I could predict a flower's sepal width based on it's sepal length.  



### Qualities of Scatterplots


Just like for histograms, there are some important qualitites about scatterplots that we need to talk about:

- Form/shape.  Is there a pattern?  Examples: linear (straight line), exponential, logistic, quadratic, etc etc.  
- Direction:  pos or neg?  Do the points seem to get bigger together (x bigger -> y bigger, going up to the right), or are they negatively related:  y gets smaller when x gets bigger.
- Strength.  How tightly do the points comform to the pattern.  









In Math 120, we'll study linear relationships in depth (even though there are others, like...)











### Correlation Coefficient

In order to measure the strength of a linear relationship, we need an objective mathematical measure.  Ie, a statistic!

We call this the linear correlation coefficient, r.

From now on, whenever you see "r" in Math 120, that -always- means "linear correlation coefficient".  

Quick example:  mpg cty vs hwy.

```{r}
cor(mpg$cty, mpg$hwy)
cor(mpg$hwy, mpg$cty)
```

The r value here is r=0.96.  This indicates a very strong, positive relationship.


### Important properties of r, the correl coeff

-1 <= r <= 1.  R can never be smaller than neg 1, never bigger than +1.

If r is close to +1, then there's a strong positive linear relationship.  If r == 1, all of the points are literall on the same line.

Smart question:  how close is "close"?  Return to this later.  For right now, understand that closer to 1 means stronger.

If r is close to -1, then there's a strong negative linear relationship.  (I.e., the line goes down).  

If r is close to zero, then there's little/no LINEAR relationship.  

WARNING:  just because there's no LINEAR relationship, that doesn't mean there's no relationship of any kind.

ALWAYS look at the scatter plot with your human eyeballs!

The correl coeff r doesn't have units.  Doesn't depend on units!  WHew!  In other words, it doesn't matter if your data is lbs/kg, ft/meters, whatever units don't affect r.  

Lastly, it doesn't matter which variable is x (ie first) and which is y (ie second)

















When talking about correlation, the x and y variables have important roles:













Important properties of *r*, correlation coefficient:

  


