---
title: "Lec 01"
author: "Ali Miller"
date: "9/01/2019"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans"))                                 #
  )       
```

## Intro

Warning!  I'm a fast typist.  You don't have to keep up with this!  I'll share it with you after class.

**Do** keep up with the code.

It's ok if you don't understand all of it.  In fact, there will be lots you don't understand!  That's ok!

Let's just play the game together.  With practice, you'll understand the logic.

Before we get going, we'll need our tools!  Here, we'll use the "tidverse"

```{r}

#load the library:

library(tidyverse)



```

The tidyverse has all of our most important data tools.

Ok, now we need data:

```{r}

deathPenaltyData <- read_csv("death penalty data.csv")


```
We've got our data.  Now what?

It's up to us!  What variables are we interested in?

Let's plot age:

```{r}

qplot(Age, data=deathPenaltyData)


```


Ok, let's look at Sex:

```{r}

qplot(Sex, data=deathPenaltyData)

```



