---
title: "Math 120 Lec 03"
author: "Ali Miller"
date: "09/04/2019"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
---

```{r setup, include=FALSE}
###########################################################################
################# DON'T DELETE THIS PART!!!!!##############################
###########################################################################
                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
library(kableExtra)                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans"))                                 #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #
### Below is where you read and store your data.                          #
                                                                          #
deathPenaltyData <- read_csv("death penalty data.csv")
                                                       

cirpdata <- read_csv("CIRP_DATA_2014_CLEAN.csv")                          #
                                                                          

                                                                          #
############# ^ ################################# ^ #######################
############# | ## DON'T DELETE THIS PART!!!!!### | #######################
############# | ################################# | #######################
```

## Summarizing Distros: Statistics

Statistics are numerical summaries of data.

Two important kinds:

1) Measure of Center: what's common
- mean is the "balancing point" of the data
- median is the "middle"
- mode is the most common

2) Measure of spread: how much variability
- standard deviation: avg distance from the mean
- iqr: width of the middle 50% (short for interquartile range)

## Summarizing Distros: Visuals

We also have visual summaries of data! 

- histogram (bar graph for one numerical variable)
- box plots (show one numerical variable)
- scatterplots (show relationship betwen two variables)

One more important quality:  shape!

- skew left means left tail
- skew right means right tail
- symmetrical means symmetrical



## Case Studies

First:  death penalty age.

```{r}

qplot(Age, data=deathPenaltyData,
      bins=10)

summary(deathPenaltyData$Age)      #useful summary!


#summary(Age, data=deathPenaltyData)  #doesn't wokr!


sd(deathPenaltyData$Age)
IQR(deathPenaltyData$Age)

```

Let's describe this data!

Center:  It's got a mean of 41.4, median of 40.  I.e., prisoners tend to be about 40 years old when they receive the death penalty!

Spread:  The avg distance from the mean (41.4) is 9.47.  In other words, on average, prisoners are within 41.4 +- 9.47 years old.  The most common half of the ages are between 34 and 47.

Shape: We see some right skew (tail), in other words: sometimes very old people get the death penalty, but rarely (never) do very young people get the death penalty.


The above shows the bones for describing data!

## In-class exercise


Look at the iris dataset:

```{r}
iris %>% head 



```


First, let's look at sepal length:

```{r}
qplot(Sepal.Length,
      data=iris,
      bins=15,
      fill=Species,
      geom="histogram")

summary(iris$Sepal.Length) 


```

It looks like sepal lengths are 5.843cm in legnth on average.  The iqr is 6.4 - 5.1 = 1.3.  I.e., if we look at the middle 50% most common flowers, they have sepal lengths between 5.cm and 6.4cm.

Shape:  Looks like there's overall right skew, and possibly bimodal (i.e., two peaks).  




