---
title: "Math 120 Week 01"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```





 
 
# Tuesday 
 
## I write most of our notes in RStudio

This will be posted online in its entirety.  I'll update it in the afternoon after each class period.

Don't worry about taking word-for-word notes.  You're not a stenographer!  Instead, summarize your thoughts and reactions to help you study later.

I'll also use the board for notes sometimes. **Be sure to incorporate boardwork into your written notes!!**


## Goals for Tuesday:

- What is a dataset?
- What do we want to say about data?

Example:  the built-in `mpg` dataset.  It's about cars.

```{r}
head(mpg)
```

Generally, we'll always work with "tidy" data.  A tidy dataset is a grid/matrix/rectangle/table of data that satisfies:

- Every row is an "observation" or "case".  In mpg: each case is a specific car!  
- Every column is a variable, i.e. a specific quality of our observations.
- Each cell contains exactly one measurement

Even if we have tidy data, it's still hard (sometimes) to understand what it means!

Answer:  metadata.  Metadata is "data about data".  It explains cases, variables, and units.  

## Types of data

Different kinds of data require different tools to analyze!  Gotta know!  

BIG DIVIDE:  number or not.  Each can be subdivided.

Quantitative data can be discrete or continuous.  

**Discrete**:  Whole units.  Countable!!!!!  Ex:  number of siblings.  Can't have 3.65 siblings.  Messy.  

**Continuous**:  Can take any value on the real line.  Usually, these are physical measurements:  height, length, duration, volume, etc etc.  Infinite decimals.   Example, height:  someone could be 5'6",  someone could 5'7", and there are an infinite number or possibilities between.  

Categorical data can also be divided into two kinds:

**Nominal**:  No natural ordering.  Ex:  hair color, nationality.  

**Ordinal**:  There is a natural, intrinsic ordering.  

- Age:  it counts something.  Probably discrete, not ordinal.

- Year in school:  fresh, soph, jun, sen.  Natural order. 

- Clothing size:  S, M, L, XL  Natural order. 

- Month of the year:  Jan, Feb, March..... natural order. ordinal.  

## Notes about `mpg` data

- "Year" is ordinal.  Even though they're numbers, just a label.  Like month or day of the week.  It is a little blurry!  In some sense, similar to discrete.  

- Note for `cty` and `hwy`:  Decide discrete vs continuous by what they *are*, not how we rounded.  These are **continuous**!

## What do we say about data?

Always first:  visualize!!!  Warning:  the options for visualization depend on the type of data!

## Quantitative visualization:  histograms

Example:  let's make a histo of the hwy variable from the `mpg` dataset.

```{r}

#note:  $ choooses variable

hist( mpg$hwy )

```

Observations:

- Bimodal.  Two groups:  one less efficient, one more.  
- Right skew (right tail).  I.e., most cars have relatively low mpg, and only a few have very high mpg.
- What's "normal"?  Looks like 25ish is a normal mpg.
- Spread?  Variability?  All are btwn 10mpg and 45mpg.  Most are between 15 and 35mpg.  
