---
title: "Math 120 Lec 10-11"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes) 
library(kableExtra)

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```

## Understanding Contingency (Two-Way) Tables

Why two-way tables?  They're great for humans!

The let us compare two categorical variables.

Ex: reading discussion data.

### Example Problems (Basics)

If a a photo is selected at random, what's the probability...

a) ... the photo was correctly identified as a fashion photo?

= 197/1822


b)  ... the photo is a fashion photo?

=309/1822

c)  ... computer thought it was a fashion photo?

219/1822


d) ... it's a fashion photo -or- the computer thought it was a fashion photo.


(197+22+112)/1822


```{r}
(197+22+112)/1822
```

For scientists, mathematicians, computer scientist, and everyone else, "or" is "inclusive or".  I.e., could be one, could be the other, could be both.

## Making Contingency Tables in R

It's easy!  Just use the `table()` command!

Example:  mpg data.  Two way table for class and drive shaft (drv).

```{r}
#make a table fofr class and drv
table(mpg$class, mpg$drv)
```

Btw, it's also easy to get sums/total


```{r}
table(mpg$class, mpg$drv) %>% addmargins()
```



## Mutually Exclusive/Disjoint Events

The mean what you think they do:  it's impossible to be in both events!

I.e., the number of outcomes in both event A and event B is zero!

One more way:  The probability P(A and B) = 0.

Ex:  Are the events "is a pickup" and "has front wheel drive" mutually exclusive?

Answer:  since there are 0 pickups with front wheel drive, they **are** mutually exclusive/disjoint.

### The "or" rule for mutually exclusive events


If A, B are disjoint, then:

$$  P(A or B) = P(A) + P(B) - P(A and B)        $$

Note:  since A, B are disjoint, the "and" prob is zero!  Thus,

$$  P(A \cup B) = P(A)+P(B)   $$

Use this on the car example:  what's the probability that a randomly selected car is either a pickup -or- has front wheel drive?  We already know they're mutually exclusive.  Use the rule!

```{r}
table(mpg$class, mpg$drv) %>% addmargins()
```

P(Pickup or FWD) = P(Pickup)+P(FWD)-P(Pickup and FWD)
                 = 33/234   + 106/234 - 0/234

```{r}
33/234   + 106/234 - 0/234
```
                 
                 
Challenge:  draw Venn for disjoint events:

(see board, two separate circles.)

## Conditional Probability

Conditional:  we know that some other event has happened.

P(A | B) = "the probability that A happens, given that B has occurred"

The given condition comes after the bar.

P(Rains | Cloudy outside) = Probability of rain, given that we see it's cloudy outside.

For the formula P(A|B), we know B happened, care about A:


P(A|B) =  P(A and B) / P(B)

that's it!  "And" on the top, "Given" on the bottom.


## Independent Events

Reading example:

P(Rept) = .3, P(Green) = .2, P(both) = .1

Since .3*.2 is not equal to .1, these are not independent events.

Important considerations:

- this is the easiest independence problem we can do
- independence is a mathematical concept, not based on feelings.  To verify independence, math has to happen!

The idea is what you expect:  two events A and B are independent if they don't affect the likelihood that the other occurs.

Mathematically:

P(A|B) = P(A), i.e., doesn't matter whether B occurred or not.  Doesn't change P(A).

-or-

P(B|A) = P(B)



### The "and" rule for independent events

If A and B are independent,

P(A and B) = P(A)*P(B)

In other words, for independent events, "and" means "times".

Example:  coin tosses are independent.

What's the probability that you get 3 heads in a row?

Answer:  P(H and H and H) = P(H)*P(H)*P(H) = 1/8 = .125

We roll two dice.  What's the probability that the first shows a "4" and the second shows an even number?

P("4" and even) = P("4") * P("even")
                = (1/6)*(3/6) = whatev
                
            


## Checking Independence

You *can* use the definition, but it's usually easier to use multiplication rule.  If it works, then yes, independent!




### Example:  mpg drv and class

First, let's make a contingency table:

```{r}
#use table() command:

table(mpg$drv, mpg$class) %>% addmargins()



```

Question:  are the events "has 4 wheel drive" and "is subcompact" independent?

Need some answers first:

- P("4 wheel drive") = 103/234

- P("is subcompact") = 35/234

- P( both) = 4/234

Check: if multiplied, do they match the "and"?

```{r}
103/234*35/234

4/234

```

Don't match, so **not** independent.  I.e., there's a dependence relation between them.  

In-class challenge: are the events "compact" and "front wheel drive" independent?


```{r}
table(mpg$drv, mpg$class) %>% addmargins() %>% kable %>% kable_styling(bootstrap_options = "striped", full_width = F)
```

P(compact) = 47/234

P(f) = 106/234

P(compact and f) = 35/234

Check independence with multiplication rule:

```{r}
47/234 * 106/234

35/234
```

Since they don't match, they're not independent.  


In-class challenge #2:

Are the events "has 4 cyl" and "is front wheel drive" independent?   Note!  You'll need to make your own table!

```{r}

table(mpg$drv, mpg$cyl) %>% addmargins() %>% kable %>% kable_styling(bootstrap_options = "striped", full_width = F)
```


P(4 cyl) = 81/234

P(front) = 106/234

P(both) = 58/234

```{r}
#check

81/234*106/234

58/234


```

Answer:  nope, not even close.  Definitely not independent. 

Question:  what if they're close?  Do they have to be "exactly equal"?

Two answers: 

- for right now, they need to be exactly equal. (Quiz 4!)
- We'll return to this later when we know "hypothesis tests".  In reality, a little bit of wiggle room is of course acceptable.





