---
title: "Math 120 Lab 6: Categorical Inference"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    theme: cerulean
    df_print: paged
  
---

```{r setup, include=FALSE}
###########################################################################
################# DON'T DELETE THIS PART!!!!!##############################
###########################################################################
                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans"))                                 #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #
### Below is where you read and store your data.                          #
                                                                          #
politicalData <- read_csv("political-data.csv")                                                                          #
library("pander")    
options(digits=5, scipen=99999)

#
                                                                          #
############# ^ ################################# ^ #######################
############# | ## DON'T DELETE THIS PART!!!!!### | #######################
############# | ################################# | #######################
```



## Intro

This is a solutions outline, not inteded to be an example of a full essay with discussion.

## Part 1: The Whole Sample

First, let's take a look at party distribution for the whole dataset:

```{r}
politicalData$Party %>% table  %>% prop.table %>% pander
```

Looks like democrats are the majority party for this sample, with $\hat{p}_D=0.53$.  Does this sample provide statistically significant evidence that Democrats are the majority party for the entire population?  Let's run the test.  Note that since there are (many!) more than 10 Dems and 10 Repubs in the sample, the CLT is satisfied and we may use the normal distribution for describing $\hat{p}$.  Whether or not this is truly an SRS is difficult to say without knowing more about the sampling method.  It's possible that the sample suffers from undercoverage or sampling bias.


```{r include=FALSE}
nrow(politicalData)
(.5328-.5)/sqrt(.5*(1-.5)/3906)
```


|  Test Component  | Result |
| --- | --- |
| Null Hypotheses| $H_0: p = .5$|
|    Alt Hypothesis      |  $H_a: p>.5$|
| Test Stat | $z= \frac{.5328-.5}{\sqrt{\frac{(.5)(1-.5)}{3906}}}=4.1$
| $p$-val | $p=$ `r 1-pnorm(4.1)` |
|Conclusion | Rejet $H_0$ |

Here, the p-val shows us the pobability of obtaining a sample proportion of 53.28% Democrats (or higher) if the true proportion is 50%.  Since this is (very) unlikely, we reject the hypothesis that p=.5; there's strong support for the claim that more than 50% of voters in this population are Democrats (again, recall that our population is merely the users of the "Show of Hands" app, not all Americans).

We can estimate the proportion of democrats using a CI.  I choose a confidence level of 95% to match with my use of $\alpha = .05$ above:

```{r include=FALSE}
.5328-1.96*sqrt(.5328*(1-.5328)/3906)
.5328+1.96*sqrt(.5328*(1-.5328)/3906)

```

$$ .5328\pm 1.96\sqrt{\frac{.5328(1-.5328)}{3906}} \longrightarrow (.517, .548)$$

We're 95% confident that the population proportion of Dems is between 51.7% and 54.8%.  This agrees with our hypothesis test:  it's very unlikely that the proportion of Dems is 50%, and we support the alternate hypothesis that more than half of voters in the population are Democrats.

The margin of error is simply the width of the interval:

```{r include=FALSE}
1.96*sqrt(.5328*(1-.5328)/3906) 
```

$$ 1.96\sqrt{\frac{.5328(1-.5328)}{3906}} \longrightarrow 0.156$$

I.e., our estimate of $p$ could be off by as much as 1.56%.  To narrow this to a 1% MOE, we'd need to use a sample size of at least:

```{r include=FALSE}
.5328*(1-.5328)*(1.96/.01)^2
```

$$(.5328)(1-.5328)\left(\frac{1.96}{.01}\right)^2  \longrightarrow 9562.7$$

Thus, we' need a 9563 people in an SRS to estimate the proportion of Dems to within 1% error.  That's a lot, but at least we know what's needed for such a high degree of precision!




## Part 2:  Educational Subgroup Analysis

First, let's remind ourselves of the options:

```{r}
unique(politicalData$EducationLevel) %>% pander
```

I'll make a comparison between people with Bachelor's Degrees and those with Doctoral Degrees.

First, we filter:

```{r}
bachelorData <- politicalData %>% filter(EducationLevel == "Bachelor's Degree")
doctoralData <- politicalData %>% filter(EducationLevel == "Doctoral Degree")

```

And, now we can look at party breakdown for each group:

```{r}
bachelorData$Party %>% table %>% prop.table %>% pander
doctoralData$Party %>% table  %>% prop.table %>% pander
```

Looks like Dems are the majority party for each group.  Let's estimate with a couple of CIS.  First for those with Bachelor's degrees:

```{r include=FALSE}
.548 - 1.96*sqrt(.548*(1-.548)/1084)
.548 + 1.96*sqrt(.548*(1-.548)/1084)
```

$$ .548\pm 1.96\sqrt{\frac{.548(1-.548)}{1084}} \longrightarrow (.518, .578)$$

We're 95% confident that between 51.8% and 57.8% of all respondants with Bachelor's degrees are Dems.  



For the doctoral group:

```{r include=FALSE}
.5621 - 1.96*sqrt(.5621*(1-.5621)/169)
.5621 + 1.96*sqrt(.5621*(1-.5621)/169)
```

$$ .5621\pm 1.96\sqrt{\frac{.5621(1-.5621)}{169}} \longrightarrow (.487, .637)$$


We're 95% confident that between 48.7% and 63.7% of all respondants with Doctoral degrees are Dems.  Notice how much larger the margin of error is for this group:  there are only 169 PhDs in the sample ,but there are 1,084 college grads.  We paid for that smaller sample size with larger MOE!

Let's investigate the difference between the two groups with both a CI and a Hyp test, in that order:

```{r include=FALSE}
.548-.5621 - 1.96*sqrt(.548*(1-.548)/1084 + .5621*(1-.5621)/169)
.548-.5621 + 1.96*sqrt(.548*(1-.548)/1084 + .5621*(1-.5621)/169)

```

$$ .548-.5621\pm 1.96\sqrt{\frac{.548(1-.548)}{1084}+\frac{.5621(1-.5621)}{169}} \longrightarrow (-0.095, 0.066)$$


We're 95% confidenct that the difference in proporiton of Dems between grads and PhDs is between -9.5% and 6.6%.  Since it's possible the difference is zero, we haven't found evidence of a difference.

Now, the hyp test:

|  Test Component  | Result |
| --- | --- |
| Null Hypotheses| $H_0: p_1 - p_2 = 0$|
|    Alt Hypothesis      |  $H_a: p_1 - p_2 \neq 0$|
|  Pooled Estimate | $p_{\text{pool}}=\frac{594+95}{1084+168}=.55$|
| Test Stat | $z= \frac{.548-.5621}{\sqrt{\frac{(.55)(1-.55)}{1084}+\frac{(.55)(1-.55)}{169}}}=-0.343$
| $p$-val | $p=$ `r pnorm(-.343)` |
|Conclusion | Fail to reject $H_0$ |

As expected, the test give the same result as the CI.  Here, there's a 36.57% chance of observing a sample difference as large as the one we saw if, in fact, the two populations have equal proportions of Dems.  Either way, the conclusion is the same:  there's not a statistically significant difference in proportion of Dems between the two educational groups.

Of course, our test coud have made an error (perhaps this was just a very unusual sample!).  Here's the rundown:

- If we'd made a Type I error, we'd have thought that there **is** a difference in proportion of Dems between bachelors and doctors, when in fact there isn't.
- If we'd made a Type II error, we'd have concluded that there's no difference in proportion of Dems between bachelors and doctors, when in fact there is.

Is either one worse?  Hard to say; it's hard to imagine a scenario where money is lost or human beings are harmed either way.




