
---
title: "Math 120 Fall 2020 Homepage"
date: "Updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: yeti

      
  
---

```{r setup, include=FALSE}

                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans")) + 
  theme(axis.text.x = element_text(angle=-90))                            #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #

```


## Video Lectures

These are all on [our Google Drive shared folder](https://drive.google.com/drive/folders/1DUJNs5i-88NPDnLlGW04BCl0NkWm700S?usp=sharing)

The work I do in those vids is availabe on this site below.  



## Math 120 Lecture Notes
*They look great on your phone, so you can read them anywhere.  Steal my code!  Don't understand something?  Ask me!*

- [Week 1](https://millera2.github.io/Math-120/Math-120-Lec-1.html)
- [Week 2](https://millera2.github.io/Math-120/Math-120-Lec-2.html)
- [Week 3](https://millera2.github.io/Math-120/Math-120-Lec-3.html)
- [Week 4](https://millera2.github.io/Math-120/Math-120-Lec-4.html)
- [Week 4](https://millera2.github.io/Math-120/Math-120-Lec-4.html)
- [Week 5](https://millera2.github.io/Math-120/Math-120-Lec-5.html)
- [Week 6](https://millera2.github.io/Math-120/Math-120-Lec-6.html)
- [Week 7](https://millera2.github.io/Math-120/Math-120-Lec-7.html)
- [Week 8](https://millera2.github.io/Math-120/Math-120-Lec-8.html)
- [Week 9](https://millera2.github.io/Math-120/Math-120-Lec-9.html)
- [Week 10](https://millera2.github.io/Math-120/Math-120-Lec-10.html)
- [Week 11](https://millera2.github.io/Math-120/Math-120-Lec-11.html)



## Next Quiz: This Wednesday

Coverage:  Last Week

Topics:  Confidence intervals.

- Constructing CIs
- the meaning of confidence
- interpreting a CI

Practice Problems:

1) Suppose in a sample of 80 DU students, we find that 33 of them prefer Cheerios to Frosted Flakes.  Estimate, with 85% confidence the proportion of all DU students who "prefer Cheerios".

    
        
2)  If we wanted to estimate the proportion of all DU students who prefer Cheerios with 98% confidence and MOE no greater than 5%, how many students would we need to include in our sample?

3)  If we constructed a 95% CI using the sample data in #1, what would be the MOE?



4)  If we increase confidence level, what happens to MOE?  Is that good or bad?


        

5)  If we increase sample size, what happense to MOE?  Is that good or bad?


        




**Solutions**  

Look away!  No, don't look yet!

Oh please, no, please try the problem first!

Ahhhhhh!!!!

Ok, here they are:

### 1
    
    middle 85% -> tail area = 15/2 = 7.5%
    
    -> zstar = 1.44
               
    -> 33/80 +/-    1.44*sqrt(33/80*(1-33/80)/80)
     

```{r}
33/80 - 1.44*sqrt(33/80*(1-33/80)/80)
```

```{r}
33/80 + 1.44*sqrt(33/80*(1-33/80)/80)

```

**We're 85% confident that between 33.3% and 49.2% of all DU students prefer Cheerios to Frosted Flakes.**

### 2

    98% confidence -> 1% tail area -> zstar = 2.33
    
    n >= (33/80)*(1-33/80)*(2.33/.05)^2
    
```{r}
(33/80)*(1-33/80)*(2.33/.05)^2
```

**We'd need at least 527 students.**

### 3 

    95% -> zstar = 1.96
    -> MOE = 1.96*sqrt(33/80*(1-33/80)/80)
        
```{r}
1.96*sqrt(33/80*(1-33/80)/80)
```
        
**The MOE would be 10.8% **       


### 4 
        
Bigger confidence level means BIGGER MOE.  This is because we'd need to widen our interval to include more of the sampling distribution.  This is BAD -- we want to be confident, but we DON'T want to have large MOE!

### 5

Bigger n means smaller MOE.  This is because n is in the *denominator* of the MOE formula.  This is GOOD!  If we're able to increase our sample size, we can decrease MOE.

## HW Assignments

[Click here for HW assignments](https://millera2.github.io/Math-120/HW.html)



## Syllabus

[Math 120 Fall 2020 Syllabus (Miller)](https://millera2.github.io/Math-120/Math-120-Fall2020-Miller.pdf)

## Office Hours 

### Prof Miller's Hours (Google Meet)

[Google Meet Room](https://meet.google.com/wtr-funk-wph)

Hours: Most weekdays before 4pm.

If you want to talk, shoot me an email (always use Denison email addresses) and propose a couple of times that work for you.

Of course, I don't check email in the evening, so we'll have to meet the next day in that case.

So, **don't wait till the night before a deadline to get help!**  

I like chatting with students -- come find me!

I think this will be better than setting aside specific times that work for everyone.



### TA Hours

*TBA*




## Shared Google Drive Folder:

[You can find all of our datasets here](https://drive.google.com/drive/folders/1tJFLMDFflBiXDHidWoTZU0SjzVi_z3zW?usp=sharing)




## Texbook Links

- [Math 120 Textbook](https://www.openintro.org/stat/textbook.php?stat_book=os)
  - *OpenIntro Statistics* is an excellent, free introduction to modern statistical analysis.
  

