---
title: "Math 120 Lec 15"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```


## Reading: Name That Distro!

Binomial, Poisson, Geom, or none of the above?  If yes, compute!  If no, explain why not.


1) According to the most recent figures from the Center for Disease Control (the CDC), 15.5% of adult Americans (age 18 and up) are regular smokers.  If we take a sample of 50 Americans, what's the probability that at most 1/2 but at least 1/4 of them are smokers?

We know it can't be geom or pois because n=50.  We notice there's a constant p = .155, there are two outcomes (smoking or no) and they're independent because big pop.

Yes, binom!

```{r}

#range of vals, so cdf

pbinom(25, 50, .155) - pbinom(12, 50, .155)

# = P(X<=25) - P(X<=12)

```


2) According to the RunKeeper app on Prof Miller's smartphone, she'd gone running 58 times in 2019 as of Sept 30.  You can google to find out that this is a time span of 39 weeks.  Assuming that her running habits don't change for the rest of the year, what's the probability that she goes running at least at least 3 times in the next 2 weeks.

Fixed number of trials?  No, not binomial.  (Note:  a week is a span of time, not a "trial".)

Fist success?  No, not geom.

Thinking poisson.  Are we counting the number of occurences over a unit of time or space?  Yes!!!  poisson!

Asking: P(X>=3).  Still need lambda.  Question:  on average, how many runs per week? 58/39

```{r}
58/39
```

Avg of 1.48179 per week.  We're talking about two week units, so times 2!

```{r}

1.487179*2
```

Got everything.

```{r}
#using lambda from above

1- ppois(2, 2.974358)

```

Note:  for poisson, we always have to do 1-(prob) whenever we need "at least" or "greater than"!



3) What strangely warm, dry weather we've had this fall!  In Granville, the wet season spans April through August; during those months, the chance of rain is about 30% on any given day. During September through March, the chance of rain is about 19%.  If we took a random sample of 40 days throughout the year, what's the probability that it rains on at least 10 of them?  

Note:  fixed n=40, so not geom or pois.  Also not binom, since p changes!

None!

4) At present, about 55% of Denison students are women.  If we sample students at random (say by choosing at random from the Registrar's records), what's the probability that 6 students sampled until we finally get a female student?  

Question theory: has to be geom.

Counting till the first success, independent, success or fail, and contsant prob p.

Note:  there's a little wrinkle in how dgeom works.  I'm not going to include it on the quiz.  So, we'll stop here.


With that, we've finished our discussion of discrete distros!! Yaay!

Next: continuous!


## Continuous Distros

Cts distros are weird!!!!  THey work different!!! Intuition fails!!

First, let's think of the simplest:  **the uniform distro**!


Example:  Suppose the amount of time it takes students to complete an exam follows a uniform dist between 40 and 60min.


Why cts?  Time is a continuous variable, and so students could take any conceivable amount between 40 and 60; i.e.

X = 45
X = 45.231
X = 51.2342345234234234

There are an infinite, uncounable number of outcomes for X.  Any real number between 40 and 60 in this case.

Interesting side note:  not countable.  Here's why.  For any two outcomes of X, it's alwyas possible to find a third that's in between them.

I.e., X=45.12345 and X=45.12346, then there are lots of values between, like X=45.123455.  Infinitely subdividable!  Can't make a list!

Becasue of this, the distro has to work differently.

Discrete:  a distro is a list of X vals and probs!

That's impossible for cts variable.  Instead, we express probability as AREA UNDER THE PDF!!!

Example:  In the exam distro, what's the probability that a student takes between 45 and 50 min on the exam?  See picture on board.

For uniform, regions are alway rectangles!  SO, multiply!

Width: 5
Height: 1/20

Prob = area = 5*(1/20) = 1/4 = 0.25.

The end!  There's a 25% chance that student takes between 45 and 50 minutes.


Example:  What's the prob it takes a student more than 52 minutes?

Helpful to draw a picture.

Width: 8
Height: 1/20

Prob = area = 8*1/20 

```{r}
8*1/20
```

In general, if a uniform distro ranges from a to b (a is the min, b is the max), how tall is the rectangle?

With discrete distros, sum of all probs is 1.


Same for cts, but here, prob is area.  Total area under pdf must be 1 !!!

Thus, the height of the uniform pdf is 1/(b-a) always.  If you know the bounds, you know the whole distro.

Exam example:  What's the prob that it takes a student **exactly** 50.000000000000-... minutes?

