---
title: "Math 120 Week 6"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

studentSurvey <- read_csv("Student Data.csv")
deathPenaltyData <- read_csv("death penalty data.csv")
housingData <- read_csv("../Data/House-Sales.csv")
#------------------------------------------------#
```

# Monday Sept 21

## Intro To Probability - Basic Examples

Recall, our goal: statistical inference is the process of making mathematically-supported conclusions about populations based on sample data.

Probability lets us do this!  It's the tool we use to make conslusions about our sample data.

Key question:  given our assumptions, how likely is our sample data?

This is pretty tough; first let's look at some basics.

Ex:  A fair coin is tossed.  What's the probability that heads shows up?

Answer:  50% = 1/2 = .5.  Why?  There are two equally likely outcomes, we're interested in one of them.  1/2

Ex A fair 6-sided dice is tossed. What's the probability that a "2" shows up?  

Answer:  1/6.  There are six equally-likely outcomes, we're talking about one specific one ("2"), thus 1/6.


Ex:  A fair 6-sided dice is tossed.  What's the prob that an even number shows up?

Answer:  3/6 = 1/2 = .5 = 50%.  Of the six possibilities (1,2,3,4,5,6), there are three evens (2,4,6).  Thus 3/6






























## Defs of Probability

Book:  The probability of an outcome/event is the percentage/proportion of times we'd observe that outcome/even if we made infinitely many observations.  (performed the experiment infinite times).


Ex: if we tossed a coin infinity times, then half of those would be heads.


This is just like "p", the population proportion.

This notion of probability is also called "classical" or "theoretical" probability.  


Another notion of probability:  experimental/relative frequency probability.  It's the proportion of occurances of our event/outcome after n repetitions of the chance experiment.

Experimental probability is just like "phat", the sample proportion.  

In probability, we HOPE to know the theoretical probability, but we're always stuck with the experimental probability.

BUMMER!

GOOD NEWS!  The law of large numbers states that as the number of repetitions/observations (n) increases, the obsererved experimental probabilities (phat) tend to get closer and closer to the "true" theoretical probability.

It's easy to see the law of large numbers in action.

Coin Toss Simulation!

We'll say that "heads" is 1, tails is "0".

















## Addition Rule for disjoint outcomes


Two events, A and B, are "disjoint" if the can't both happen at the same time.

There are no possible outcomes that are in both A and B.

Ex:  The lights are either on or off.

Ex:  A coin shows either heads or tails.

    P(A and B) = 0
    
No chance of both!

Note:  "disjoint" == "mutually exlcusive"

"Addition rule for disjoint outcomes" says:

If A, B are disjoint, then
  
    P(A or B) = P(A) + P(B)
    

Ex:  We flip a coin.  What's the probability it shows either heads or tails. 

P(H) = .5
P(T) = .5

    P(H or T) = .5 + .5 = 1 = 100%
    
Ex:  We roll a dice.  What's the prob that the face showing up is either a 1 or a 5?


    P(1 or 5) = 1/6 + 1/6 = 2/6 = 1/3 = 33.333333333%
    
    
WARNING:  This only works for DISJOINT events!!!






## General Addition Rule

If two events A, B are -not- disjoint, what then??  How can we find the P(A or B)?

Intuition:  Venn Diagram

General addition rule:  Don't want to double count!

    P(A or B) = P(A) + P(B) - P(A and B)
    

Ex:  Roll a 6-sided dice.  What's the probability that the number on top is either an even number or a prime number?

WARNING:  not disjoint!  Use general addition rule:

Even:  2, 4, 6 (three of them)
Primes: 2, 3, 5 (three of them)
Even and prime:  2 (one of them)

    P(Even or Prime) = 3/6 + 3/6 - 1/6 = 5/6
    
Ex:  In a standard 52-card deck, there are 4 suits:

- hearts, diamonds, clubs, spades

Each suit has 13 cards:

- 2, 3, 4, 5, 6, 7, 8, 9, 10, J, Q, K, A

Where J, Q, K are "face cards"

There are two colors:  

- the hearts and diamonds are red
- the clubs and spadeas are black


Ex:  Addition rule.  If we randomly select a card from the deck, what's the prob that the card is either RED or a FACE CARD?

WE know:

    P(Red or Face) = P(Red) + P(Face) - P(Red and Face)
     
- Reds:  26 of them (13 hearts, 13 diamonds, ie half the deck)
- Face: 12 of them ( 3 in each of the 4 suits)
- Red AND Face: 6 of them (JQK of hearts and diamonds)

So:

    P(Red or Face) = 26/52 + 12/52 - 6/52 = 32/52
    
```{r}
32/52
```





    
    







## Probability Distribution

A sample distribution is just a list of all the values we observed in our data (for a variable).

A probability distribution is a list of 

1) all possible outcomes, and 
2) the probability for each one.

These are commonly represented as tables.

Important difference:  sample dists describe what *did* happen, while probability dists describe what *may* happen.


Ex:  We roll a dice, record the face showing up.  Construct a probability distro.

| X  | P(X) |
|----|------|
| 1  |  1/6 |
| 2  |  1/6 |
| 3  |  1/6 |
| 4  |  1/6 |
| 5  |  1/6 |
| 6  |  1/6 |

**Rules/Properties** for probability distributions:

All probability distributions must satisfy the following:

1)  0 <= P(X) <= 1 for all X.  Ie., all probabilities must be at least 0, no greater than 1.  
2)  If we add all probabilities, must get 1:  sum(P(x)) = 1 (for all x)

Ex:  Suppose we have the following probability dist:

| X   |  1  |  2  |  3  |
|P(X) |  .7 |  .2 |  ?  |

Must be the case that P(3) = .1, since all add to 1.


Ex:  In the above dist, what's P(1 or 2)?

Disjoint events, so:

    P(1 or 2) = P(1)+P(2) = .7+.2 = .9
    



























Def and rules

## Complement of an Event

## Independence and the Multiplication Rule

Examples p90

## Contingency Tables - Marginal and Joint Probabilities


## Conditional Probability






