---
title: "Math 120 Week 11"
author: "Prof. Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: cerulean
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```


# Week 11 Vid 1




## Hyp Tests

Last week, we focused on Confidence Intervals.  CIs were the first tool for statistical inference we've learned in this class:  they let us **estimate** a population parameter based upon a sample result.

This week, we'll be learning the other major tool for statistical inference:  **hypothesis tests**, or "hyp tests" for short.  Hyp tests are formalized procedures for **asking questions** about a population parameter.

**There's a lot to take in at first**, but just like CIs, **all hyp tests have the same structure.**  Once you get that down, you understand how all hyp tests work.

These might seem intimidating at first, but hang in there!  With a little practice, they're really not too tough.

I think the easisest way to understand hyp tests is to compare them to modern **criminal trials**.  Let's dive in!







## Hyp tests are criminal trials


Hypothesis tests are just like criminal trials.  The two are philosophically identical procedures.  We'll break the process down into four parts:

1) In a criminal trial, we wonder:  innocent or guilty?

In a hyp test, the NULL HYPOTHSIS (H0) is on trial.  The null hyp is a statement about the population.  The alternate hypothesis is logical opposite.  (Ha).

In a criminal trial, we assume innocent until proven guilty.  In a hypothesis test, we assume H0 is true, until we find evidence that's it's false.

The whole reason we're doing the test is becasue we suspect H0 is false!  I.e., the alt hyp (Ha) is our claim about the population.













2) In a criminal trial, we have witnesses.  In a hyp test, we have sample data!  We'll use special statistics, called "test statistics", to summarize the evidence.


3)  In a criminal trial, there are laws and lawers and arguments.  In a hyp test, there's probability!  (p-value)

A p-value is the probability of observing a sample result as or more extreme than the one we got, assuming H0 is true.  The assumption is important: we can't compute probabilities unless we make an assumption about the population!

Finding p-vals are identical to the **forwards problems** for sampling distributions that we practiced in Week 8 (the first post-apocalypse week).












4) In a criminal trial, a judge or jury weighs the strength of the evidence.  In a hyp test, we decide if our p-value is SMALL ENOUGH.  

If p is small, it means our sample result is unlikely under H0.  In this case, we REJECT H0, we trust our data!  Small p-val -> reject H0.  Support Ha.

Every hyp tests ends in one of two ways:  if p is small, we reject H0.  If p is large, we fail to reject H0.  

Wait, how small is "small"?   Totally up to us!  It's called the "significance level", "alpha".  Generally, we use alpha = .05 by default.  In other words, if our sample has less than 5% chance of occuring, we reject H0.

















## Example 1:  Hyp test for proportion

In Fall 2015, my stats class had 16 women out of 24 students.  Based upon this, I suspect that the gender ratio at DU was higher than 50% female.  Does this sample result support my claim?  Perform a hyp test.  Use alpha=.05.


Note:  phat = 16/24


1)  Hypotheses.  

H0: p = .50  (Assume by default that there's not an imbalance)

Ha: p > .50  (Think:  proportion of female students is greater than 50%)

2)  Test stat.  Think:  z score!

z = (obs-exp)/stdev

obs:  phat = 16/24

exp:  p = .5

stdev:  sqrt(.5*(1-.5)/24)


```{r}
(16/24 - .5)/sqrt(.5*(1-.5)/24)
```

3)  p-val:  Probability of getting a sample result as or more extreme, assuming that p = .5.

P(Z>1.63) = 1-.9484

```{r}
1-.9484
```

Here, p = .0516.  

4)  Remember, if p is small (p<.05), then reject H0.  Otherwise, if p>.05, we fail to reject H0.

Here it's close, but still .0516>.05, so we fail to reject the H0. 

We didn't find strong evidence to suggest that the proportion of female students at DU is greater than 50%.














## Example 2:  Hyp test for proportion

After a night of bad luck gambling, we suspect that dice is unfair.  Specifically, we think it rolls a 1 more frequently than it should (1/6 of the time).

To test this, we rolled the dice 100 times, and found that 25 of them were 1s.

Does this provide evidence that the proportion of "1" rolls is higher than expected?


1)  H0:  p = 1/6, Ha: p > 1/6

2) test stat

```{r}
(25/100-1/6)/sqrt(1/6*(1-1/6)/100)
```

3) p-val:

Look up on table,
```{r}
pnorm(2.24)
```

Thus p = 1-.9875

```{r}
1-.9875
```

4)  Since p<.05, reject H0.  There's strong evidence that the proportion of "1"s rolled with the dice is greater than 1/6.










# Week 11 Vid 2

## Hyp Tests (Continued)


CIs and Hyp Tests are our two big tools for "inference"

CIs are used for estimates.  What is it?

Hyp Tests are used for answering questions.  Is our support for our claim?










## Four key steps of hyp tests

### 1) Hypotheses

Two competing hyps:

H0:  the null hyp.  This is the assumption we make about the population.  It's ALWAYS a statement of equality about a pop parameter.

Ex:  H0:  p = .34  (note that p is the pop proportion)

It's needed to compute p-val!  We can't find probabilities unless we make an assumption about pop.  That's H0.

Ha: the alt hyp.  The alt hyp is our claim about the population.  It's what we think is "up".  It's the reason we're doing the test.  We hope to support it!  

The alt hyp is always a statement of INEQUALITY.  <  >  !=

Example:  Ha:  p < .34

The alt hyp is the logical opposite of the null hyp.  Only one can be true!
















### 2) Test statistic

In short: a summary of our data.  

Example:  phat

Almost always, expressed as a z score.  How about for **proportions**?

z = (phat - p)/sqrt(p(1-p)/n)

**Important note**:  All occurances of "p" here will use the value stated in the H0!  That's why the H0 is so important -- we can't compute our test statistic without first stating our assumptions!

Later, when we do hyp tests for means, we'll see a different test statistic that uses the center and spread for **xbar** instead.  But still, it's a z score!












### 3) p-val

The p-val is the probability of observing a sample that's as (or more) extreme than the one we got, assuming H0 is true.

In short, to find:  pnorm.  Find **tail area** past the test stat.  Exactly the same as "forwards" problems for phat and xbar.

You'll always need to consider: will my tail area be on the right or the left?  This is always easy to tell:  **look at the Ha!**

- If HA has "<", then your tail is on the left side (less than)

- If Ha has ">", then your tail is on the right side (greater than)

Remember, the z table always shows left-side probabilities.  So, this matters a lot!















### 4) Conclusion

If our p-val is small, then H0 is saying that our data is unlikely.  But of course we trust our data, not the hypothesis!

So, if p is small, we reject H0.  We have evidence to support Ha.

(Remember, "small" means "alpha", alpha = .05 by default).  Alpha = "significance level"

If p is large (p>alpha), then we FAIL TO REJECT H0.  We didn't find the evidence we're looking for to support Ha.  We have sad feelings, our paper doesn't get published, try again!

Remember, there are only two possible conclusions:

- If p<alpha, then we reject H0.  We found strong evidence to support Ha.  We're happy!  Our claim was validated!
- If p>alpha, then we fail to reject H0.  We didn't find enough evidence to support Ha.  We're kinda bummed out.  We didn't find evidence to support our hypothesis.

**Important note**:  Whaterver your conclusion is, **it must always contain two sentences**.  

1) Either reject or fail to reject H0.
2) Either find evidence to support Ha, or not enough evidence to support Ha.













## Example: 5.32 p205 (slight modification)

It is believed that nearsightedness affects about 8% of all children. In a random sample of 194 children, 21 are nearsighted. Conduct a hypothesis test for the following question: do these data provide evidence more than 8% of children are nearsighted?


Sample: phat = 21/194

```{r}
21/194
```


The sample result IS BIGGER than 8%.  Is this result extreme enough to convince us that the population proportion really is greater than 8%?  How likely is this sample? What human feeligns do we have?







1) **Hypotheses**

    H0:  p = .08
    
    Ha:  p > .08
    
    
    
    
    
    
    
    

2) **Test stat**

    phat = 21/194

```{r}
# find zscore.  note: center = p, se = sqrt(blaahblahs;as)


# idea:  (phat - p)/sqrt(p*(1-p)/n)

(21/194 - .08)/sqrt(.08*(1-.08)/194)


```

  (Human feelings:  Not an unusual score!)
  
  
  
  
  
  
  

3) **p-val**

Look up z=1.45 on z table:  left tail area is 

```{r}
pnorm(1.45)
```

Thus, p = 1-.9265 = **0.0735**

    Note:  this is a "right-tail" test.  You can tell from Ha.
















4) **Conclusion**

    Is p small?  Is p<alpha?  Is p < .05?

    No!  p is large!  We didn't find strong evidence against H0.  The sample results doesn't seem unlikely under H0.  We fail to reject H0!

    We didn't find the evidence we're looking for to suggest that more than 8% of children are nearsighted.  Nobody publishes our result.  We are sad and dejected.  
    
    
    
    