---
title: "Math 120 Week 12"
author: "Prof. Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: cerulean
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```












# Week 12 Vid 1

## Hyp Tests for mu (population mean)

Last week, we learned about hypothesis tests.  A hyp test is a formal procedure for asking a question about a population parameter.

All of the tests we did then were about categorical data.  In other words, the tests asked questions about population **proportions**.

This week, we'll explore hypothesis tests for quantitative data.  That means, we'll be asking questions about population **means**, mu.

Let's dive in with an example.













## Example 1

The height for adult men ages 20-29 in the US is 69".  However, a researcher suspects that male students at DU have a higher average height than the general population.

To investigate this, he takes a sample of 40 men at DU and finds their average height to be 70.2" with std dev 3.4".

Does this sample data provide evidence to support the researcher's suspicion?



Before we begin, a few observations:
- We're asking a question about a population:  is the mean height (mu) for male DU students higher than 69"?
- We have sample data, and we see that the sample mean is in fact larger than 69"
- However, we know that this is just a sample.  Does it provide strong evidence that the whole **population** mean is larger than 69?  That's what the hyp test is designed to show.
- The alternate hypothesis is always stated in the question.  Here:  Ha: mu > 69.  That means we know what the null hypothesis has to be also.


Let's perform the test:

1)  Hypotheses

H0: mu = 69
Ha: mu > 69

2) test stat = (obs - exp)/stdev

obs: xbar = 70.2
exp: mu = 69   [Note:  expectation always comes from H0]
stdev: s/sqrt(n) = 3.4/sqrt(40)   [Note:  only have sample std dev, s.  So, use t-table, not z table]

```{r}
(70.2 - 69)/(3.4/sqrt(40))
```

t = 2.23

3) p-val = "tail area"

Note:  use t-table for means.  (Use z-table for proportions).  

Here, df = n-1 = 40-1 = 39.  Df = 39.  

From table, we see .01 < p-value < .025

4)  Conclusion

Since alpha not stated, use alpha = .05.  Q:  is our p-val smaller than alpha?  A: yes!

Here:  p-val < .025, so p-val < alpha.  Ie, our p-val is small!

Thus we reject H0.  There's strong evidence that the true mean height of men at DU is greater than 69".  









## Left-tail / Right-tail / Two-tail tests

Remember the definition of p-value:

p-val = The probability of observing a sample result that's as (or more) extreme than the one we got, assuming H0 is true.  Right now, we're going to focus on one part:  "as (or more) extreme".

What makes a sample result "extreme"?  Well, that depends on what we're trying to show.  In other words, it depends on the **alternate hypothesis**, Ha.  Remember, there are only three options for Ha:  either > (greater than), < (less than), or != (not equal to)

If we think the parameter (mu/p/whatever) is **smaller**, then Ha has to have "<".  Thus, our notion of "extreme" here is **smaller than**.  This is called a "left-tail" test, because the tail area is literally on the left-side of the curve.

If we think the paramater (mu/p/whatever) is **larger**, then Ha has to have ">".  Thus, our notion of "extreme" here is **larger than**.  This is a "right-tail" test.

The third kind is new for us in this class:  not equal.  In this case, we don't necessarily think "bigger" or "smaller"... it could be either!  That means, our tail area comes from **both sides**!

There's a good picture here:  https://www.fromthegenesis.com/wp-content/uploads/2018/06/Types-of-Hypothesis-Tests.jpg









Since the normal curve (and t curve) are symmetric on both sides, it's easy to get this area:  just multiply by two.  (Think:  there are two tails, so p = 2*tail area.)

If you're looking at the t-table, both 1-tail and 2-tail areas are shown (althought that's kinda unneccesary, since you literally just multiply by 2 to find 2-tail area).

Remember, you can always tell what kind of test it is from the Ha:

- If Ha has "<", then "left-tail" (i.e., one-tail)
- If Ha has ">", then "right-tail" (i.e., one-tail)
- If Ha has "!=", then "two tail"













## Example 2

The mean IQ for the general population is 100 (the test is designed and tweaked to maintain this).  

Suppose a researcher suspects that professional musicians have a different mean IQ than the general population.  To investigate, she administers an IQ test to 55 randomly-selected professional musicians.  From this, a mean IQ of 103.2 with stdev 9.2 is calculated.  

Does this sample provide evidence to support her suspicion?


1) Hypotheses

- H0:  mu  = 100
- Ha:  mu != 100

2) Test Stat

t = (xbar - mu)/(s/sqrt(n))

```{r}
(103.2 - 100)/(9.2/sqrt(55))
```

t = 2.58


3) p-val (two tail)

Df = 55 - 1 =54

Note:  use df=50 since 54 isn't on the table.

From the t table, .01 < p < .02

```{r}
2*(1-pt(2.58, 54))
```



4) Conclusion

Since p < .05, we reject H0.  There's strong evidence that mean IQ of professional musicians differs from 100.














## Example 3

A biologist suspects that the sex ratio of a species of Tiger Snake differs from 50/50.  To test this, they examine 100 tiger snake embryos, and find that 59 of them are male.  Does this provide evidence to support the biologists claim?

1) Hypotheses


2) Test Stat


3) P-val


4) Conclusion











