---
title: "Math 120 Lec 13"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```



## New Special Distro: The **Poisson Distro**

It’d be rad if it was pronounced “Poison Distro”, but it’s not. It’s French: pwah-sawn


Just like the binomial, the poisson is a discrete distro.  I.e., all possible values of X are whole numbers.

Remember, if X is binomial, then X counts the number of successes over a fixed number of trials (n).

If, on the other hand, X is a poisson RV, then X counts the number of occurences over a fixed interval of time or space.

Ex:  Reading assignment 4.32.  Scenario:  A very skilled court senographer makes one typo per hour on average.

d) Calculate the probability that this stenographer makes at most 2 typos in a given hour.

Here:  X = the number of typos in one hour is poisson RV.  

All poisson variables must satisfy three properties:

- events happen at a constant average rate (lambda)
- all occurences are independent of eachother
- the chance of an event occuring is proportional to the width of the interval (time or space)

In 4.32:

- lambda = 1
- assume independence
- If the stenographer spends more time time typing, they'll probably make more typos.

If X satisfies these properties, then it has the following pdf:

f(x) = e^(-lambda)*lambda^x/x!

(note, you won't use this by hand).

In R, it's easy:  dpois()!  


Note:  dpois is pdf, and ppois is cdf (just like for binomial)

Let's anser 4.28d:  P(at most two typos).  

First:  what x vals?

Note: "at most" means "less than or equal to"

X=2 yes! X=1, X=0  Need all threee

```{r}
#first, let's use pdf three times:

# P(X=0) + P(X=1)    + P(X=2)
dpois(0,1)+dpois(1,1)+dpois(2,1)
  
  


```

There's a 91.97% chance that our stenographer makes at most 2 typos in an hour.  The end.   

Note:  could also use the cdf ppois()

```{r}
#  P(X<=2)
ppois(2,1)


```

Example 1)

At an archeological dig site of 25square meters, a total of 534 fragments were found at the site.  In a particular 1 square meter section, what's the probability that at least 20, no more than 22 fragments are found?

Here, we're thinking about #fragments / sq meter.  Occurences over space, not time!

Matt:  avg should be 534 fragments divided by 25 sq m

```{r}
#lambda

534/25



```


Put it all togther:  

```{r}
# X = 20, 21, 22,    lambda = 21.36

dpois(20, 21.36) + dpois(21, 21.36) + dpois(22, 21.36)

# try again, but with ppois.  should be same!

ppois(22,21.36) - ppois(19,21.36)

```

### Important differences between poisson and binomial

In poisson, there's no fixed number of trials.  In poisson, there's a unit of time or space  (think:  1 hour, 1sq meter), but the thing we count is the number of success.

Related idea:  "support".  The support for a RV is the set values for which it has nonzero probability.

I.e., all possible values of X.

If x is binomial, its support is X=0,1,2,3,...,n
If x is poisson, its support is X=0, 1, 2, 3, 4, ..., 300, ... infinity!

(It's at leasty hypothetically possible that our stenographer makes tons of errors, lol at stenographer.)

Related question:

What's the prob that the stenographer makes more than 3 typos?


P(X>3) = P(X=4)+P(X=5)+P(X=6)+...+P(X=23423423) + ...





















