---
title: "Math 120 Lab 03 - Exploring Discrete Probability Distributions"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: false
    toc_depth: 3
    code_folding: show
    theme: yeti
    df_print: paged
  
---

```{r setup, include=FALSE}
###########################################################################
################# DON'T DELETE THIS PART!!!!!##############################
###########################################################################
                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans"))                                 #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #
### Below is where you read and store your data.                          #
                                                                          #
youShouldRenameThis <- read_csv("death penalty data.csv")
                                                                          #
                                                                          #
                                                                          #
############# ^ ################################# ^ #######################
############# | ## DON'T DELETE THIS PART!!!!!### | #######################
############# | ################################# | #######################
```



## Introduction

In this lab, we'll be using discrete probability distributions to answer questions about real-world scenarios.  As we've seen in lecture, the distributions are easy to use in R.  But, deciding which distributions are appropriate is much more interesting.  


## Tools and Distros

The distributions we'll use for this lab are:

- The Binomial Distribution
- The Poisson Distribution
- The Geometric Distribution

Note that we haven't discussed the Geometric distribution in class.  One of the goals of this lab is for you to explore the geometric distribution on your own.  You can find everything you need in our book, Open Intro Statistics (4th ed) on page 145-146.  Also, there's a big wide internet out there full of answers; your professor is especially fond of Wikipedia.


This lab presents you with 5 scenarios, each with a question about the data.  For each, you'll need to adress the following:

- Which distribution (if any!) can be used to answer the question.
- A discussion of why the properties/requirements of the distribution are satisfied (for example, binomial RVs must satistfy four criteria described in Lec 13).
- A plot of the distribution.  For the distributions with infinite support (see Lec 14 for a definition of *support*), I leave it up to your good judgement where to stop the graph.  For example, below is a graph of a poisson distribution with lambda = 3.  I chose to stop it at X=12, even though the poisson has infinite support.

`qplot(1:12, dpois(1:12, 3), geom="col")`

```{r}
qplot(1:12, dpois(1:12, 3), geom="col", xlab = "x", ylab = "P(x)")
```

- The probability in question.  Here, you'll use the appropriate pdf or cdf.  Include a brief explanation of how you computed the probability and what it means.
- A discussion about whether the event in question is "unusual".  The expected value and variance should feature prominently in this discussion.  You may use, as a rule of thumb, that observations which are more than two standard deviations away from the expected value are "unusual".  The graph of the distribution could also be a good reference...
- If the scenario **doesn't match** any of the probability distributions we've learned (binomial, geometric, or poisson), then write a brief explanation of why not.  

## The Scenarios

In this section, your scenarios are described.  You should structure your report accorgingly.  For each scenario, write a paragraph that addresses the considerations listed above.  As always, use clean formatting, tight grammar, and crisp technical writing.  

1) According to the most recent figures from the Center for Disease Control (the CDC), 15.5% of adult Americans (age 18 and up) are regular smokers.  If we take a sample of 50 Americans, what's the probability that at least 1/2 but at most 1/4 of them are smokers?

2) According to the RunKeeper app on Prof Miller's smartphone, she'd gone running 58 times in 2019 as of Sept 30.  You can google to find out that this is a time span of 39 weeks.  Assuming that her running habits don't change for the rest of the year, what's the probability that she goes running at least twice this week?

3) What strangely warm, dry weather we've had this fall!  In Granville, the wet season spans April through August; during those months, the chance of rain is about 30% on any given day. During September through March, the chance of rain is about 19%.  If we took a random sample of 40 days throughout the year, what's the probability that it rains on at least 10 of them?  

4) At present, about 55% of Denison students are women.  If we sample students at random (say by choosing at random from the Registrar's records), what's the probability that we don't choose a female student until the 6th student that we sample?  


## Deadline and Grading

This lab is due **Monday, Sept 29th at 11:59pm**.   **I'll charge a 20% late fee per day late.**  

I will assign grades on this project ”holistically”, not using a scoring sheet where each item is worth a certain number of points.

- **B**: A ”B” project is very well done. It should have sound analysis, use appropriate statistical tools, support analysis with the appropriate graphs and other visual aids, and be a well-written document.

- **C**: A ”C” project is solid, but usually lacking in one of the areas listed above. 

- **D/F**: Projects below a C are severely deficient in the basic criteria or have failed to complete the assignment.

- **A**: An ”A” project is truly above ordinary; it has exceptional qualities that make it stand out distinctly. An ”A” project is not the default grade.
  - Does the project fulfill the requirements of the assignment?
  - Does this report have a ”thesis” or ”story” that is the focus? Does the rest of the report support the thesis?
  - Does the report use the subject matter from this unit appropriately? Are there tools being used in the wrong way? Is there an obvious tool that might have been used but was overlooked?
  - Are the graphics factually correct? Are they formatted well? Do they give the first visual impression that is supportive of the thesis/story of the report, or do they indicate a different story?
  - Is the report well-written? Is there good organization of ideas? Is the story/thesis communicated clearly? Is the analysis grounded in reasonable use of statistics? Is the paper free of mechanical and grammatical errors?
  - Does the report provide the right context for the target audience? Is the right information provided? Have the right assumptions been made about what the audience knows or does not know?
  
  
  
  
  
  
  
  
