---
title: "Math 120 Lec 10-11"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```

## Understanding Contingency (Two-Way) Tables

Why two-way tables?  They're great for humans!

The let us compare two categorical variables.

Ex: reading discussion data.

### Example Problems (Basics)

If a a photo is selected at random, what's the probability...

a) ... the photo was correctly identified as a fasion photo?

= 197/1822


b)  ... the photo is a fashion photo?

=309/1822

c)  ... computer thought it was a fashion photo?

219/1822


d) ... it's a fashion photo -or- the computer thought it was a fashion photo.


(197+22+112)/1822


```{r}
(197+22+112)/1822
```

For scienteists, mathematicians, computer scientist, and everyone else, "or" is "inclusive or".  I.e., could be one, could be the other, could be both.

## Making Contingency Tables in R

It's easy!  Just use the `table()` command!

Example:  mpg data.  Two way table for class and drive shaft (drv).

```{r}
mpg %>% head

table(mpg$class, mpg$drv)
```

Btw, it's also easy to get sums/total


```{r}
table(mpg$class, mpg$drv) %>% addmargins()
```



## Mutually Exclusive/Disjoint Events

The mean what you think they do:  it's impossible to be in both events!

I.e., the number of outcomes in both event A and event B is zero!

One more way:  The probability P(A and B) = 0.

Ex:  Are the events "is a pickup" and "has front wheel drive" mutually exclusive?

Answer:  since there are 0 pickups with front wheel drive, they **are** mutually exclusive/disjoint.

### The "or" rule for mutually exclusive events


If A, B are disjoint, then:

$$  P(A or B) = P(A) + P(B) - P(A and B)        $$

Note:  since A, B are disjoint, the "and" prob is zero!  Thus,

$$  P(A \cup B) = P(A)+P(B)   $$

Use this on the car example:  what's the probability that a randomly selected car is either a pickup -or- has front wheel drive?  We already know they're mutually exclusive.  Use the rule!

```{r}
table(mpg$class, mpg$drv) %>% addmargins()
```

P(Pickup or FWD) = P(Pickup)+P(FWD)-P(Pickup and FWD)
                 = 33/234   + 106/234 - 0/234

```{r}
33/234   + 106/234 - 0/234
```
                 
                 
Challenge:  draw venn for disjoint events:

(see board, two seperate circles.)

## Conditional Probability

Conditional:  we know that some other event has happened.

P(A | B) = "the probability that A happens, given that B has occurred"

The given condition comes after the bar.

P(Rains | Cloudy outside) = Probability of rain, given that we see it's cloudy outside.

For the formula P(A|B), we know B happened, care about A:


P(A|B) =  P(A and B) / P(B)

that's it!  "And" on the top, "Given" on the bottom.


## Independent Events




### The "and" rule for independent events



## Checking Independence

### Example:  mpg drv and class


### Example:  Death Penalty data

```{r}
deathData <- read_csv("death penalty data.csv")

deathData%>% head

table(deathData$Region, deathData$Method) %>% addmargins()



1175/1442*158/1442

152/1442
```


