---
title: "Math 120 Week 4"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

studentSurvey <- read_csv("Student Data.csv")
deathPenaltyData <- read_csv("death penalty data.csv")
housingData <- read_csv("../Data/House-Sales.csv")

#------------------------------------------------#
```


# Monday Sep 14

## Quiz 











## Effects of outliers on linear regression

Always need to *look* at the scatter plot to check for outliers.  Outliers can have a variety of effects.

- An outlier could make a weak relationship seem strong. (iris examle)
- Or, an outlier could make a strong relationship seem weak (mpg example)
- It might not have much effect at all.

Morals of the story:

- outliers can have all sorts of effects (stronger, weaker, mnot much chang)
- always need to look at scatterplot
- perform analysis both ways: with and without outliers.












## What's an outlier -- IQR Method

Intuition: outliers are "unusual" values in a dataset.

Need an objective tool to decide what values are (and are not) outliers.


IQR Method:  All numbers that are either

- smaller than Q1 - 1.5*IQR or
- bigger than Q3 + 1.5*IQR

are outliers.

**Example:  Exam scores:**

75, 76, 79, 80, 80, 81, 82, 83, 84, 99

We see that Q1 = 79, Q3 = 83, IQR = 4

Compute cutoffs:

Low:
```{r}
79 - 1.5*4
```
Upper:
```{r}
83+1.5*4
```

Since 99 is larger, it's an outlier!

Note, this is an arbitrary (but frequently used) standard.  

Some texts define an "extreme" outlier as one that's 3 IQRS away from Q1, Q2 (as opposed to 1.5).










## What's an outlier - z score method

For a value, the z score ("standard score") is the number of std devs above or below the mean that the value is.

Example:  average height is 64", std dev is 3".  Jill is 70" tall.  What's her zscore?

Her z-score is +2!  She's 2 std devs above the mean.

z = (obs - exp)/stdev
  = (x - xbar)/s_x

Z-scores are a measure of how "unusual" a value is.  Large positive z score means way above the mean!  Large negative z-score means far below the mean.


Sometimes, outliers are defined to be any value with a z-score greater than 3 or less than -3.

We'll always prefer the IQR method, since IQR is robust and stdev isn't.  










