---
title: "Math 120 Lab 5 Example Code"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: yeti
    df_print: paged
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```




Take samples of transmission type from each category of "cyl" in the mpg data, and compute CIs for the proportion of cars with "auto(l4)" transmissions:

```{r}

# Filter the data

mpg %>% filter(cyl == 4) -> fourCyl
mpg %>% filter(cyl == 6) -> sixCyl
mpg %>% filter(cyl == 8) -> eightCyl

# Take samples

fourSample <- sample(fourCyl$trans, 30)
sixSample <- sample(sixCyl$trans, 30)
eightSample <- sample(eightCyl$trans, 30)

# Compute proportions for "auto(l4)" transmission

phat_4cyl <- sum(fourSample == "auto(l4)")/30
phat_6cyl <- sum(sixSample == "auto(l4)")/30
phat_8cyl <- sum(eightSample == "auto(l4)")/30



```

Find true propotion for proportion of cars with "auto(l4)" transmission:

```{r}
# Use all data, not the samples

prop_autol4_4cyl <- sum(fourCyl$trans == "auto(l4)")/nrow(fourCyl)
prop_autol4_4cyl

# similarly for 6cyl and 8 cyl (not shown here)
```


Compute 95% CIs 

```{r}

phat_4cyl - 1.96*sqrt(phat_4cyl*(1-phat_4cyl)/30)
phat_4cyl + 1.96*sqrt(phat_4cyl*(1-phat_4cyl)/30)

#similarly for 6cyl and 8cyl (not shown here)

```

Compute 99% CIs

```{r}
phat_4cyl - 2.576*sqrt(phat_4cyl*(1-phat_4cyl)/30)
phat_4cyl + 2.576*sqrt(phat_4cyl*(1-phat_4cyl)/30)
```

Present resulsts in a table:

| Cylinder Type | True proportion for "auto(l4)" |     95% CI      |   99% CI       |
| ------------  |  ---------------------------   |  -------------  | -------------  |
|   4           |  29.6%                         |  (19.4%, 53.9%) | (14.0%, 59.3%) | 
|   6           |           ...                  |     ...         |     ...        |
|   8           |           ...                  |     ...         |     ...        |
















How to find proportions in R?

First, take a sample.  Use iris, sample species.

```{r}
irisSample <- sample(iris$Species, 30)
irisSample

```

Q: how to find the proportion of virginicas?

A: make R count all the viginicas

```{r}
sum(irisSample == "virginica")
```


Neat trick for making tables:

|Var 1  | Var 2 |  Var 3 |
|------ | -----|  ----- |
| a     |  1     | blue  |
| b     | 2     |  green |







