---
title: "Math 120 Week 7"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

studentSurvey <- read_csv("Student Data.csv")
deathPenaltyData <- read_csv("death penalty data.csv")
housingData <- read_csv("../Data/House-Sales.csv")
#------------------------------------------------#
```



# Monday Sept 28

## Complement of an Event

Complement means "not"

Notation:  If our event is called "A", then the complement is "A^C"

$$A^C$$

Sometimes: A'

Ex:  Suppose we roll a dice.  A is the event that a prime number shows up.  Compute/express A^C:

Here:  A = {2, 3, 5}

So, A^C = {1, 4, 6}

Complements:

Since, for any event A, our entire set of all possible outcomes is everything in A OR in A^C

SO:

    P(A)+P(A^C) = 1
    
More frequently expressed as:

    P(A^C) = 1-P(A)
    
**Ex** What's the probability that we roll a dice and DON'T get a "2"?

Could do:  1/6+1/6+1/6+1/6+1/6 = 5/6

Easier to use the complement rule:  1 - 1/6 = 5/6







## Independence and the Multiplication Rule

Recall, two events A,B are "disjoint" if it's impossible for both A and B to occur at the same time.

    P(A and B) = 0
    
Review:  If A and B are disjoint, then P(A or B) = P(A)+P(B)


Ex:  If we roll a a dice, what's the probability that it's a "3" or an "even"?

These are disjoint, so P(3 or E) = 1/6 + 3/6 = 4/6

Now:

Two events A, B are "independent" if the outcome of A doesn't affect the outcome of B, and vice verse.

WARNING:  this is completely unrelated to "disjoint"/"mutually exlusive".  In fact it's impossible for two events to be both disjoint and independent. (Explanation below)

Example:  Roll two dice.  The outcome of one dice doesn't affect the outcome of another.

Example:  coin toss.  Doesn't matter if you get 10 heads in a row, the probability for the next toss is still 50/50.  

To check for independence (or to use it):  the multiplication rule for independent events:

If A, B are independent, then P(A and B) = P(A) * P(B)

This works for more than two events.  If A,B,C,D are indep:

    P(A and B and C and D) = P(A)*P(B)*P(C)*P(D)
    
If indep, "and means times"

**Example**  Suppose we roll two dice.  What's the probability that we get a "2" both times?

    P("2" and "2") = 1/6 * 1/6 = 1/36
    
```{r}
1/36
```

**Example** Suppose we roll two dice. What's the prob that the first is a "2" and the second is a "5"?

    P(2) = 1/6,  P(5)=1/6
    
Since indep:

    P(2 and 5) = 1/6*1/6 = 1/36
    
Same thing!

**Example** Suppose we roll four dice. What's the prob that:
 - the first is a "2"
 - the second is an "even"
 - the third is strictly greater than 4
 - the fourth is a prime number?
 
 
     P(2 and E and 4 and P) = 1/6 * 3/6 * 2/6 * 3/6 = 18/6^4

```{r}
18/6^4
```

**Super important example**:  Suppose we roll a dice 10 times.  What's the probability that at least one outcome is a "2".

WARNING:  this is describing many, many possible outcomes!  Tens of thousands of possibilities!

2 x x x x x x x x x 

x 2 x 2 x x x x x x 

x x 2 2 2 x 2 x x x

How to solve?  Complement!

Our event:  At least one "2".  

Complement:  NONE OF THEM ARE 2s!!!

There's only one type of solution:

x x x x x x x x x x x 

For each:  P(not a 2) = 1 - P(2) = 5/6

(5/6)(5/6)(5/6)*....*(5/6)  = (5/6)^10

P(at least one "2") = 1 - (5/6)^10

```{r}
1-(5/6)^10
```



**Example** If we toss a coin 20 times, what's the probability that at least one is a heads?

For "at least one", use the complement!

Here, the complement is:  only get tails!

P(T) * P(T) * ..... * P(T) = (P(T))^20 = (1/2)^20

So, P(a least one heads) = 1 - (1/2)^20

```{r}
1 - (1/2)^20
```


Examples p90





## Contingency Tables - Marginal and Joint Probabilities

A "contingency table" (also called a "two-way table") is a summary of counts/frequencies for two categorical variables (one in the rows, one the columns.)

Let's go make one:  Student Data in Google Sheets

In  a pivot table, there are two kinds of probabilities we can find

[remember, these are all EXPERIMENTAL/RELATIVE FREQUENCY probabilities because they're based on data]

- "joint" probabilities.  In other words:  "and" probabilities.

**Ex**  Find the probability that a student is a junior AND has blonde hair.

SInce there are 5 students who are both blonde and junior, and since there's 42 total students, we get a joint probability of 

```{r}
5/42
```

Note that in our student contingency table, there are 12 differernt joint probabilities (4 years, 3 colors  4*3=12)


- Next type of prob from contingency table:  **marginal** prob

**Ex** What's the probability that a student is a Soph?

Since there are 5 soph (check the margin) and 42 total, there's a 

```{r}
5/42
```

**Ex** What's the probability that a student is a senior?

There are 19 seniors, 42:

```{r}
19/42
```


### Finding "or" probabilities from contingency table

Recall:

    P(A or B) = P(A) + P(B) - P(A and B)
    
In table, we can find these:

- P(A) and P(B) are both MARGINAL probs!
- P(A and B) is a JOINT prob!

**Ex**  What's the probability that a stuent is a senior OR has brown hair?

    P(Sen or Brn)  = P(S) + P(B) - P(S and B)
    
From table:

- P(S) = 19/42
- P(B) = 28/42
- P(S and B) = 15/42

All together:

    P(S or B) = 19/42+28/42-15/42
    
```{r}
19/42+28/42-15/42
```
  
**Ex**  What's the prob that a student is Blonde or Fresh?

1:  notice it's an "or"
2:  remember, have a "general or formula" that always works

    P(Bl or Fr) = P(Bl) + P(Fr) - P(Bl and Fr)
                   6/42 + 6/42  - 0/42
                   
Btw, this means the events are "disjoint" or "mutually excl" (careful, not the same as "independent").

```{r}
12/42
```

###  Checking for independence in contingency table

WARNING:  future quiz question

**Ex** Are the events "student is a senior" and "student is blonde" independent?


Remember, if events indepenent then:

    P(A and B) = P(A)*P(B)
    
Note: this is how we always check for idependence!

If both sides are equal, then YES indep.  Otherwise, no, not indep.

Here:


    P(Sr and Bl) = P(Sr)*P(Bl)
    
- P(Sr and Bl) = 1/42
- P(Sr) = 19/42
- P(Bl) = 6/42

Let's check!  Left side:

```{r}
1/42
```

Right side:

```{r}
(19/42)*(6/42)
```

Is it true that .0238 =? .0646  No!  Thus, they're not independent!

**Ex**  Are the events "brown hair" and "junior" independent?


## Conditional Probability

Sometimes, our knowledge of one event changes how likely another event is.


Ex:  If it's cloudy outside, it's more likely to rain.

Ex:  If the skies are clear, rain is unlikely.

Notation:

    P(A|B)
    
Read as:  "probability of event A, GIVEN that event B has occurred.

The thing after ther | is the "given condition".  Ie, we know that thing has occurred.

Goal:  find the probability.

From picture, we have formula:

    P(A|B) = P(A and B)/P(B)
    
Note:  the letters don't matter, but the given ALWAYS has to be the bottom!

     P(M|G) = P(M and G)/P(G)
     
Ex:  Suppose the probability of being both cloudy and raining is 30%.

If the probability it's cloudy is 40%.

If it's cloudy, what's the probability that it rains?

    P(R|C) = P(R and C)/P(C) = .3/.4 = 75%
    
**Ex**  Contingency Table

Studend data for hair color and year.

What's the probability that a student has brown hair, given that the student is a soph?

If a student is a soph, what's the prob that they have brown hair?

    P(Br | S) = P(Br and S)/P(S) = (4/42) / (5/42) = 4/5 = 80%
    
**Ex**  We randomly select a card from a 52-card deck.  What's the probability that the card is a face card, given that it's black.

    P( F | B)  = P(F and B)/P(B)
    
Need to count:

- How many are black and face?  6/52
- How many are black?  26/52

Finish:

    P(F | B) = 6/52 / 26/52 = 6/26
    
```{r}
6/26
```

  
So far:

 - general "or" formula
 - "not" probabilities (complement) [related: at least one]
 - "and", but ONLY for independent!!
 - contidional probs
 
 
## General "And" formula

Recall:      P(A|B) = P(A and B)/P(B)

Note that this always works for all A,B, so we can rearrange to get the "and" formula:

For any two events, A,B:

    P(A and B) = P(A|B)*P(B)
               = P(B|A)*P(A)
               
Both versions always give the same number, but often only have info for one!

**Ex**  Suppose 51% of DU students are female (nearly true).  Of all female students, 45% are STEM majors (making this one up).

What's the probability that a randomly selected student is both female AND a stem major?

Translate:

- P(F) = .51
- P(STEM | F) = .45

Need to find:  P(F and STEM)

Two versions:

    P(F and Stem) = P(F | Stem) * P(Stem)
                  = P(Stem | F) * P(F)   <---- have to use this one!
                  
                  = .45 * .51
                  
```{r}
.45*.51
```

About 23% of all DU students are female STEM majors.

d
                  
        
    
    



















  




















## Probability Trees

Probability trees provide a handy visual aid for organizing all of the various probabilities we care about.

*Ex*  Suppose a test for a disease advertises itself as "95% accurrate.  This means:  if the test is administered to someone who *has* the disease, the test will show "+" 95% of the time.

Suppose that if a person doesn't have the disease, then they'll test "-" about 85%.

Suppose that 3% of the population has the disease.

Let's draw a probability tree to summarize the probabilities.

List of Probabilities:

P(+|D) = .95
P(-|D) = .05
P(+|D^C) = .15
P(-|D^C) = .85


P(+ and D) = .03*.95

```{r}
.03*.95
```


P(D and -) = .03*.05

```{r}
.03*.05
```

```{r}
.97*.15
```

```{r}
.97*.85
```

In summary, the tree gives and intuitive layout for all conditional probabilities and "and" probabilites.


Two important uses:

1)  Q:  what percent of people test positive?  Can easily figure it out from the tree.  There's only four possibilities!

  Want to find:  P(+)?  There are only two options for this scenario:  D and + or D^D and +.  So, add up the probabilities!
  
P(+) = .0285 + .1455

```{r}
.0285+.1455
```

About 17.4% of all people in the population will test positive.

2)  What's the probability that, if you test +, you actually have the disease??  Observations:

- this is what YOU care about if you're taking the test!
- "backwards" from what we have right now!  RIght now, we know P(+|D).  We want the other way:  P( D | +)?  

From the tree, we know everyting.  Use the conditional probability formula:


    P(D|+) =  P(D and +)/P(+) = .0285/.174
    
```{r}
.0285/.174
```

If you test positive, there's only about 16.4% chance that you really do have the disease!  Woah!  
    
    
Notes:

- 1 above is the "law of total probability".  To find P(+), we added all the possible probabilities:


    P(+) = P(D and +) + P(D^C and +)
    
- 2 above is Bayes Rule!  Bayes rule is a formula/law for "reversing" conditional probabilities.  Here: we started with P(+|D). But actually, we're interested in the probability P(D|+).

Recall, the test advertises itself as "95% accurate".  BUT, if you get a + result on the test, there's only a 16.4% chance that you really have the disease!!

Think of Bayes this way:  it reverses conditional probability using info from the tree and the definition of conditional probability.


**Example:  Bayes**  Suppose: 

- 0.1% of the American population currently has lung cancer (event "L") 
- 90% of all lung cancer cases are smokers (event "S") 
- 21% of those without lung cancer also smoke. 

(These values are fairly close to the values given on the American Lung Association web site

Construct a probability tree to find the following probabilities:

- P(S^C|L) = .1

- P(S and L^C) = .20979


- P(S) = .0009 + .20979 = 0.21069

```{r}
.0009 + .20979
```



- P(L|S) = P(L and S)/P(S) = .0009/.21069

```{r}
.0009/.21069
```




- P(L|S^C) = P(L and S^C)/P(S^C) = .0001/(.0001+.78921)

```{r}
.0001/(.0001+.78921)
```

Q:  how much more likely is it for a smoker to have lung cancer than a non-smoker?

Divide:  smoking lung rate/non smoking rate:

```{r}
0.004271679/0.0001266929
```

Smokers are nearly 34 times more likely to have lung cancer!


```{r}
hist(runif(10000))
```


# Friday 10-9

## Continuous Distributions - Normal Dist

Context:  let's get to know the most famous probability distribution: normal dist, ie, the bell curve.

Interesting fact:  the normal dist arises naturally in many real life/data applications.  

Cool fact:  the normal dist is seen whenever we examine sums or **means** of random numbers.

```{r}
runif(1)
```

Random real number between 0 and 1.


```{r}
runif(10)
```


```{r}
hist(runif(10000))
```

Punchline:  what if we do this a bunch of times (10000) and look at a histogram of the means/averages:

```{r}
hist(replicate(50000, sum(runif(10000))))
```

Moral of the story:  any time we study averages of randomly selected number (each one from the same population), the means (or sums) will follow a normal dist.

Intuition:  most values are in the middle.  Very large (or very small) results are increasingly unlikely.

The normal dist also arises frequently in biological variables: ex human height is normally distributed.  


Weird thing about the normal dist:  it's a "continuous" distribution.  In other words, we can't list out all the possible options for X.

Ex: Human height.  Not possible to make list of all possible human heights.  Infinitely many!

Because of this, we can't make a distribution table!

Instead:  use a density curve.  To find probability for a normal dist, we need to find area under the density curve.

Ex: Height for adult men in the US follows a normal dist, with mean 69" and std 2.7".

What's the probability that a randomly selected man is less than 70" tall?

What's the probability that a randomly selecte man is between 67 and 68" tall?

**Problem** It's easy to draw the picture, it's hard to find the area!  No simple geometric formula!

If you've taken calculus:  this is an unsolvable integral!  No closed-form solution!

So, we NEED computational assistance.  

1)  Refer to a table
2)  Use software (including Google Sheets)

**One more problem**  There are MANY normal distributions (depending on center and spread).  

Answer:  use the Standard Normal Dist, ie the normal distribution with center = 0 and spread (std dev) = 1.

To perform any of these computations, we'll need to convert our data to a standard normal.  (More on this later).

Let's do some example computations.

NOTE:  The standard normal distribution is ALWAYS denoted by "Z".  

*EX* Suppose Z has a standard normal dist.  Compute the following:

a) P(Z<1.28) = .8997

b) P(Z>1.28) = 1 - P(Z<1.28)   [complement rule]

```{r}
1-.8997
```

c) P( -0.82  < Z < 2.51 ) = P(Z<2.51) - P(Z< -0.82) = .9940 - .2061

```{r}
.9940 - .2061
```



Problem:  if we look up z=2.51 in the table, we'll get ALL the area below that cutoff.  But that's too much!  So, let's subtract off the area below -.82:




















