---
title: "Math 120 Lec xx"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```




# 09-01 Wednesday

## First "Real" Day


Hi!  This is how I take notes in class!

I can do math right here!

```{r}
2+2
```

## Statistics -- WHat's that?

We've got data!  Ie, measurements and observations about individuals (people/animals/objects/things).  Based on data, want two things:

1) Descriptive Statistics:  describe the individuals!  Summarize, visualize, etc.

2) Inferential statistics:  say smart things about the **population** that the people/objects/things come from based on the data!

Example:  student data.  Let's data about *you*, and say smart things about you!

Note, our data is "rectangular data"/"tabular data":

Three important things:

1) Each column (up/down) is a **variable**.  Ie, a specific quality of cases/individuals.
2) Each row (left/right) is a **case**/**individual**.  Ie, the specific people/things/objects we're collecting data from.
3) Each cell (each rectangle) has exactly **one** measurement.

ALL DATA we work with in Math 120 is rectangular!

Now:  let's do some descriptive stats!  What does this data say about YOU ALL?

Q:  What sorts of summaries might we care about?

Ex:  The MAX number of siblings we have is 5.  

Ex:  What about average siblings?  

Q:  What are GOOD summaries to use?

Example:  mean hometown population is 6+ million.  BUT, only 4 of you come from a hometown of more than 1 million.  

SEEMS LIKE AVERAGE/MEAN IS DOING A BAD JOB!


## Numerical Summaries (Quantitative Data)

- Mean/Average.  STORY OF THE FORMULA:  The mean is "balancing point" of the data.

- Median.  STORY OF THE FORMULA:  Middle element (when data is ordered)

- Mode.  The most common.

NOT ALL OF THESE ARE EQUALLY GOOD (ALL THE TIME)!

Idea: we want a good **Measure of center**, ie, what's typical for this data?

Ex: Hometown:

  - mean = 6.3 million
  - median = 0.125 million
  
Seems like median is doing a better job!  Seems more representative of "typical"!

Why is mean worse here?  SHAPE!  

Notes about our first histo:

- Bad!  Almost all data is in one bar, the last is an outlier.  Can't see the shape!  Fix:  bins!

Population shape:  strong right skew!  (skew == "tail")













# 09-01 Wednesday

## The real first day

This is how I take notes!

I like it because I can do math:

```{r}
2+2
```

## Statistics, What the heck is it?


Want to say smart things about data, ie measurments and observations about people/things/objects/whatever.

Two types of statistics:

1)  Descriptive statistics:  descirbing/summarizing/visualizing the people/objects/whatever in the dataset.

2)  Inferential statistics:  making empirically supported conclusions about the population that the people/objects/things come from, based on the data.

Let's look at some data:  you!

Note:  this data is **rectangular data**.  Three key qualities:

1) Columns (up/down) are **variables**, ie qualities of individuals
2) Rows (left/right) are **individuals/cases**.  The people/objects that we're measuring/observing.  
3) Each cell (rectangle) has exactly ONE measurement/observation.

WE ALWAYS USE RECTANGULAR DATA IN 120!


Types of data:

- Quantitative
- Qualitative

How we describe data depends on it's type!

What can we say about you?

Q:  What's the mean hometown population?

A:  1.97 million

Woah!  Doesn't seem like a good measure of "typical"!  Most students from hometowns less than 1 mil!

What's the right tool for the job?

## How to describe quantitative variables

Three key observations:

- measure of center.  what's typical?
- measure of spread.  In general, how far away from the center are people?  Are all close to center, or are many far away?
- Shape.  In what regions do most data lie?

## Measures of Center (Quant variable)

What's typical?

- Mean: STORY OF THE FORMULA:  The balancing point of the distribution.
- Median.  STORY OF THE FORMULA:  Middle number (if in order)
- Mode.  Story:  Most common element.

For population, mean and median are totally different!
     
     mean = 1 something mil
     median = .043 million
     
Median seems much more representative of the "typical" value for this data!
















# 09-01 Wednesday

## First "real" class day

This is how I take notes!

I can do math!

```{r}
2+2
```


## Stats - what the heck is it?

We've got data (observations about people/things), wanna say smart things about it!

Two kinds of stats:

1) Descriptive stats:  Describe data with summaries, visuals, etc.  What can we say about the individuals in the dataset?

2)  Inferrential stats:  Making empirically supported conclusions about the population the individuals came from.  

Example:  your data!

**Rectangular Data** has 3 properties:

- Each **column** (up/down) is a **variable**, a characteristic of an individual
- Each **row** (left/right) is an **individual/case**, ie the people/things that we collect data from
- Each **cell** contains exactly one measurement.

WE ALWAYS USE RECT DATA IN 120!!


## Types of Data

- Numerical/quantitative (number)
- Categorical/qualitative (not a number)

The tool we use depends on the type of data!

Ex: Hometown population. 

- What's the average?  A:  1.4 million

Woah!  Most students are from a hometown of less than 1 mil.  1.4 mil seems unrepresentative!

## What to say about quantitative variables

Three key features:

- Measure of center.  Ie, what's typical?
- Measure of spread.  In general, how far from the center are you?  How much are values spread apart?
- Shape.  In what ranges do data lie?

## Measures of center

- mean.  STORY OF THE FORMULA:  "balancing point"
- median.  STORY OF THE FORMULA: middle element (when in order)
- mode.  STORY:  most common

Problem:  we've seen that these can be drastically different!

    Pop:  mean = 1.4mil, median = 0.047 mil
    
Answer:  shape!  Since there are outliers, the mean is strongly affected (balane), but the median is not (middle).  

How to decide the measure of center:

- If the data has outliers/skew, use median (not affected).
- If the data is mostly symmetric, use mean (simial to median).



# Fri Sept 03

## Saying smart thigns about data

Last time, two types:

- quant (numerical)
- qual  (categorical)

## Smart things to say about quant variables

Three important characteristics/qualities for quant variables:

- measure of center.  what's typical?  
- measure of spread.  how far from center is typical?  how spread out are observations?  
- shape


## Measures of center

- mean (balance point)
- median (middle)
- mode (most common)

## Measures of spread

- standard deviation (stdev, sd)

Formula:

     s = sqrt( sum(xi-xbar)^2 / n-1)
     
Ex:  compute the stdev of:  1, 2, 3

xbar = (1+2+3)/3 = 2

     salmost =   [ (1-2)^2 + (2-2)^2 + (3-2)^2 ] / (3-1)
             =   [1 + 0 + 1]/2
             =   2/2 = 1
             
     ->  s = sqrt(1) = 1
     
STORY OF THE FORMULA:  Stdev is the "average distance" between data points and the mean.  

Q:  Why squared?  To prevent + and - differences from cancelling.  

FOR STUDENTS:  DON'T need to compute by hand.  DO need to understand story of formula.  

## Measures of spread (ctd)

- IQR (Interquartile range).  Quartiles:  25% marks.  (just like median is the 50% mark).

To find:  break data in half, find median of each.

Ex:  Find quartiles of  1, 2, 3, 4, 5, 20

lo half:  1,2,3  -> Q1 = 2

upper half:  4,5,20  ->  Q3 = 5

Then,

      IQR = Q3 - Q1
      
STORY OF THE FORMULA:  IQR is the range of the middle half of the data.


## Shape of a quant variable/distribution

Ie, shape of historgram.  Common shapes:

- symmetrical:  mirror image about the center. 
- skew.  THINK:  SKEW MEANS TAIL!!
- bimodal/multimodal

A measurement is is ROBUST if it's NOT strongly affected by skew/outliers.


Q:  Which is robust, the mean or median?

A:  The median is robust!  Middle element not affected by extremes!  Ex:  hometown population!   Mean not robust, strongly affected by tail.

Exercise:  For the following distributions, describe shape.  Which measure of center should we use?  Which will be larger?

a)  Data:  the heights of men at DU

b)  Data:  scores in a large class on an easy exam.


A:

a)  Expect symmetric.  Most in the middle.  Use mean (probably close to median).  Mean and median about the same.

b)  Skew left.  Use median to measure center (robust).  










































