---
title: "Math 120 Lec 25"
author: "Ali Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```







## Hyp Tests

If DU faculty want to change the faculty handbook, they need a "supermajority", 2/3 vote or more.

Prof Miller wants to gauge support for her proposal. She takes a sample of 50 faculty members, and finds 38 of them support her.

```{r}
38/50
```

Of course, that doesn't mean that 76% of all faculty support her.  Is there statistically significant evidence that a supermajority (>2/3) support her?  Conduct a hyp test.

### 1 Hypotheses

H0:   p=2/3  (we'll use this for p-val)
Ha:   p>2/3  (right tail test)

### 2 test stat

phat = 38/50

```{r}

(38/50-2/3)/sqrt(2/3*(1-2/3)/50)


phat <- 38/50
p <- 2/3
n <- 50


(phat-p)/sqrt(p*(1-p)/n)


```

### 3 p-val

```{r}
1-pnorm(1.4)
```

### 4 conclusion

Since p>.05, we fail to reject H0.

We didn't find strong evidence that the proportion who support Prof Miller is greater than 2/3.

Follow-up question:  based on the sample result phat = 38/50, construct a 95% CI for p.

```{r}
phat <- 38/50
n <- 50
zstar <- 1.96

#lower bound

phat - zstar*sqrt(phat*(1-phat)/n)
phat + zstar*sqrt(phat*(1-phat)/n)



```

We estimate the proportion of faculty who support Prof Miller between (64.2%, 87.8%).

Based upon this, is there evidence against H0?  Is there evidence to support Ha?

H0:  p = .667,   Ha:  p > .667

Based upon the CI, the H0 seems totally reasonable!  Certainly, we'd fail to reject, same conclusion from our p-val.  

Hyp Tests and CIs are connected in this way:  if the H0 value is contained in the CI, we fail to reject the H0.  Note, that alpha and confidence level must match!!  C=95%, alpha = .05

(Btw, that means our default conf level and default significance level are actually the same!)  As long as they match, CIs and p-vals will -always- give the same result!  You can conduct every hyp test with a CI!!