---
title: "Math 120 Lec 21"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```





## Confidence Intervals for p

Formula:



$$ \hat{p} \pm z^* \sqrt{\hat{p}(1-\hat{p})/n} $$

Idea:  we want to estimate the population proportion p, based on our sample proportion phat.

Another phrasing:  we're estimating parameters using statistics.

Note: when a statistician says "estimate", she always means "construct a confidence interval".

### Critical values

zstar is the connection to the normal dist -- what's the middle C% of the distro?

All confidence intervals have the same form:

(point est) +/- (crit val)*(SE)

The point est and the SE, you know right away from the sample.  
The real work:  finding crit val.  Easy: use `qnorm`

Example:  find crit val for 95%:

Know: the lower tail has 2.5%.  qnorm!

```{r}
qnorm(.025)
```

Exercise:  what's the crit val for 90%?

```{r}
qnorm(.05)
```

Typical rounding: zstar = 1.645

Exercise:  what's the crit val for 99%

```{r}
qnorm(.005)
```

Typical rounding:  zstar = 2.576

Note that 90%, 95%, and 99% are very common confidence levels.  So, you'll see 1.645, 1.96, and 2.576 a lot!

### Examples constructing CIs

Example:  In a sample of 240 giraffes, 86 of them enjoy water skiing.

Based upon this sample, estimate the true proportion of giraffes who enjoy water skiing with 95% confidence.

First, let's make sure we're justified in using normal dist!

Here, easy to see that 

n*phat >10 

and

n*(1-phat)>10

Now, since p hat is normal, find zstar!

```{r}
qnorm(.025)
```

zstar = 1.96 (as always for 95%).  Now, plug and chug!

```{r}
# phat +/- zstar*sqrt(phat(1-phat)/n)

86/240 - 1.96*sqrt(86/240*(1-86/240)/240)
```

```{r}
86/240 + 1.96*sqrt(86/240*(1-86/240)/240)
```

So, we can be 95% confident that between 29.8% and 41.9% of all giraffes enjoy water skiing.  

Important note:  the sentence is formulaic, conclusion should always be about POPULATION.

Exercise:  In a sample of 300 DU students, 94 of them play videogames regularly (at least once a week).  Estimate the proportion of DU students who play videogames regularly with 99% confidence.


```{r}
#phat: 94/300
#n : 300
#zstar:2.576

94/300 - 2.576*sqrt(94/300*(1-94/300)/300)
```

```{r}
94/300 + 2.576*sqrt(94/300*(1-94/300)/300)

```

We are 99% confident that the between 24.4% and 38.2% of all DU students play videogames at least once a week.

### Margin of error

MOE: it's the stuff after the +/-  

It determines the width of the CI

Of course, we want MOE to be small!

Two important considerations:

- Sample size:  if n increases, then MOE decreases.  Yay!  This is what we'd hope for!  It means, we have some control over error. Math: n is in the denominator for MOE.

- Confidence level.  Bigger confidence level means bigger margin of error.  Bummer!!  More confidence comes at a price: more error.  More confidence means bigger zstar means bigger margin of error.  

Making a CI is like trying to capture a little creature (the paramater) with a net.  My chances of capturing the creature are higher if I have a really big net, rather than small!

Students always ask:  why not just use a really, really confidence level?  Confidence comes at a cost: our intervals get really, really big.  Dunno the real value!