---
title: "Math 120 Lec 20"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```

## Practice Problem

About 82% of DU students are from out-of-state.  If we take a sample of 100 students, what's the probability that at least 90% of them are from out-of-state?

A few notes:  population is all DU students, we're taking a random sample of 100 students.  We're interested in the proportion of students who are from out-of-state.  

Here:  p = .82
Considering:  phat>.9

What distro does phat have?  Normal!  How do you know?

Check the "success-fail" condition:  need both 10 success and 10 failures.  I.e., np>10 -and- n(1-p)>10.

Here:  n*p=100*.9 = 90
Also:  n(1-p) = 100*(1-.9)=10

Which to use, p or phat?  If you have p, it's best to use it.  Generally, you never will!  In that case, use phat.  

Generally, they both give same results.  here with p:

100*.82 = 82
100 * (1-.82) = 18

Still good, all at least 10.

This means that n is "large" enough for the CLT to kick in:  we know phat has a normal dist.

Ok, we know we're using pnorm.  What are the center and spread for the normal?

mu_phat = p = .82
se(phat) = sqrt(p*(1-p)/n) = sqrt(.82*(1-.82)/100)

```{r}
sqrt(.82*(1-.82)/100)
```

Use pnorm to find P(phat>.9)

```{r}
1-pnorm(.9, .82, .03841875)
```

There's about a 1.8% chance of obtaining a sample result like this.  Sounds pretty unlikely!  ...

## Confidence Intervals

(Note: I'm following the book closely.  This is all 5.2)

Remember our big goal:  statistical inference is the process of making conclusions about the population based upon the sample result we obtained.  

So far, our conversation about sampling dists has been backwards:  how in the world would we know what the population proportion is?????  The whole reason we conduct samples is because we -don't- know what's going on with the pop, and we want to find out.

Today's technique:  Confidence Intervals, or CIs.

The idea is simple: the result from out sample is just a starting point, "point estimate".  For categorical data, phat is our point estimate.  We know its distribution:

mu = p
se = sqrt(p(1-p)/n)
shape is normal

Picture (on board): Confidence intervals show the middle C% (i.e. 95%, 99%, etc) of the normal distribution for our point estimate.  C = Confidence Level

Big question:  how do we find the low and high bounds for this?  

Answer:  We use z-scores backwards!

Algebra:

low:  phat - zstar * SE(phat)
high: phat + zstar * SE(phat)

Where zstar ("critical value") marks off the middle c% of the standard normal dist.  

For 95%, we need cutoffs for lower .025 and upper .025

The function in R:  qnorm.  qnorm is backwards normal.

```{r}
qnorm(.025)
```

```{r}
pnorm(-1.959964)
```

With cdf, we input a cutoff and get an area/prob.

With qnorm, it's backwards:  we input an area and get a cutoff.

Upper cutoff:

```{r}
qnorm(.975)
```

This is the "critical value" for 95% confidence:  zstar = 1.96.

Let's stick it all together for the CI:

low = phat - zstar*SE(phat)
    = phat - zstar*(sqrt(p*(1-p)/n))
    
upper = phat + zstar*SE(phat)
      = phat + zstar*sqrt(p(1-p)/n)
      
Example:  13 of you are women.  There are 22 students present.  Based upon this sample result, estimate the proportion of DU students who are female with a 95% confidence interval.

n = 22
phat = 13/22
se(phat) = sqrt(13/22*(1-13/22)/22)
zstar = 1.96

Low bound for CI:

```{r}
13/22-1.96*sqrt(13/22*(1-13/22)/22)
```

Upper:
```{r}
13/22+1.96*sqrt(13/22*(1-13/22)/22)
```


Based upon this sample result, we're 95% confident that the true proportion of female DU students is between 38.54% and 79.63%.  

Katie's question:  why bad?



Answer: huge range!! what's the real value??

Example:  Suppose in a sample of 200 DU students, 143 of them are from out-of-state.

Compute the 95% CI for the true proportion of DU students who are from out of state.

phat = 143/200
n = 200
zstar = 1.96

Low bound:

```{r}
143/200-1.96*sqrt(143/200*(1-143/200)/200)
```
Upper bound:
```{r}
143/200+1.96*sqrt(143/200*(1-143/200)/200)

```

Conclusion:  we can be 95% confident that the true proportion of DU students from out-of-state is between 65.2% and 77.8%.  

What about 99% confidence?

zstar:

```{r}
qnorm(.005)
```

zstar: 2.576

99% CI for p:

```{r}
143/200-2.576*sqrt(143/200*(1-143/200)/200)
```
```{r}
143/200+2.576*sqrt(143/200*(1-143/200)/200)
```

We can be 99% confidenct that the true proportion of DU students from out-of-state is between 63.3% and 79.7%.  

Wed quiz:  Lec 19 (no CIs)