---
title: "Math 120 Lec 22"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```





## The Meaning of "Confidence"

When we say "95% confident", what does that mean?  

Example:  let's estimate the proportion of 4cyl cars in the mpg dataset with automatic transmission.  Specifically, "auto(l5)":

```{r}
mpg %>% head
```

Let's do it for the whole dataset.

Q:  How many 4 cyl cars have auto(l5) transmission?

First, we need to filter out the 4cyl:

```{r}
mpg %>% filter(cyl == 4) -> fourCyl
head(fourCyl)
```

Next quetion:  what proportion of them have automatic transmission?

Let's count!

```{r}
sum(fourCyl$trans == "auto(l5)")  #count up auto(l5) trans
```

How many total?

```{r}
nrow(fourCyl)
```


Proportion:  6/81

```{r}
6/81
```

What if we take a sample, contsturct a CI based on that?

```{r}
sample(fourCyl$trans, 30) -> fourCylSample
```

Based on this sample, estimate the proportion of cars with auto:

```{r}
sum(fourCylSample == "auto(l5)")
```

Since n=30, our phat:

```{r}
4/30
```

So, there's error.  BUT, will a confidence interval work?  Let's try!  How about 95%:

```{r}
phat <- 0/30
n <- 30

#lower bound:

phat - 1.96*sqrt(phat*(1-phat)/n)
phat + 1.96*sqrt(phat*(1-phat)/n)



```

Our population proportion was 7.4% with auto.  Did our CI work?

Interpretation:  We are 95% confident that the true proportion of 4cyl cars with auto(l5) transmission is between 1.1% and 25.5%.

Since the true proportion was 7.4%, and that lies within the confidence interval, our confidence interval was successful!

In real-world scenarios, there's no way to know.  That's what the confidence is for:  confidence level is the "success rate" of this method.  If we take many samples, and contstruct CIs from all of them, about 95% will be successful, i.e., contain the true paramater.  

Not every sample will generate a "successful" CI.  That's what confidence level is all about, it tells us the success rate of our method for making CIs.

One last note about confidence:  95% DOES NOT MEAN that there's a 95% chance that p is in our specific interval!


## Planning ahead for Margin of Error

You know that increasing n decreases MOE.  We can use to plan in advance!  What if there's a specific magin of error that we want?  How big would n have to be in order to guarantee that MOE?

Example:  In 2015, the gender ratio at DU was 57% female.  A researcher wonders, how has the proportion of female students changed?  She wishes to construct a 95% CI with MOE no greater than 1%.  How big must n be in order to guarantee this?

Idea:  solve MOE backwards for n!

Since $$MOE = z^*\sqrt{\frac{p(1-p)}{n}},$$ 
we get

$$n = p(1-p)\left(\frac{z^*}{MOE}\right)^2$$

In this example, our estimate for p is .57, our zstar is 1.96, and desired MOE is 1%, .01.  Plug and chug:

```{r}
.57*(1-.57)*(1.96/.01)^2
```

We'd need to sample 9,416 DU students!  Not possible!  Our goal of 1% MOE is unreasonable! What about 10% MOE?

```{r}
.57*(1-.57)*(1.96/.10)^2

```

Need 95 students, more reasonable!  We can use this to plan our study!