---
title: "Math 120 Lec 13"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: simplex
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```

## Special Discrete Distribution:  **Binomial Dist**

Goal for next week or so:  let's get to know the most famous probability distros.

Our first new distro:  binomial!

## Example: tossing three coins

Suppose we flip a coin three times in a row.

a)  What are all possible outcomes?  How many?

There are 8 outcomes.

b) What's the probability that we observe exactly 1 head?

Since exactly three of the outcomes have exactly one head, that's 3/8 = .375.

Why this example?  Notice that each one has equal probability.

P(HHH) = P(HTH) = P(HHT) = ..... = P(TTT) = 1/8.  So, doesn't it seem kinda wasteful of us to count out all options, when each is the same?  Must we do each seperately?

What if we tossed 5 coins?  There's 32 ways.  What if 10 coins?  There's 1024 ways!  I don't wanna list those!

Fortunately, there's a shortcut:  The binomial dist.



## Binomial Random Variables

A discrete RV, X, is "binomial" if it counts the number of success out of a fixed number of trials.

Think:  success is "heads", # of trials is 3 (in last example).

Binomial variables must satisfy 4 properties:

- Only two possible outcomes:  success and failure.
- Fixed number of trials, n.  Decided in advance.  
- The probability of success is constant, i.e never changes.  We call this probability "p".
- Each trial must be indepndent.  

For our coin example:

- H or T (only two outcomes)
- n = 3
- p = 1/2  (prob of heads is constant)
- coin tosses are independent


## The distribution

If X is binomial, there's an easy way to find probailities.  (i.e., a special formula).

Every distribution has two versions:

1) The "pdf" or "probability density function" is what you guys have learned so far:  a list of all possible options, and their probs.
2) The "cdf" shows "cumulative" probability.  I.e., prob of that value or smaller.  Everything the left.


###  binom pdf
If X is binomial, n trials, chance p of success:

$$ f(x) =  {n \choose x} p^x(1-p)^{(n-x)} $$

where $$ {n \choose x} = \frac{n!}{(n-x)!x!} $$

This is pronounced out loud as "n choose x".

Let's digest all the pieces:

- ${n \choose x}$ gives us the total number of ways to get x successes in n trials
- $p^x$ tell the probability of having x successes. ( $\overbrace{p\cdot p\cdot p \cdots p}^n$)
- $(1-p)^{(n-x)}$ tell the probability of having n-x failures.  $\overbrace{(1-p)\cdot (1-p)\cdot (1-p) \cdots (1-p)}^{n-x}$

For the first time ever in my class, **you won't need to use this by hand**!!  Instead, know the "story" of the formula:  what are the pieces, and what do they mean?

We don't do it by hand because it's 2019.  Let's let non-humans do it, i.e. R!  It's easy:

```{r}
#Find P(X=1) with n=3 tosses, p=1/2 chance heads

#use the dbinom(x,n,p) function

dbinom(1, 3, 1/2)


```


Yaaaay! It gave the same answer, but so much easier.

In-class challenge:  If we toss a coin 50 times, what's the chance that exactly 12 of them are heads?

```{r}
#x=12, n=50, p=1/2
dbinom(12, 50, 1/2)
```

In-class challenge 2:

If we roll a dice 100 times, what's the chance we get a "4" exactly 18 times?

```{r}
#n=100     x=18     p=1/6

dbinom(18, 100, 1/6)
```

There's about a 9.7% chance.

###  cdf

Question:  what if we're interested in a range of outcomes?

Example:  If we toss a coin 100 times, what's the probability that we get less than 60 heads?

The hard way:

P(X<60) = P(X=0) + P(X=1) + P(X=2) + P(X=3) + ... + P(X=58) + P(X=59)

Oh no!  I don't wanna do all that!

Cdfs exist to solve this problem.

F(x) = the probability of observing x **successes or fewer**.  Cdfs always tell the "left-side" probabiliy.

In R:  use `pbinom()`.  This is a common pattern:  "d" means "density", and "p" means cumulative probability.

For individual outcomes, use "d".  For ranges, use "p".

Let's find our prob of less than 60 heads:

```{r}

#n=100, x<60, p = 1/2

pbinom(59, 100, 1/2)

```

There's about a 97.1% chance of getting fewer than 60 heads.

In-class challenge 3:

What's the probabilty of getting between 50 and 60 heads (not including 50 and 60)?  Still tossing the coin 100 times.  Use the cdf, `pbinom`

```{r}
#use cdf for x<60

pbinom(59, 100, 1/2) - pbinom(50 , 100, 1/2)
```


### Reading example 4.30

a)  Binomial?

- n=100 (check!)
- Two outcomes:  either yes pox or no pox.
- p = .9
- Independent?  Yes!  Pop of US is about 320 million, much much bigger than our sample (think of the 5% rule).


b) Since we're asking about one oucome (X=97), use dbinom() (i.e. the pdf):

```{r}
# n = 100     p = .9      x = 97

dbinom(97,100,.9)
```

There's about a .58% chance that 97 of the 100 had pox.

d) At least 1 out of 10

```{r}
# P(X>=10) = P(X=1) + .... + P(X=10)
#use cdf

pbinom(10 , 10, .9) - pbinom(0,10,.9)


```

## Examples
###  Sampling DU students

At DU, about 55% of students are female (this is true).  If we take a sample of 45 students, what's the probability that 

a) exactly 30 of them are female?

```{r}
dbinom(30,45, .55)
```


b)  more than 25 and less than or equal to 34 of them are female?

```{r}
pbinom(34, 45, .55) - pbinom(25 , 45, .55)
```

There's a 41% chance of obtaining a sample like this.

Note for the future:  this will help us interpret our sample results!

One little note: we can only use the binomial here because there are at least 1000 women at DU, i.e., large pop and thus independent.






###  Multiple Choice Exam

A student is taking a mult choice test with 30 questions, four options each:  ABCD.  Tragically, she hasn't studied and must guess on each question.

What's the probability that she passes, i.e., gets at least half of them correct?


```{r}
#x >= 15   n =30    p = 1/4

1 - pbinom(14 , 30, 1/4)


```

Note:  sum of all probs is 1 for any RV.

```{r}
pbinom(30,30,1/4)
```


###  Baby Tiger Snakes

[next time]

## Mean and variance of binomial

We've seen that every RV has a center (expected value) and spread (variance/std dev).

But, consider the exam example above.  In order to find the expected value, we'd need to do:

0*P(0)+1*P(1)+2*P(2)+..... + 29*P(29)+30*P(30)

Oh no! I'm way too lazy to do that!


Fortunately, there's a shortcut.  If X is binomial with n trials and prob p, then:

$$E[X] = n\cdot p$$

Huge shortcut from big nasty expected value formula!

$$Var[X] = n\cdot p\cdot (1-p)$$

thus

$$ STDEV[X] = \sqrt{n\cdot p\cdot (1-p)} $$

**Warning** These shortcuts only work for the binomial!!!


Example:  Multiple Choice Exam

We saw n = 30 question, p = 1/4 chance correct.  What's the expected and varaince?

```{r}
#Expected value = n*p

30*(1/4)
```

In the long run, we'd exped our guessing student to get about 7.5 questions correct on average (oh no!).

```{r}
# Variance

30*(1/4)*(1-1/4)


# std dev

sqrt(30*(1/4)*(1-1/4))

```

Here: the variance is 5.625 and the std dev is 2.37.  Note about the units:  units on variance are problems^2, which is hard to think about.

However, the std dev is intuitive:  we exepct in the long run that guessing students will get an average of 7.5 questions correctly (E[X]), on average students will be within 2.37 questions within this range:

7.5+-2.37 are common values for this dist.










###  Interpreting examples above

