---
title: "Math 120 Week 01"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```


# Wed Jan 19

Let's have class!

Check it out, we can calculate:

```{r}
2+2
```

Grey boxes:  calculation!

## What's a dataset?

In 120, care about:  **rectangular data**.

Three qualities:

- Each row (left-right) represents an **individual** ("case")
  Ex:  students (you guys!)
  
- Each column (up-down) represents a **variable**
  Ex:  our data has 4 variables
  
- Each "cell" contains exactly ONE measurement/observation.


## Descriptive Statistics

Say smart stuff about data!

WARNING!!!!  Q:  What **kind** of data is it?


Broadly:  two types of data:


- number data!  quantitative/numerical

break down:

  - discrete.  COUNTABLE!  whole numbers.  Ex:  siblings.
  - continuous. Measureable!  Any real number.  Ex:  height.
  
  WARNING:  Decide based on the variable, NOT the rounding
  
  Generally:  Physical measurements are continuous!  height, length, weight, volume, duration (time)
  
  Money:  let's agree it's continuous
  

- not numbers!  categorical/qualitative 

break down:

  - ordinal.  has intrinsic ordering.  Ex:  year in school
  - nominal.  no intrinsic ordering.   Ex:  genre
  
  
## Metadata!



















