---
title: "Math 120 Week 03"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```

# Monday Mar 1

## Effects of outliers (ctd)

More examples.

### Ex:  mpg data.  Predict cty mpg based on displ

```{r}
qplot(mpg$displ, mpg$cty)
```
```{r}
cor(mpg$cty, mpg$displ)
```

What if the following outlier is added:  x=20L, y=100mpg

Before outlier:  
After outerlier:  


Moral of the story:  

### Ex:  same mpg example (x=displ, y=cty).

Try the following outlier:  x=5, y=70.

Correl before:  
Correl after:  

Some effect, but not drastic.  Why?

Answer:  Leverage.

-Outliers with x-values that far outside the range of observations ....

-Outliers with x-vals that are within the range of observations  ...

Moral of the story:   doesn't change much.  
                     
### Final example:  mpg data, x=cty y=hwy

```{r}
qplot(mpg$cty, mpg$hwy)
```

```{r}
cor(mpg$cty, mpg$hwy)
```


Outlier:  x= 100, y=115.

Correl before:   
Correl after:   

Moral of the story

### BIG MORAL OF THE STORY:  

Outliers could do anything!  Have to look!




## Analyzing the model (ctd)

So far:  outliers

Now:  residuals.

Recall:  for an observation x,


    resid = 
    
(the vertical distance between a point and the line)

Think:  residuals are errors!

As always, visualize first!   Tool:  residual plot!

    x-axis:  
    y-axis:  
    
    
Example: mpg x=displ y=cty

Original:
```{r}
qplot(mpg$displ, mpg$cty)+geom_smooth(method="lm")
```


Residual plot:
```{r}
mpgModel <- lm(mpg$cty~mpg$displ)
qplot(mpg$displ, mpgModel$residuals)
```



Qualities of "bad" residuals:

- ...
- ... 

### Examples:  mpg x=displ and y=cty 

Residual plot shows ...

### Example:  mpg x=cty and y=hyw

```{r}
mpgModel2 <- lm(mpg$hwy~mpg$cty)
qplot(mpg$cty, mpgModel2$residuals)

```

...


### Good residuals follow a normal distribution

### Ex:  mpg, x=cty, y=hyw

```{r}
hist(mpgModel2$residuals)
```

...

### Ex:  mpg, x=displ, y=cty

```{r}
hist(mpgModel$residuals)
```

Can see right skew...



# Wed March 3

## Evaluating the model (ctd)

So far:

- Effects of Outliers
- Residuals (resid plot and dist of the residuals)

Today:

- Coefficient of Determination, r^2, (R^2).  

The coeff of det, r^2, tells us ...


Discussion:  ...

Math is easy:  literally (r)^2 !!  

    coef of det = (correl coeff)^2
    
### Ex:  

Suppose that, in a dataset about 50 men, the correlation coeff between height and weight is computed to be r=0.76.  In a linear model to predict weight based on height, compute and interpret the coef of det.


x = height
y = weight

Compute:

```{r}
.76^2
```

Interpret:  ...

### Ex:  mpg x=displ y=cty

```{r}
cor(mpg$displ, mpg$cty)
```

```{r}
(-.7985)^2
```

Interpretation: ...








## Ch 3 - Probability

### Two Defs

A random experiment/chance experiment/random process is any process that could result in two or more unknown outcomes.

Ex:  Flipping a coin.  Could be heads or tails, no way of knowing.  

Ex:  Rolling a dice.  Could be 1,2,3,4,5,6, no way of knowing.  

Ex:  Select a random DU student, record their height.  

Ex:  Take a random sample of 50 DU students, and record the proportion of them that are       Freshmen.

Goal:  how likely/unlikely is any outcome?  Probability!  Two main definitions:

1) Classical/Theoretical Probability.  ...

Ex:  Flip a "fair" coin.  What's the probability of seeing H?

     
     
Ex:  Roll a dice.  What's the chance that a prime number shows on top?  Prime numbers:  2, 3, 5.  

     
 

2)  Experimental/Relative Frequency Probability.  ...

Ex)  An imprisoned mathematician flips a coin 10,000 times and observes 4,981 Heads.


### Law of Large Numbers

... 

Good news! ...


Important notes about the picture on p82: ...







### Disjoint Events / Addition Rule

Two events A and B are disjoint (mutually exclusive) if ...

#### Ex:  

Are the following events disjoint?  Random process:  roll a dice.  

    A = the event that an even shows on top
    B = the event that a prime number shows
    


#### ADDITION RULE:  

If events A and B are disjoint, then:

     P(A or B) = P(A) + P(B)
     
Think:  For disjoint events, "or" means "+".  

#### Ex:  Roll a dice.  

     A = event that an odd shows up
     B = event that a number GREATER than 5 shows up
     
Compute P(A or B)


     
     
     





### Not Disjoint Events / General Addition Rule 

What if A, B have outcomes in common?  Venn Diagram!

For ANY two events A and B:

    ...
    
    
Think:  ...

Q:  What if we use this formula on disjoint events?  

A:  Not a probelm, since for disjoint A and B:  ...

#### Ex:  Roll a dice.

     A = event that an even shows up
     B = event that a prime shows up
     
Calculate P(A or B):


     ...
               
#### Ex:  Two way tables  (Contingency Tables)

mpg data, comparing year vs drive type

4 = 4 wheel drive
f = front wheel drive
r = rear wheel drive

```{r}
table(mpg$year, mpg$drv) %>% addmargins()
```

What's the probability that a random car is either a 1999 model or has rear wheel drive?

   
               
#### Deck of cards:

4 suits:  diamonds, hearts, spades, clubs

Each suit has 13 cards:

2, 3, 4, 5, 6, 7, 8, 9, 10, J, Q K, A

All together: 52 total cards

There are two colors:

 - All diamonds and hearts are red
 
 - All spades and clubs are black
 
The following in each suit are "face" cars:  J, Q, K


#### Ex)  

If a card is selected at random, what's the probability that it's either a face card or a red card?



### Probability Distribution

For a chance experiment, a probability distribution is:

- a list of all possible outcomes
- the probability for each

Usually, we'll express it as a table.  


Ex:  Suppose we flip a coin twice, and record the number of heads (x).  Construct a probability distribution for x.


    
    










### Complement/Sample Space

### Independence/ Mult Rule

### Sampling w/ Replacement vs Not

### Conditional Probabity 

### General Multiplication Rule

### Trees, Total Probability, and Bayes























