---
title: "Math 120 Week 11"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```



# Monday May 3

## Chi square goodness of fit (GOF)

Idea:  test an an entire distribution (not just a single category).

Test stat:

     chisquare = sum( (observed-expected)^2/expected)
               = sum( (O-E)^2/E )


Ex)  Do babies have a color preference for toys?  Experiment:  100 babies are each presented with four identical toys, except for color.  One's blue, one's green, one's red, and one's yellow.

We record which one they reach for first.  We observe the following:

    Blue   Green  Red  Yellow
    22      18     35   25

1)  Hypotheses

H0:  no preference (1/4 chance for each color)
Ha:  at least one color is picked a higher proportion of times

Expected:  25 for each color.

2) Test stat

    chisquare = (22 - 25)^2/25 + (18-25)^2/25 + (35-25)^2/25 + (25-25)^2/25

```{r}
(22 - 25)^2/25 + (18-25)^2/25 + (35-25)^2/25 + (25-25)^2/25

```

3) p-val

Look at chisquare table (dist for stdev)

df = #num categories - 1 = 4-1 = 3

     .05 < p < .1
     
4)  Since p>.05, we fail to reject H0.  There's not enough evidence to suggest that that proportion of colors chosen differ amongst the colors.  


Ex)  The Mars Company claims that the colors of MnMs follow this distribution:

     Blue  Yellow  Red  Green Brown
     20%    15%     35%  20%   10%

We use a bag of MnMs as a sample, and observe:

      Blue  Yellow  Red  Green Brown
       34    25     40     8     15
       
1)  Hyps

H0:  no difference in distribution
Ha:  at least one proportion differs from Mars's claim

Observed:

    Blue  Yellow  Red  Green Brown
       34    25     40     8     15

Expected:  total

     Blue   Yellow  Red   Green  Brown
     24.4    18.3   42.7  24.4   12.2

2)  Test stat

```{r}
(34-24.4)^2/24.4 + (25-18.3)^2/18.3 + (40-42.7)^2/42.7 + (8-24.4)^2/24.4 + (15-12.2)^2/12.2
```

    df = 5-1 = 4
    
    
3)  p-val


      .001 < p-val < .005
      
4)  Since p<.05, we reject H0.  We've found strong evidence that the color of MnMs doesn't follow the distribution that Mars claims.



##  Chi square test for independence

Two events A and B are independent if....

    P(A and B) = P(A)*P(B)
    
Ex)  The following data was collected from a sample of students

            Brown   Black   Other  Total
    Fresh    35      25       10    70
    Soph     30      35       15    80
    Total    65      60       25    150
    
Expected counts = (row total)*(column total)/(table total)

Expected:
             Brown  Black   Other
    Fresh    30.33   28    11.67
    Soph      34.67  32    13.33





1)  Hypotheses

H0:  yes independent (no relationship between year and hair color)
Ha:  no not independent (yes, some kind of dependence relation)

2)   chisquare = (35-30.33)^2/30.33 + (25-28)^2/28 + (10-11.67)^2/11.67 + (30-34.67)^2/34.67 + (35-32)^2/32 + (15-13.33)^2/13.33

```{r}
(35-30.33)^2/30.33 + (25-28)^2/28 + (10-11.67)^2/11.67 + (30-34.67)^2/34.67 + (35-32)^2/32 + (15-13.33)^2/13.33
```

df = (#rows-1)(#cols-1) = (2-1)(3-1) = 2

3) p-val

    pval > 0.3
    
4)  Since p>.05, we fail to reject H0.  We did not find strong evidence of a relationship between hair color and year in school.

Ex)  Is there a relationship between age group and opinion about whether climate change is anthropogenic.  Data:

            Yes    No   Total
    20-29   45     15     60
    50-59   32     28     60
    Total   77     43     120
    
Expected:

            Yes     No   
    20-29   38.5    21.5
    50-59   38.5    21.5

```{r}
(45-38.5)^2/38.5 + (15-21.5)^2/21.5+ (32-38.5)^2/38.5 + (28-21.5)^2/21.5
```

1)  Hyps

H0:  age group and opinion are independent (no relationship)
Ha:  they're -not- independent (yes relationship)

2) Test stat

(above)    chisquare = 6.12

df = (2-1)(2-1) = 1

3) p-val

     .01 < p-val < .02
     
4)  Since p<.05, we reject H0.  There's strong evidence of a relationship between age group and belief in anthropogenic climate change.





# Fri May 7

## Test for correlation

Recall:  r measures the strength and direction of a linear relationship.

If:

  - r>0,  pos lin rel
  - r<0,  neg lin rel
  - r=0,  no lin rel
  
Note:  "r" is a statistic!  Based on sample data!

What about population?  Population version of r:  rho.

Bad news:  the sampling dist for rho is very complicated.

Good news:  much simpler if we assume rho=0 (no linear relationship).

H0:  rho = 0
Ha:  rho >, <, != 0

test stat:

    t = [ r*sqrt(n-2) ]/[ sqrt(1-r^2) ]
    
    df = n-2
    
Ex)  iris

```{r}
cor(iris$Sepal.Length, iris$Sepal.Width)
```

For sepal length and sepal width, r = -0.118, n=150

Q:  is there evidence of a linear relationship between sepal length and sepal width?


1) Hyps

H0:  rho = 0
Ha:  rho != 0 

2)  test stat

```{r}
(-.118*sqrt(150-2))/sqrt(1-.118^2)
```

df = 150 - 2 = 148

3)  p-val

    .1 < p-val < .2
    
4)  Since p>.05, we fail to reject H0.  There's not strong evidence of a linear relationship between sepal length and sepal width.

Ex)  mpg data:  cty mpg vs hwy mpg

```{r}
cor(mpg$cty, mpg$hwy)
```

```{r}
nrow(mpg)
```


r = .956, n=234

Q:  Is there evidence of a positive linear relationship btwn cty and hwy mpg?

1) Hyps

H0:  rho = 0
Ha:  rho > 0 

2) test stat

```{r}
.956*sqrt(232)/sqrt(1-.956^2)
```

df = 232

3) p-val

    p-val << .05
    
4) Since p<.05, we reject H0.  There's super duper strong evidence of a positive linear relationship.  


WARNING:  The test for rho only shows if there's SOME linear relationship.  It DOESN'T tell you how strong that relationship is (specifically, what's rho).

How strong is the relationship?  You decide!





## Chi square practice

In survey, the following data was collected about species and preffered caffeinated beverage:

                Tea   Coffee   Total
     Giraffes   17     32      49
     Elephants  20     26      46
     Totals     37     58      95
     
     
Q:  Is there a relationship between species and beverage preference?


              Tea   Coffee   
     Giraffes  19.08 29.91
     Elephants  17.92 28.08

```{r}
(17-19.08)^2/19.08+(32-29.91)^2/29.91+(20-17.92)^2/17.92+(26-28.08)^2/28.08
```

1)  Hyps

H0:  no relationship (independent)
Ha:  yes relationshp (not independent)

2)  test stat

chisquare = 0.76

df = (2-1)(2-1) = 1


3) p-val

    p-val > .3
    
4) Since p>.05, fail to reject H0.  There's not strong evidence of a relationship between species and beverage pref.

##

```{r}
57+330+11+6
```

Observed:

    af-am   asian    white   latinx
     57      11       330      6


Expected counts:

    af-am   asian    white   latinx
   71.508   12.928   296.536  23.028


1) Hyps

H0:  the distribution matches the population demographics
Ha:  at least one is different


2) 
```{r}
(57-71.508)^2/71.508 + (11-12.928)^2/12.928+(330-296.536)^2/296.536+(6-23.028)^2/23.028
```
    
df = 4-1 = 3

3) p-val

     p-val < .001
     
4) Since p<.05, we reject H0.  We have found strong evidence that the demographic proportions in commercial representation doesn't match the population demos.





In a random sample of 64 Coca Cola cans, the mean volume of soda in the cans was computed to be 11.9oz, with stdev 0.05 oz.

Estimate the true mean volume contained in Coca Cola cans with 95% confidence.


    (statistic) +/- (crit val)*(stdev for sample)
    
    
df = 63,  tail area (each one) = .025

    tstar = 2.00
    
```{r}
11.9-2.00*.05/sqrt(64)
```
```{r}
11.9+2.00*.05/sqrt(64)

```
    
We're 95% confident that the true mean volume of soda in all the cans is between 11.89 and 11.91 oz.
  

## Central Limit Theorem

CLT:  as long as our sample size is "large enough", sample statistics (like xbar and phat) follow a normal distribution, NO MATTER WHAT POPULATION WE'RE SAMPLING FROM!

For xbar:  "large enough" means n>=30

For phat:  "large enough"  means both np and n(1-p) >= 10.

## Forwards vs backwards


Forwards:  Have a measurement, looking for a probability (area)

Ex:  Women in the us have mean height 64" stdev 2.4".  What percent of women are taller than 70"?

Backwards:  Have a region under curve/area/probability, looking for measurement.

Ex:  How tall must a woman be in order to be in the top 10% of heights for women in the us?


