---
title: "Math 120 Week 06"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```






# Mon April 5

Recall:  Statistical inference is the process of making mathematically supported conclusions about population parameters based on sample data.

Idea:  say something smart mu based on xbar.  

       say something smart about p based on phat.  
       
First:  

## Confidence intervals for p

So far:  know info about parameter, use it to find probabilty of a sample.

In reality:  don't know paramters!  Don't know mu, don't know p.  

Want to use our sample result to say something about population.

Confidence interval:  a range of reasonable values for a parameter.

Ex:  We're 95% confident that the **true population proportion** of DU who enjoy waterskiing is between 12% and 24%.  

Note:  CIs alwasys describe PARAMETERS (population), not statistics (sample).  

Idea:  middle % problem!  Based on our sample result, what are the midde 90%/95%/99% most likely values for parameter!



Ex:  In a sample of 100 DU students, 18 enjoy water skiing.  Estimate the proportion of DU students who enjoy water skiing with 95% confidence.


Check: is phat normal?

Check:

```{r}
100*(18/100)
```

```{r}
100*(1-18/100)
```

Both larger than 10, yes phat is normal!

Also:  at least 10 in the category, at least 10 out of the category.  

z = +/- 1.96

For phat:

      z = (phat - p)/sqrt( p(1-p)/n )

Idea:  solve for p!

Problem:  what's p(1-p)/n ??  Approximate with phat intestead.  

Lo:

     -1.96 = (18/100 - p)/sqrt( 18/100*(1-18/100) / 100)
     
     p = -1.96*sqrt( 18/100*(1-18/100) / 100) + 18/100


```{r}
-1.96*sqrt( 18/100*(1-18/100) / 100) + 18/100
```

Hi:

```{r}
1.96*sqrt( 18/100*(1-18/100) / 100) + 18/100
```

We are 95% confident that population proportion of all DU students who enjoy water skiing is between 10.5% and 25.5%.


## CI Concepts


Picture:

          (---------- statistic ----------)
          (-------------phat--------------)
          
Stat is always in the center.  Best guess!

The length of the "arms":  Margin of Error (MOE).  

Formula:  If phat is normal  (at least 10 in category and outside):

              phat +/- zstar*sqrt( phat(1-phat) / n)
              
              
where zstar is the z-score cutoff for the "middle %" (confidence level).  


Ex)  In Spring 2016, Prof Miller's stats class had 16 women out of 25 total students.

Estimate the proportion of all DU students who are female with 90% confidence.


Check:  is phat normal?

No!  16 in the category, 9 out of the category.  Fails the check!

With skepticism, let's find the CI:

    phat = 16/25
    
Find zstar:  1.645 (since 1.64 and 1.65 equally good).

CI:

low:

```{r}
16/25 - 1.645*sqrt(16/25*(1-16/25)/25)
```

```{r}
16/25 + 1.645*sqrt(16/25*(1-16/25)/25)
```

Based on this sample, we're 90% confident that the proportion of women at DU is betwen 48.2% and 79.8%.

Here, MOE:  

```{r}
1.645*sqrt(16/25*(1-16/25)/25)
```

THoguht: large margin of error.

Ex:  Same data (16 out of 25).  Construct a 99% CI for the proportion of women at DU.

Use zstar = 2.575 (between 2.57 and 2.58)

CI:  everything's the same, but zstar:

```{r}
16/25 - 2.575*sqrt(16/25*(1-16/25)/25)
```

```{r}
16/25 + 2.575*sqrt(16/25*(1-16/25)/25)

```

We're 99% confident that the proportion of women at DU is between 39.3% and 88.7%.

MOE:

```{r}
2.575*sqrt(16/25*(1-16/25)/25)
```

Greater confidence means greater MOE!  Bummer!











# Wed April 7

## CI for p

Formula

    phat +/- zstar*sqrt(phat(1-phat)/n)
    
zstar: z-score cutoffs for the "middle percent" based on confidence level

Q:  What does 95% confident mean?

Q:  What does it mean for a confidence interval to be successful?

Goal:  estimate p (pop proportion)

A "successful interval"  for p is one that CONTIAINS the true population proportion!

Let's investigate with a sample:

### Setup:  Coin Tosses

Fair coin.  Tossing 20 times, recording the proportion of heads.

phat = proportion of heads

Q:  does phat follow a normal dist?

np >=10 and n(1-p)>=10

n = 20, p = .5

```{r}
20*.5
```

```{r}
20*(1-.5)
```

yes!  n is large enough!  phat follow a normal dist!

Simulation:  pick random numbers, 0s and 1s.  1 = heads, 0 =tails.

For each sample, construct a 50% CI for p.

zstar:  look up area = .25  -> zstar = 0.67

## Meaning of Confidence

What does it mean to be 50% confident?  This:  if we took many samples of size n and constructed confidence intervals from each one, about half of them would be successful.  
Confidence level = SUCCESS RATE of this method.

NOT TRUE:  There's a 50%/95%/whatever% chance that the interval we got is correct.

No way of knowing!  

## Effects on MOE

We saw last time that bigger confidence means bigger MOE.  Bummer!


A CI is like a net:  bigger means more likely to succeed! 

Good news:  if n gets larger, MOE gets smaller!  Yes!

## Planning for sample size

Suppose we want certain confidence level and a particular margin of error.  How big must n be?


Recall, MOE:

    MOE = zstar*sqrt(phat(1-phat)/n)

Solve for n!

    n > phat(1-phat)*(zstar/MOE)^2
    
Need n to be whole number

Ex)  Last time:  99% CI for gender ratio based on Prof Miller's class had MOE of 24%.  (Data:  phat = 16/25).  

If wish a follow-up study to estimate the gender ratio with 95% confidence and MOE no greater than 1%, how many students must we sample?


    phat = 16/25
    n = 25
    MOE = .01
    zstar = 1.96
    
Plug and chug!

```{r}
(16/25)*(1-16/25)*(1.96/.01)^2
```

Uh-oh!  Sample is impossible!  Aren't that many DU students.  

Adjust expectations:  90% confidence, MOE of no greater than 5%.

   zstar = 1.645
   
```{r}
(16/25)*(1-16/25)*(1.645/.05)^2
```
  
Need at least 250 students to get a 90% CI with MOE no greater than 5%.  


Q:  What if we have no idea about the proportion!  No prior data!

A:  Use p=0.5, the worst case scenario.   Gives biggest possible MOE!  

Ex)  Prof Miller would like to estimate the proportion of DU students who enjoy water skiing with 96% and MOE no greater than 7%.

How many students would she need to include in her sample?

    phat = 0.5 (no prior data)
    MOE = 0.07
    
Critical value:  96% area in the middle, each tail has 2%.  Look up area=0.02 in table:

Use zstar = 2.05.  


```{r}
.5*(1-.5)*(2.05/.07)^2
```

In order to estimate p with confidence 96% and moE no greater than 7%, we'd need to sample at least 215 students.


# Friday April 9

## CI for mu

Ex)  In a sample of 35 DU students, the mean IQ was computed to be 113.4.

Construct a 95% CI for the true mean IQ of DU students.

Assume that the population stdev sigma = 15.  (!!!)

zstar:  95% confidence -> tail area = .025, zstar = 1.96

lo:

```{r}
113.4 - 1.96*15/sqrt(35)
```

hi:

```{r}
113.4 + 1.96*15/sqrt(35)

```

We are 95% confident that the true mean IQ of DU students is between 108.4 and 118.4.

Problem:  If we don't know what mu is (reason for the CI), why would we know sigma?????

We don't!  Best thing we can do:  use s instead!  (sample stdev)

Problem!  If we use s:

      (xbar - mu)/(s/sqrt(n))
      
does NOT follow a normal distribution!

# The "t" distribution

Very similar to the normal dist!  BUT, it accounts for not knowing sigma.

Similarities:

- same shape:  bell-shaped and symmetric
- both centered at zero

Differences:

- t distribution has thicker tails
- t critical values are bigger than zstar values!
- MOE for t CIs are bigger!  Bummer!
  - This is the price we pay for not knowing sigma.  Sad, but honest.  
- t dist depends on sample size.  Called "degrees of freedom", "df".  

        df = n-1
        
- As n (df) increases, the t dist shape gets closer and closer to z dist.  For very large n, there's not much difference between t and z.  



Ex) Suppose we have a samle of size n=33, and we wish to construct a 95% CI with t.  What's the critical value?

Here:  95% in the middle, each tail has 2.5%.  

Df = 33 - 1 = 32

From table:  tstar = 2.04.

Ex)  In a sample of 20 men at DU, the mean height is computed to be 70.1" and stdev 3.2".

Construct a 98% CI to estimate the true mean height of men at DU.

Note:  human height follows a normal dist, so it's ok to have small n!


tstar:  98% in the middle, each tail has 1%.  df = 20-1 = 19

tstar = 2.54

PLug and chug!

```{r}
70.1 - 2.54*3.2/sqrt(20)
```

```{r}
70.1 + 2.54*3.2/sqrt(20)

```

We are 98% confident that the true mean height of men at DU is between 68.3" and 71.9".












