---
title: "Math 120 Week 11"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```




# Wed April 21

## Power of a hyp test

So far:

    alpha = prob of type I = prob that we reject H0, but H0 is true
    
    beta = prob of type II = prob that we fail to reject H0, but H0 is false
    
Now:

    power = prob that we reject H0 (yay!) given that H0 is false (also yay!).
    
Of course, we hope power is big.  What effects it?

Problem:  if H0 is false, we can't compute the probability!  Don't know mu/p/sigma/whatever!  Same sitch with beta.  BUT:

           power = 1 - beta [Ie, complementary!]
           
So what if....

- alpha is bigger?  
  - If alpha is bigger, we reject more!  Thus, power is bigger!
  - Yay!  But, beta (type II) is bigger.  Bummer!
- beta is bigger?  Then power is smaller!  Bummer, but Type I less likely.

What about sample size?  Bigger sample size is better!

- Lowers both alpha and beta
- Increases the power

## Identifying Error Types

Ex)  A researcher wonders if a coin is unfair, ie, shows heads more than half the time.  

To investigate, he flips the coin 100 times, and the results lead him to reject the H0.

In reality, the coin is fair.  <- Magic Stats Genie!

What error, if any, has he made?


Here: hypotheses:

H0:  p=.5
Ha:  p>.5

Conclusion of test:  p>.5
Reality:             p=.5

Since we reject H0, but H0 was true, this is type I!

Ex)  A researcher suspects that mean height for women at DU is less than 64".  He conducts a hyp test, computes a p-val of 0.08.

If, in reality, the mean height of women at DU is 64", what error (if any) was made?  <- Magic Stats Genie!

H0: mu=64
Ha: mu<64

Conclusion:  Maybe mu=64 (no evidence against it).

Reality:           mu=64

No error!

           
## Connection Between Hyp Tests and CIs

Suppose a researcher wants to estimate what proportion of the time a coin shows heads.

She takes a sample of 50 coin tosses, and constructs a 95% CI:  

            (.493, .524)
            
If she'd conducted a hyp test to see if the proportion differs from 50%, what would the conclusion of her test be?

H0: p  = .5 
Ha: p != .5


According to her CI, the H0 is very reasonable!  We FAIL to reject H0!

Ex)  Based upon the delivery times from Prof Miller's past 30 pizzas, she constructs a CI for mean delivery time to be:

            (21.3min, 27.4min)
          
If she conducted a hyp test to investigate the mean time was greater than 20 min, does this sample support her?

H0: mu=20
Ha: mu>20

According to the CI, the H0 is not reasonable.  We'd reject H0!  Support claim that mean time >20.

In general:

- If H0 value is inside CI, then fail to reject H0.  
- If H0 value is NOT in Ci, then reject H0.  Seems unreasonable!

WARNING:  Confidence level and alpha have to "match"!

IE:

  - If 95% confidence, then alpha = .05
  - If 99% conf,       then alpha = .01









# Fri April 23

Quiz:

Prof Miller suspects that a coin is not "fair", in other words that the proportion of "heads" differs from 50%.

To test this, she flips the coin 100 times, and observes that 44 of them are heads.

Conduct an hypothesis test to see if this sample data supports Prof Miller's claim.  Show all steps as outlined in class.

1)  Hypotheses
H0: p =.5 
Ha: p!=.5    <- two tail

2)  Test stat:

```{r}
(44/100-.5)/sqrt(.5*(1-.5)/100)
```

3) p-val

tail area = 0.1151

-> p-val = .2302

4)  Since p>.05, fail to reject H0.  Not enough evidence to suggest the coin is unfair.




2) Prof Miller suspects that the mean IQ of DU students is higher than the national average of 100.  To investigate, she takes a random sample of 50 Du students to take IQ tests.  Their mean IQ is calculated to be 107.2  (stdev 10.4).

Does this sample data support her claim?  Conduct an hypothesis test, showing all steps as outlined in class.


1)  Hypotheses

H0:  mu=100
Ha:  mu>100

2) Test stat

```{r}
(107.2-100)/(10.4/sqrt(50))
```

3) p-val:  close to zero

4)  Since p<.05, we reject H0.  There's strong evidence to support the claim that mean IQ at DU is greater than 100.



3)  Hypotheses:

H0: drug has no effect         mu = ....
Ha: drug lowers cholesterol    mu < ....


##  Tests for two means (two-t tests)

So far: answered questions about one pop mean.

Now:  let's ask questions about how they compare!  Think: medical trial, pysyc study, etc etc.


For one mean:  approximate mu with xbar.  Samp dist:

         expected = mu       stdev = s/sqrt(n)
         
         
Q:  what if we two populations, and sample each?  Info:

     xbar1       xbar2
     s1          s2
     n1          n2
     
How do we study the difference?   mu1 - mu2!!!

Samp dist:

    observed:   xbar1 - xbar2

    expected:   mu1 - mu2
    
    stdev:    sqrt( s1^2/n1 + s2^2/n2 )
    
    
Test stat:

    t = (obs - exp)/stdev
      
      = (xbar1 - xbar2 - (mu1-mu2) )/sqrt( s1^2/n1 + s2^2/n2 )
      
      
CI:  

      observed +/- (crit val)*(stdev)
      
      (xbar1-xbar2) +/- tstar*sqrt( s1^2/n1 + s2^2/n2 )
      
      
Ex hyp test:

Prof Miller is teaching two sections of calc.  On the final exam, the first section scores an average of 76.1 (stdev 9.5).  The second section scores average of 78.4 (stdev 8.3).  There are 45 students in the first section, 39 in the second.

Is there statistically significant evidence of a difference in mean performance between the two sections?


     xbar1 = 76.1   xbar2 = 78.4
     s1    = 9.5    s2    = 8.3
     n1    = 45     n2    = 39
     


1) Hyps

H0:  mu1-mu2  = 0   ( mu1=mu2)   <- ALWAYS the H0 for 2-t tests!!!
Ha:  mu1-mu2 != 0   (same as mu1 != mu2)


2) test stat

```{r}
(76.1-78.4 - 0)/sqrt( 9.5^2/45 + 8.3^2/39 )
```

t = -1.18.   df = smaller of the two = 39-1 = 38

3) p-val

(note:  look for t=1.18, same tail area)

p-val > .2  (on the far left, two tail test)


4)  Since p>.05, fail to reject H0.  There's not strong evidence of a statistically significant difference in mean performance between the two sections.  


## CI for two means

A new drug has been developed to lower cholesterol (hopefully).

Two groups of study participants with similar ages and health markers participate in the study.

The 32 people in control group (no drug) have mean cholosterol 150.3, stdev 7.41.

The 36 people in the experimental group (yes drug) have mean cholesterol 138.2, stdev 9.8.

Construct a 95% confidence interval for the difference in mean cholesterol between the two.

      xbar1 = 150.3, s1=7.41, n1=32
      xbar2 = 138.2, s2=9.80, n2=36
      
tstar: one tail area = .025, df = 32-1 = 31

   -> tstar 2.04
   
Plug and chug!

```{r}
150.3-138.2 - 2.04*sqrt( 7.41^2/32 + 9.80^2/36  )
```

```{r}
150.3-138.2 + 2.04*sqrt( 7.41^2/32 + 9.80^2/36  )

```

We're 95% confident that:

                       7.8 < mu1 - mu2 < 16.4
                       
Q: Is there evidence that the drug lowers choleterol?

Ie, is mu1 > mu2?

Yes!  Since the difference is verly likely to be positive, the first one is bigger!  


