---
title: "Math 120 Week 06"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```




# Wed March 31

Simulation for the dice distribution.

Sample of dice, n=10:

```{r}
mean(sample(1:6, 1000, replace = TRUE))
```

Idea:  take MANY samples, find their means, make histogram.  













20000 Dice Rolls (n=1)

```{r}
n=1
qplot(replicate(20000, mean(sample(1:6, n, replace=TRUE))))
```






Thousands of means (xbar), n=2 dice:

```{r}
n=2
qplot(replicate(10000, mean(sample(1:6, n, replace=TRUE))))
```



Thousands of means (xbar), n=5 dice

```{r}
n=5
qplot(replicate(30000, mean(sample(1:6, n, replace=TRUE))),bins=30)
```



Thousands of means, n=30 dice

```{r}
n=30
qplot(replicate(10000, mean(sample(1:6, n, replace=TRUE))),bins=18)
```






Woah!  Looks familiar!

Summary:

- When looking at sample averages (xbar), smooshed towards the center.  Less likely to get very small xbar or very xbar.  
- As n (sample size) gets larger and larger, the sampling distribution for xbar gets closer and closer to a normal distribution!  Woah!
- We know the probability dist for xbar.  Ie, we can say how likely or unlikely SAMPLE MEANS are!

Summary:

The Central Limit Theorem says:  [...]


For xbar, "large enough" generally means [...]  

Know the shape:  normal.  

    z = (x-mean)/stdev

Center:  the mean stays the same!  

- the expected value for xbar is the same as the expected value for x

Spread:  stdev for xbar is 

    s_x / sqrt(n)
    
    = stdev for original population / sqrt of sample size
    
Note:  as n gets big, s_x/n gets small!  Bigger samples have less variation!


### Ex 1:  

Height for adult men is normal, mean=69", stdev = 2.7".  What's the probability that a randomly selected man is at least 70" tall?

    z = (70-69)/2.7
    
```{r}
(70-69)/2.7
```
    
Look up z = 0.37

P(X>70) = 

```{r}
1-0.6443
```

About 35.57% of men are at least 70" tall.


### Ex 2:  

Suppose we take a sample of 4 men and calculate their average height.  What's the probability that xbar is at least 70"?  

Ie,  P( xbar >= 70)  

mean:  stays the same.  mean for xbar = 69"

spread: smaller.  spread for sbar = 2.7/sqrt(4)  = 1.35

     P(xbar >= 70) = P(z > 1.35)
                   = 1-.9115
     
```{r}
1-.9115
```


### Ex 3)  

If we take a sample of n=30 men what's the prob that their mean height is at least 70"?  

mean:  69 (doesn't change)
stdev:  2.7/sqrt(30)

      P(xbar >= 70) = P(Z > (70-69)/ ( 2.7/sqrt(30) )   )

```{r}
(70-69)/( 2.7/sqrt(30) )
```

                    = P(Z > 2.03) = 1-.9788
                    
```{r}
1-.9788
```
                    

Observations:

- ...
- ..


### Ex:  

The life span of a brand of lightbulb has mean of 10,000 hours with stdev 500.  If we take a sample of 60 lightbulbs, what are the cutoffs for the middle 70% most likely sample averages?  

Look up area = .15 in table:  z = +/- 1.04


     z = (obs - mean)/stdev
     
     z = (xbar - mean)/(stdev/sqrt(n))
     
Low:

```{r}
-1.04*(500/sqrt(60))+10000
```

hi:

```{r}
1.04*(500/sqrt(60))+10000
```

The middle 70% most common sample means are between 9932.9, 10067.1.  



# Friday April 2

Last time:  samp dist for xbar.  

Idea:  find probability of a particular sample mean. 

Center: stays the same

     mu_xbar = mu_x
     
Think:  "the mean of the mean is the mean".  


Spread:  gets smaller with n

     sigma_xbar = sigma/sqrt(n)
     
Shape:

   - If X (original pop) is normal, then xbar (samp dist) is normal no matter what n is.  
   
   - CLT:  As long as n is large (n>=30), then xbar is normal no matter what original population we sample from!!  Holy cow!
   
   
   
   
   
### Ex:  

Giraffe height is normal with mean 18', stdev 2.5'.  If we take sample of 5 giraffes, what's the prob their mean height is at least 19' tall?

mu_x = 18  (population mean)
sigma_x = 2.5 (population stdev)

n=5

Find:  P(xbar >= 19)

Q: Can we use Z table?  Is xbar normal?  Since giraffe height is normal, xbar is normal even for small n.  

Dist for xbar:

mu_xbar = 18
sigma_xbar = 2.5/sqrt(5)

z:
```{r}
(19-18)/(2.5/sqrt(5))
```

Area on right:
```{r}
1-.8133
```

There's about an 18.67% chance of observing such a sample mean.  


## Proportions

Idea:  xbar is the statistic for quant data (numbers)
       phat is the statistic for qualitative data (categories)
       
       
Sampling Dist for phat:

Center:  doesn't change

     mu_phat = p
     
Spread:  decreases with n

     sigma_phat = sqrt( p(1-p)/n )

Shape:  CLT works for phat also!

     If n is "large enough", phat follows a normal dist no matter what.
     
### Ex)  

At DU, 19% of students are international students.  

If we take a sample of 100 DU students, what's the probability that at least a quarter (25%) of them are international?

    p = .19
    
    n = 100
    
Need: 

    P(phat > .25)

Q:  can we use Z table?  Is phat normal?

    Need:  np>=10  and n(1-p)>=10
    
```{r}
100*.19
```

```{r}
100*(1-.19)
```

YES!  CLT IS SATISFIED!  UZE Z TABLE!

Get z-score:

    z = (obs - exp)/spread
    
      = (.25 - .19)/sqrt( .19 * (1-.19) / 100 ) 

```{r}
(.25 - .19)/sqrt( .19 * (1-.19) / 100 )
```
z = 1.53

On table:  area to left = .9370

P(phat >= .25) =

```{r}
1-.9370
```

There's about a 6.3% chance of observing such a sample proportion.

### Ex)  

At DU,  77% of students are from out-of-state.  

In a sample of 60 students, what's the probability that at least half are from out of state?

    p = .77
    n = 60
    
Can we use z?

```{r}
60*.77
```

```{r}
60*(1-.77)
```

Find z for phat = .5

    (.5 - .77)/sqrt( .77 * (1-.77) / 60)

```{r}
(.5 - .77)/sqrt( .77 * (1-.77) / 60)
```

P(Z > -4.97) = almost 1

It's nearly certain that more than half of students in the sample are out-of-state.
















