---
title: "Math 120 Week 11"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```







# Monday April 12

## Hyp Tests


So far:

Confidence Intervals estimate a population paramater based on sample data.

Hypothesis tests answer a question/claim about a population parameter.


Ex:  We wonder if mean cholesterol is lower for patients who take new drug.

Hyp tests are criminal trials!

In a criminal trial:

1)  Q:  Innocent or guilty?  Assume by default that they're innocent, but the reason we're doing the trial is because they're think they're guilty!

2) Get evidence in the form of witness testimony and experts, etc.

3) A jury/judge decides how strong the evidence is.

4)  Conclusion:  if evidence is strong enough, then convict!  Guilty!  If evidence isnt' strong enough, can't convict.   Innocent.


Hyp test:  pop parameter is on trial!

1)  Two competing hypotheses:
    - H0:  The status quo.  Our assumption about the paramater.  Innocent!
    - Ha:  Our claim/question.  The thing we're trying to support.  Guilty!
    
    Note:
    
      - The H0 is always a statement of equality (=) about a pop parameter.
      - The Ha is always an inequality (<, >, !=).  
      
    H0 and Ha are always logical opposites.  
    
2)  The evidence is data!  Specifically:  z-score!  called "test stat"

3)  "Strength" of evidence comes from probability!  Called "p-value"

    The p-value is the probability of observing a sample result that's as (or 
    more) extreme than the one we got, assuming H0 is true!!!!
    
4)  Conclusion.  Small p-vals are evidence agains the H0.

     If p-val is small, then H0 says our data is very unlikely.
     Ie, if p-val is small, then H0 and data disagree!!
     
     If p-val is small, reject the H0!  There's strong evidence to support the
     Ha!
     
     If p-val is large, then H0 and data agree.  We did not find evidence
     to support Ha.  Boo!
     
     
## Hyp Tests for p (proportions)

Ex)  In a samle of 100 DU students, 62 have brown hair.  Does this provide strong evidence that more than half of DU students have brown hair?


Note:  phat is normal:

100*(62/100) > 10
(100)(1-62/100) > 10

1)  Hypotheses

H0:  p = .5
Ha:  p > .5


2) Test stat

z-score:

```{r}
(62/100 - .5)/sqrt(.5*(1-.5)/100)
```

3) p-val

```{r}
1-0.9918
```

p-value = .0082

4)  Q:  how small is small enough?  The "significance level" is the cutoff.  

    symbol:  alpha
    
    By default:  use alpha = .05.
    
Since p-val < alpha, reject H0.  We have strong evidence to support our claim that more than half of students have brown hair.

# Wed April 14

## Notes about hyp tests

- p-vals are always tail areas.  But:  which tail?  A:  Ha!

  If Ha has ">", then right tail.  
  If Ha has "<", then left tail.  
    
    
    
- Every conclusion has two parts:

  - If we have small p-val (p-val<alpha), we reject H0.  If p is large (>alpha), we fail to reject H0.
  - If we reject H0, then we've found strong evidence to support Ha.  Otherwise, we don't have enough evidence to support Ha.
  
WARNING:  If we fail to reject H0, this doesn't prove H0 is true.

## Significance Level

How small is small enough for p-val?  You decide!  The "cutoff" is called alpha, the significance level.  

If alpha not stated, assume alpha = 0.05.  

Ex)  Prof Miller has a proposition to change the faculty handbook.  For it to succeed, more than 2/3 of the faculty must vote to support it.  

To investigate whether there's enough support, she takes a sample of 50 facult members, and finds that 39 of them support her proposition.

Q:  Does this provide strong evidence that at least 2/3 of all faculty support her?


1) Hypotheses

H0: p=2/3
Ha: p>2/3    <- right tail test

2) test stat

```{r}
(39/50-2/3)/sqrt(2/3*(1-2/3)/50)
```

Note:  Ci for p:

     phat +/- zstar*sqrt(phat(1-phat)/n)


3) p-val

Justify:

```{r}
50*2/3
```
```{r}
50*(1-2/3)
```
Yes!  phat is normal!

left tail area = .9554

p-val:

```{r}
1-.9554
```

If H0 is true, there's a 4.46% chance of observing a sample result like the one we got.


4) conclusion

    p < alpha
0.0446 < 0.05

Reject H0.  We have strong evidence that more than 2/3 of faculty support the proposition.

## Two tail tests

So far:

Ha:  <   -> left tail test
Ha:  >   -> right tail test

Q:  what if Ha has "!="  (not equals)

A:  Then, both very large and very small results give evidence against H0.

Q:  Only one test stat, so only one tail area.  What's the other?

A:  For two tail tests, double the tail area for p-val!

Ex)  Prof Miller suspects that a dice is unfair.  In particular, she thinks the proportion of times that "3" shows up differs from 1/6.

To investigate, she rolls the dice 100 times and observes that "3" comes up 13 times.

1) Hypotheses

H0:  p  = 1/6
Ha:  p != 1/6    <- two tail test!!!

2) Test stat

```{r}
(13/100 - 1/6)/sqrt(1/6*(1-1/6)/100)
```

left-tail area:  0.1635

p-val:  2*.1635

```{r}
2*.1635
```

p-val = 0.327

4)  Conclusion


The p-val is greater than 0.05, so we fail to reject H0.  We didn't find strong evidence that the dice is unfair.  

## Hyp tests for mu (mean!)

Ex)  Prof Miller suspects that men at DU are taller, on average, then the national mean height for adult men of 69".  

To investigate, a sample is taken of 45 studnents.  Their mean height is computed to be 69.8" with stdev 2.5".

Does this provide strong evidence to support the claim?


1) Hypotheses

H0:  mu = 69
Ha:  mu > 69

2) Test stat

Note:  always use t-dist for means!

t-score:

```{r}
(69.8-69)/(2.5/sqrt(45))
```

df = 45 - 1 = 44


3) p-val

Use t table!

  .01 < p-val .025
  
  
4) Conclusion

Since p<.05, we reject H0.  There is strong evidence that the mean height is greater than 69".












    
    
 


# Friday April 16

## Error In Hyp Tests


Consequences of error.   For the following, describe the practical consequences of type I error and type II error.  Which do you think is worse?


1)  Criminal trial:

   H0:  innocent
   Ha:  guilty
   
   Questions:  do we think they're guilty?  are they really guilty?
   
Type I error:  We think they're guilty, but actually they're innocent!

Type II error:  We think think innocent, but they're actually guilty!

Could argue that type I is worse!  Seems worse to punish the innocent rather than risk letting the guilty go free.

2)  An EPA enforcement officer suspect that the mean concentration of Toxin A in a local water supply is above the federal limit of 0.8g/ml.  

To investigate, she takes water sample from 30 sources in the area, and computes the mean concentration of toxin A.


H0: mu = 0.8
Ha: mu > 0.8 

Error:  what do we think about water, whats true?

Type I:  We think the water is unsafe, but actually it's fine.  
Type II: We think the water is safe, but actually it's dangerous!

Could argue that type II is worse: in this case, people drink unsafe water!

## Effects on Type I and Type II errors

Impossible to eliminate error, but possible to tip the scales.  

Prob of type I error = alpha, the sig level!

Prob of type II error:  inversely proportional to alpha!  

If Type I is really bad (criminal trial), then choose small alpha.  

If Type II is really bad (toxic water), then choose larger alpha.














