---
title: "Math 120 Week 03"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```


# Monday Feb 15

## Z-scores

Motivatating question:  how "unusual" (or not) is a particular element in a dataset (quantitative).


First tool:  "z score".  Also called:  "standard score".  

For a paricular value, x, in a quant dataset, the z-score is:  ...


Similar to deviation:  x - xbar.  BUT!  Z-scores aren't ...

Picture:  


Math:

    z = 
    
    
One important application:  comparing values from different distributions!

Example:  The SAT and ACT are both college entrance exams.  The mean score on the ACT is 17, stdev of 4.3.  The mean score of the SAT is 765, stdev 52.

Amy took the ACT and scored 26.  Beth took the SAT and scored 872.  Who scored "better" relative to the pool of test takers?



## Empirical Rule (68-95-99.7 Rule)

*If a distribution is roughly bell shaped ('normal')*,  [btw, this is common!], then we can say something about where the data lies in the distribution:

1)  About 68% of the data in the dataset lies within +/- 1 stdev from the mean.  
2)  About 95% of the data in the dataset lies within +/- 2 stdevs from the mean.
3)  About 99.7% of the data lies within +/- 3 stdevs.

We can use this to "carve up" the normal distribution.  Picture!

Practical takeaways:

If your z-score is within +/-1, you're a very "typical" observation.  Grouped in with the middle 68%.  

If your z-score is between +/-1 and +/- 2, Still fairly normal.  Over a quarter (27%) of the data has a z-score close to yours.

BUT, if your z-score is larger the 3 or smaller than -3, YOU'RE QUITE UNUSUAL!  You're part of only 0.3% of the distribution!

If you see see a z-score above |4|, that's coco puffs!

Ex)  Giraffes have average height 18', stdev 2.5'.  What percent of all giraffes are between 15.5' and 23' tall?







# Wednesday Feb 17

## Practice:  Empirical Rule

In the US, the mean height of men ages 20-29 is 69" with stdev 2.7" (this is true).  For women ages 20-29, the mean height is 64" with stdev 2.4" (also true).  For both, heights follow a normal (bell-curve) distribution (yep, true!).


a)  If we select a man at random (rando man), what's the probability that he's between 63.6" and 71.7"?




b)  If we select a rando woman, what's the probabilit that she's over 71.2" tall?




## 5-number summary, boxplots, and outliers


Outliers are values/measurements in a dataset that are substantially unlike the others.  Unusual!  Way out there!

Outliers can have a significant effect on our analyses.  We want to identify them and consider their impact.

### 5-number summary

It's these 5 numbers:   


Ie, the 5 number summary cuts up our distro into ...


Ex:  56, 69, 71, 82, 83, 85, 88, 95.

Motivating question:  is "56" an outlier?

First:  5-number summary:



## Outliers

In a dataset, any value that's **larger** than ...


or **smaller** than ...


is considered an **outlier**.

Note: the spread of the distro is key for deciding who's an outlier!

Ex:  are any of the scores in our data above outliers?









# Friday Feb 19

## Boxplots




Boxplots are anothter visualization for a quant variable.  In short:  boxplots are visualizations of 


Notes:

- Boxplots are especially useful for ....

Ex:  compare sepal length across species

```{r}
boxplot(iris$Sepal.Length~iris$Species, horizontal=T)
```

Since all are plotted on same axis, we can compare:  setosa sepal lengths tend to be shorter with less variability.  Virgicina sepal lengths tend to be longer with more variability (even an outlier!).


```{r}
hist(iris$Petal.Length)
```

- Limitation of boxplots:  ...


## Correlation - The relationship between TWO quant variables

Book 8.1-8.2

Ex:  Height and weight.  Expect, in general, taller people to be heavier!

Always, first tool for analyzing relationships:  picture of course!  To study relationship between two quant variables:  scatterplot.

A *scatterplot* is an x-y plot of two quant variables in a rectangular dataset.  

Ex:  Let's visualize the relationship between displ and cty in the mpg data:

displ:  size of the engine (in liters)
cty :   fuel efficiency (in mpg)

```{r}
qplot(mpg$displ, mpg$cty)
```

Woah!  Bigger cars tend to be less efficient!














































































