---
title: "Math 120 Week 06"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```


# Mon March 8

Probability Dist Example:  Sum of two dice.

#### Ex:  Sum of Two Dice

We roll two dice, record x = their sum.  Construct a prob dist for x.  



Terminology:  The set of all possible values for X (everything that has non-zero probability) for a distribution is called the "support" for the distro.  

Previous example:  Support:  X = 2, 3, 4, ...., 12

####  Important properties of Prob Dists

1)  
2)  








### Complement

Notation:  The "complement" of event A is:  A' or A^C:  All possible outcomes NOT contained in A.  

Probability:

     P(A')  = 1 - P(A)
     
Think:  A and A' together have all possibilities.  So:

     P(A) + P(A') = 1
     
     
#### Ex.  "At least one"

Suppose we flip a coin three times.  What's the probability that at least one shows heads?

Note:  this is a big complex event!




#### Ex:  At least one

At DU, 19% of students are international.  If we take a random sample of 10 students, what's the probability that at least one of them is an international student?








### Independence/ Mult Rule


Recall:  Two events are **disjoint** if ...

We say that two events A,B are **independent** if ...


Otherwise, we say that they're "dependent" (or not independent).  

Ex:  The event "it's cloudy" and the event "it's raining" are likely not independent.  

Ex:  If we roll a dice twice, the outcome of the first doesn't affect the outcome of the second.  Independent!


#### Multiplcation Rule for Independent Events:

If A,B are independent:

     P(A and B) = P(A)*P(B)
     
Think:  for independent events, "and" means "times"

NOTE:  WORKS BOTH WAYS!  If P(A and B) = P(A)*P(B), then A and b are indep!


Ex:  Suppose we flip a coin 5 times.

a)  P(H, H, H, T, H) = ...


b)  P(H, H, H, H, H) = ...


#### Ex:  

Roll a dice.  What's the prob we get 1, then an even?



#### Ex:  

We pick a card from the deck (52 card poker).  Without replacing it, we pick a second.  What's the probability that both are hearts?

  

Q:  What if we'd replaced the first at random?  


Idea:  "sampling with replacement".  If not, events might NOT BE INDEPENDENT!  

#### Ex:  

There are 2300 Students at DU.  1200 students are women, 1100 students are men.  If we take a sample of two students, what's the probability that they're both women?

a)  "Without replacement"


b)  "With replacement"


# Wed Mar 10

### Sampling w/ Replacement vs Not


MORAL OF THE STORY:  If our population is "large", we can assume that sampling is INDEPENDENT.  Probabilities are basically the same.


What's "large enough"?  Answer:  ...


#### Ex:  

Suppose we take a random sample of 30 DU students.  Can we assume that these events (each student) are independent?



#### Ex:  

If we select two cards at random from a 52 card deck, can we assume indep?











### Conditional Probabity 

    P(A | B) = "probability of A given B"
             = if we know that B occured, how likely is A?
             
Ex:

     P(Rains | Cloudy) = ?
     
     
Make a picture!

Think:  "and on top, given on the bottom"


#### Ex:  

At DU, 51% of students are female, 17% of students are both female and a STEM major.

If we select a random female student, what's the probability that she's a STEM major?



#### Ex:  mpg

```{r}
table(mpg$year, mpg$drv) %>% addmargins()
```


a)  If we randomly select a car from 1999, what's the probability it has front wheel drive?


b)  Are the events "from 2008" and "rear wheel drive" indep?



MORAL:  for indep, check mult rule!

### General Multiplication Rule

So far:

A,B are disjoint ->
  
    P(A or B) = P(A) + P(B)
    
For ANY A,B:
    
    P(A or B) = P(A) + P(B) - P(A and B)
    
If A,B are independent:

    P(A and B) = P(A)*P(B)
    
    
Q?  What if A,B indep???? What about P(A and B)????

A:  Conditional probability!  Algebra:

    P(A and B) = ...
               
               
#### Ex:  

I'm walking my dog.  If he sees a squirrel, there's an 82% chance he'll bark.  If he doesn't see a squirrel, there's 27% chance he'll bark.  In my neighborhood, there's a 71% chance he'll see a squirrel.

On a random walk, what's the prob that my dog sees a squirrew and barks?






### Trees, Total Probability, and Bayes









