---
title: "Math 120 Week 01"
author: "Prof Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: yeti
    df_print: paged
  
  
---

```{r setup, include=FALSE}

#---------- RSTUDIO STARTER V 2.0  --------------#
#                    -Prepared with care by  AM ;D
                
                                                                          
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)      
library(tidyverse)                
library(ggthemes)                    

theme_set(theme_tufte() +                                     
  theme(text=element_text(family="sans")))  

#------------------------------------------------#
```




# Wednesday Feb 3


This is Rstudio!  I take notes here!

It's cool, because I can do math:

```{r}
1+1
```

Note:  the stuff in grey is code.  Unless it's just a mathematical computation, you don't need to worry about it.  But, DO worry about mathematical computations!

I can look at data!

```{r}
head(iris)
```

## Section 1.2 - Data Basics

What's data?  What's a dataset?

Here in 120, we'll be using "rectangular data".

In rectangular data:

- 
- 
- 


In 120, when we say say "dataset", we *always* mean rectangular data.  

Our goal:  say something smart about the individuals based on the data!


## Type of data

Not all data is created equal!  Important!  The tools we use for analysis vary drastically depending on what type of variable we're studying.


Main categories:  

- 
- 


We use -radically- different tools for analyzing each.  Ex:  doesn't make sense to find "mean species".  


We can further subdivide:

Quantitative data can be subdivided into two types:

-  
- 

Interesting caveats:

- Height:

- Money:   


Qualitative data can also be broken down into two types:

- 
- 

## Metadata

Sometimes, it's not always clear what the data represents.

Example:  mpg dataset

```{r}
head(mpg)
```

Not clear what all the variables are!

Answer:  metadata!  Data about data!  Describes:

- 
- 
- 


# Friday Feb 3

## Describing Data - Visualizations

Our goal:  say smart things about the individuals in our data.

Always first:  visualize!  Pictures's worth a 1000 words!


### Visualizing a single quantitative variable

Most important visulization:  

Every time we make a histogram, there are three important observations to make:

- 
- 
-  

Ex:  Number of siblings.  We observe... 

IMPORTANT WARNING ABOUT HISTOGRAMS:  The "bucket size" (the width of the rectangles, ie the number of rectangles) is ...

Another important shape word:  "bimodal" data...  


































