
---
title: "Math 220 Spring 2022 HW Solutions"
date: "Updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: yeti

      
  
---

```{r setup, include=FALSE}

                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans")) + 
  theme(axis.text.x = element_text(angle=-90))                            #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #

```


# Abbreviated Even Solutions

## HW 2


### 2.26

a)  P(A1^C) = 1-.12 = .88
b)  Since P(A1 or A2) = P(A1)+P(A2)-P(A1 and A2), we plug in:

.13 = .12 + . 07 - P(A1 and A2) 

Thus P(A1 and A2) = .06

c)  P(A1 and A2 and A3^C) = P(A1 and A2) - P(A1 and A2 and A3) = .05

d) P(at most 2) = 1 - P(A1 and A2 and A3) = .99

### 2.30

a)  8P3

```{r}
8*7*6
```

b)  30C6

```{r}
choose(30,6)
```

c) 8C2 10C2 12C2

```{r}
choose(8,2)*choose(10,2)*choose(12,2)
```

d) 

```{r}
83160/593775
```

e) P(all Zin or all Merlot or all Cab)

```{r}
(choose(8,6)+choose(10,6)+choose(12,6))/choose(30,6)
```


### 2.34

a)
```{r}
choose(25,5)
```


b) 
```{r}
choose(6,2)*choose(19,3)
```


c)

P(4)+P(5)
```{r}
(choose(19,4)*choose(6,1)+choose(19,5)*choose(6,0))/choose(25,5)
```


## HW 1

### 1.38

a)  Reported median is 125.
b)  New reported median is 130.  Conclusion:  even though the median is **not** sensitive to small fluctuations in data, it **can be affected** by grouping or rounding.

Related topic:  The shape of a histogram can change drastically depending on how wide your bins/categories are!

### 1.42

- Mean:  

$$\bar{y} = \frac{\sum{x_i+c}}{n} = \frac{x_1+x_2+\cdots +x_n+c+c+\cdots +c}{n}$$
$$=\frac{\sum{x_i}}{n}+\frac{nc}{n} $$

$$= \bar{x}+c$$


- Median:  Re-order the elements $x_i$ so that $x_1 \leq x_2 \leq \dots \leq x_n$.

If $n$ is odd, the median element is $x_{(n+1)/2}$.

Since this ordering also places the $y_i$s in ascending order, and so 

$$y_{(n+1)/2} = x_{(n+1)/2} + c$$

is the median.  If $n$ is even we can say the same about $\frac{x_{n/2}+x_{n/2+1}}{2}$

b. Similar arguments show that 

$$\bar{y} = c\bar{x}$$

and

$$ y_{(n+1)/2} = c\cdot x_{(n+1)/2}$$


### 1.44

a) range = 182.6-180.3 = 2.3


b) (note that I did not follow the "hint"; ok if you did)

$$s^2 = \frac{(180.5-181.4083)^2 + \cdots + (180.5 - 181.4083)^2}{11} = .5245$$

c)  $s = \sqrt{s^2} = .7242$

d)  

$$ \frac{394913.6 - \frac{4738894}{12}}{12-1} = .5245 $$








