
---
title: "Math 220 Spring 20203 Lecture Notes"
author: "Alice Miller"
date: "Updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: yeti

      
  
---

```{r setup, include=FALSE}

                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans")) + 
  theme(axis.text.x = element_text(angle=-90))                            #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #

```





## Math 220 Notes



R is cool, because we can do calculations together:

```{r}
2+2
1+5
```

(grey = code chunk)


## Data Basics

There are MANY formats for data.  In 220, we only care about one:  RECTANGULAR DATA ("tidy" data)!

3 things:

-  rows are CASES (individuals).  the people/animals/things/whateve that we're collecting data from.
-  columns are VARIABLES.  Ie, the qualities of the individuals that we record.
-  each "cell" contains EXACTLY ONE observation/measurment

## Types of Data

Two basic types:  quantitative (numerical), qualitative (categorical)

In all, 4 total types:

- DISCRETE numerical.  "countable" 
- CONTINUOUS numerical.  "measureable"
    Ex:   weight, volume, most phyiscal measurments
      Philospical question:  units?  What do we REALLY mean?
      
      Age = total duration of time since birth.   continuous.
      Age = social cohort.  discrete/categorical.             <-  Prof Miller's bias
      Height = measure of vertical length.  continuous.       <- Prof Miller's bias
      Height = categeory?
      
      Money.  pennies?  discrete.  crypto/interest.  Classroom agreement:  money is continuous.
    
      
- ORDINAL categorical data.  INTRINSIC ordering.  
- NOMINAL categorical data.  No natural ordering.

## Say smart things about data. (descriptive statistics)

Categorical variables:  proportions!  Ex:  in our class, 6/18 students had black hair.

phat = 0.3333....


Numerical data:  much more complex!





































