
---
title: "Math 220 Spring 20203 Lecture Notes"
author: "Alice Miller"
date: "Updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: yeti

      
  
---

```{r setup, include=FALSE}

                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans")) + 
  theme(axis.text.x = element_text(angle=-90))                            #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #

```





## Math 220 Notes




Hi!  This is RSTudio!


```{r}
2+2
```

# Week 1

## Wed Jan 18

### Data Basics

Many formats for data!

Here (and in all DA classes), we only care about RECTANGULAR DATA ("tidy data"):

3 things:

- rows are CASES (individuals).  Ie, from what person/animal/thing/whatev did we record?  Eg: you guys!
  remember:  rows are horizontal!
- columns are VARIABLES.  Ie, the qualities of the individuals that we care about!
   columns are vertical!
- each "cell" must contain ONLY one observation/measurement

### Types of Data

Big breakdown:  numerical (quantitative) or categorical (qualitative)
               ie:  number or not!
               
All together, 4 types:

-  discrete numerical data.  "countable".  Actually counting real object/things/ideas.
-  continuous numerical data.  "measureable". 
   Eg:  weight, volume, duration of time, etc.  
     
        Philosopical note:  be careful about WHAT you're measuring.  Ex:
        Age = integer number of years.  discrete.
        Age = duration of time living on earth.  continuous.  
        height = measuring tape.  continuous.  Possible to be 66.1", 66.7", 69.23423423"
        
        Weird example:  money.  gas?  pennies?  currency conversions? interest!
              Class agreement:  money is continous. 
              
- nominal categorical.  categorical, no INTRINSIC order.  
- ordinal categorical.  categrocial, YES natural order

MORAL OF THE STORY:  Not all data is the same.  We use DIFFERENT TOOLS for different types!

### Saying smart things about data (descriptive/summary stats)

Summary statistics for Categorical data:  proportion!

Ex:  In our class, 9/15 students had black hair.

phat:
```{r}
9/15
```


Summary statistics for Quantitative data:

For quant variables, always describe:

1) center.
2) spread.
3) shape.





