
---
title: "Math 220 Spring 20203 Lecture Notes"
author: "Alice Miller"
date: "Updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: yeti

      
  
---

```{r setup, include=FALSE}

                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans")) + 
  theme(axis.text.x = element_text(angle=-90))                            #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #

```





## Math 220 Notes




Hi!  This is RSTudio!


```{r}
2+2
```

# Week 1

## Wed Jan 18

### Data Basics

Many formats for data!

Here (and in all DA classes), we only care about RECTANGULAR DATA ("tidy data"):

3 things:

- rows are CASES (individuals).  Ie, from what person/animal/thing/whatev did we record?  Eg: you guys!
  remember:  rows are horizontal!
- columns are VARIABLES.  Ie, the qualities of the individuals that we care about!
   columns are vertical!
- each "cell" must contain ONLY one observation/measurement

### Types of Data

Big breakdown:  numerical (quantitative) or categorical (qualitative)
               ie:  number or not!
               
All together, 4 types:

-  discrete numerical data.  "countable".  Actually counting real object/things/ideas.
-  continuous numerical data.  "measureable". 
   Eg:  weight, volume, duration of time, etc.  
     
        Philosopical note:  be careful about WHAT you're measuring.  Ex:
        Age = integer number of years.  discrete.
        Age = duration of time living on earth.  continuous.  
        height = measuring tape.  continuous.  Possible to be 66.1", 66.7", 69.23423423"
        
        Weird example:  money.  gas?  pennies?  currency conversions? interest!
              Class agreement:  money is continous. 
              
- nominal categorical.  categorical, no INTRINSIC order.  
- ordinal categorical.  categrocial, YES natural order

MORAL OF THE STORY:  Not all data is the same.  We use DIFFERENT TOOLS for different types!

### Saying smart things about data (descriptive/summary stats)

Summary statistics for Categorical data:  proportion!

Ex:  In our class, 9/15 students had black hair.

phat:
```{r}
9/15
```


Summary statistics for Quantitative data:

For quant variables, always describe:

1) center.
2) spread.
3) shape.












## Friday Jan 20

### Quantitative Variables (saying something smart)

Note:  it doesn't make sense to talk about ONE number in a dataset.  Instead, we care about ALL of them, at the same time!  "Distribution".

ALWAYS MAKE A PICTURE FIRST!


For numerical data:  HISTOGRAM!

Ex:  class data:

```{r include=FALSE}
ourData <- read.csv("StudentData.csv")
```


```{r}
head(ourData)
```
Q:  what can we say about your height?

Hist!

```{r}
hist(ourData$Height..inches.)
```
Observations:  
 - typical height is around 67/68
 - spread:  looks like "most" values are btwn 62 - 72
 - looks symmetrical "ish"
 
 
 Q: what if we use more than 5 rectangles?
 
```{r}
hist(ourData$Height..inches., breaks = 10)
```
 Interesting new feature!  Bimodal!  Q:  could it be because of sex?
 
 MORAL OF THE STORY:  Follow the "Tao of the Histogram"!
 
 Ex:  our siblings
 
```{r}
hist(ourData$Number.of.Siblings, breaks = c(0,.99,1.99,2.99, 3.99))
```

Tao of the histogram!
 
### Summary Statistics
 
Ie, numerical summary of distribution.
 
- Measures of Center
  - Mean:  average!  "balancing point"
  - Median: "middle number"
  - Mode:  most common number
    
- Measures of Spread
    - Range = max - min.
       Con:  highly susceptiple to outliers.  
    - Standard deviation = "average distance btwn points and the mean"
        Formula:  on the board
        Note: must square to measure the TOTAL variation!
    - IQR.  Quartiles:  cutoffs for 25% (Q1) and 75% (Q3) in the dataset.
        IQR = Q3 - Q1
        
        Ex:  Data = {15, 16, 16, 18, 20, 27}
        
        Compute IQR = 20 - 16 = 4
        
        (M = 17, Q1 = 16, Q3 = 20)
    
         Meaning:  range of the MIDDLE 50% of the data!
         
         
A statistic is ROBUST if it is NOT strongly affected by skew or outliers.

Ex:  Data = {1,1,1,1,1,19}

Mean = 4, Med = 1.  Seems like median is better!


# Week 2





## Wed Jan 25






Distribution:  The salaries of all employees at a major corporation.

Which would provide a rosier picture of compensation:  the mean or the median?

A:  expect strong right skew (tail)!  expect mean to be bigger (since not robust).

### Outliers

Q:  what "counts" as an outlier? Need a test!

Q:  There are two sections of calculus.  Both take the same exam, both class means are the same.

Is it possible that a student who scores 95 in one section could be an outlier, but a student with same score in the other is NOT an outlier?


Yes!  Must factor in the spread!

Let's use IQR!  (robust!)

"Cutoffs" for outliers:

low =   Q1 - 1.5(Q3-Q1) 
                  IQR
hi  =.  Q3 + 1.5(Q3-Q1) 
                  IQR
                  
Ex:  Data = { 1, 2, 3, 4, 5, 6, 7, 8, 8, 20}

Q:  are there outliers?

M = 5.5 Q1 = 3, Q3 = 8,  IQR = 8-3 = 5

Cutoffs:

low:

```{r}
3 - 1.5*5
```

hi:
```{r}
8 + 1.5*5
```

### Boxplots

Great for comparison!

Ex:  iris dataset

```{r}
head(iris)
```

Ex:  sepal length

```{r}
boxplot(iris$Sepal.Length, horizontal = T)
```
Q:  compare species?

```{r}
boxplot(iris$Sepal.Length ~ iris$Species, horizontal = T)
```


### Ch 2 - Probability!

Plan:  stay to close to book!

Key building blocks:

A random process / chance experiment is any scenario where more than one unknown outcome could occur!

Ex:  Tossing a coin, rolling dice, sampling random student

Sample space = S = the set of all possible outcomes
Event.       = A = any subset of the sample space (what we care about!)

Ex:  Rolling a dice.   Let A be the event that we get an even number.
                       Let B be the event that we get a prime.

S = {1,2,3,4,5,6}
A = {2,4,6}
B = {2,3,5}

goal:  combine events into useful statements

3 tools:

 - Complement.  "not". A' is everything in S, NOT in A
 - Union.  A U B = all elements in either a OR b (could be both)
 - Intersection A intersect B = all elements in both A AND B
 
 Find:
 
 A' = {1,3,5}
 A U B = {2,3,4,5,6}
 A intersect B = {2}
 
 
 



## Friday Jan 27

2.2 Axioms of probability

The very basics!

For some event A, the probability of event A, denoted P(A), satisifies...

1). P(A) >= 0   (for any event A in our sample space S)
2). P(S) = 1.   (since S is all possibilities)
3). If A, B are mutually exclusive, then:

            P(A U B) = P(A) + P(B)

Ex) Using only the axioms, prove that P(A') = 1 - P(A)

Ex) Axioms, prove P(A) <= 1 for any A

Ex). P(A U B) = P(A) + P(B) - P(A and B)

Exercise:  prove from axioms!  (it's in the book!)

Ex). Cards!  If we draw at random, what's the prob that our card is a heart or a face?

   P(H U F) = P(H) + P(F) - P(H and F)
   
            = 13/52 + 12/52 - 3/52
            
### 2.3 - Counting Techniques

The PRODUCT RULE:  if we make a sequence of k decisions, 
        - with n1 options for the first, 
        - n2 options for the second, ....
        - nk options for the kth decision.
        
If the decision at any one stage doesn't affect any other, then there are:

         n = n1 * n2 * n3 * ... * nk total possibilities
         
Ex). We toss a coin 3 times.  How many possible outcomes?  Ex:  H H H

Ex). Roll 2 dice.  How many outcomes?

          6 * 6 = 36
        
Ex). In OH, license plates have 3 letters followed by 4 numbers.

How many plates are possible?

```{r}
26^3*10^4
```

Ex). Above, but no letter "O", can't end on an odd number.

### Permutations

Ex) How many arrangements of the letters in ALICE are there?

Ex) How many arrangments of the letters in BANANA are there?

Ex) There are 20 children.  3 swings.  How many seating arrangements are there?

###  Combinations

Ex). There are 20 children.  We need to pick a team of 3 children.  How many teams?

Problem:  teams have no arrangement/order!

What's the difference?  ORDER!!!   

Perms - order matters!

Combs - doesn't










 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 

