
---
title: "Math 220 Spring 2023 Homepage"
date: "Updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: yeti

      
  
---

```{r setup, include=FALSE}

                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans")) + 
  theme(axis.text.x = element_text(angle=-90))                            #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #

```





## Lecture Notes

These look great on your phone - easy to study on the go!

[Click here for notes](http://millera2.github.io/Math-220-S23/Math220Lec.html)


## Reading (Texbook)

I'll update each week.   

- Week 1: Ch. 1
- Week 2: Ch 2.1-2.3


## Homework

- For full credit, be sure to write neatly.
- Be sure your phyical (handwritten/printed) HW is in the pile before I take attendance at the beginning of class.  
- <b><big>Once attendance begins, I will no longer accept HW. </b></big>  
- I drop 2 low scores, in case of an emergency.

### HW 1 (Due Friday Jan 27)

Reading: All of chapter 1.  Some HW problems refer to conepts in the reading.

Practice Odds: 1.17, 1.39, 1.49 *[Solutions in the back of the book!]

**Evens to turn in:  $\leftarrow$ [The important part!]**  

- 1.38, p35
- 1.42, p36
- 1.44, p44
- 1.56, p46
- 1.24, p27 [You may use R for help, sketch your result on paper.  Example below.  Don't forget the Tao of the histogram.]

```{r}
yourData <- c(1,2,3,4,5,6)
hist(yourData, breaks = 3)
```




<br><br>

## Office Hours

I'm in Fellows 212.  I'll be there and available at:  

- Mon:  2:30-4pm
- Tues: 9-10:30am
- Wed:  2-4pm


<br><br><br><br>
<br><br><br><br>

