
---
title: "Math 220 Spring 2022 HW Solutions"
date: "Updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: yeti

      
  
---

```{r setup, include=FALSE}

                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans")) + 
  theme(axis.text.x = element_text(angle=-90))                            #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #

```


# Abbreviated Even Solutions

## HW Week 6



### 4.60

a) 
```{r}
pexp(200,.01386)-pexp(100,0.01386)
```

b) $P(X>\mu+2\sigma)$
```{r}
lambda <- .01386

1-pexp(1/lambda + 2/lambda, lambda) 

```

c) Set $F(x)=.5$, solve for $x$:

```{r}
-log(.5)/lambda
```

### 4.62 

**Note**: I didn't grade this one, but multiple students asked about it.  

First, note that $E[X]=0$, since this pdf is symmetric about $x=0$.  This isn't too tough to show algebraically, but also you could see immediately from a graph of $f(x)$.  

Now, for $\lambda$.  We know $\sigma = 40.9$, so let's use the familiar shortcut:

$$\sigma^2 = E[X^2] - \left(E[X]^2\right)$$
Using the above fact about expected value, we have:
$$\sigma^2 = E[X^2]$$

So, let's calculate that.

$$E[X^2]=\int_{-\infty}^{\infty}x^2\cdot f(x) dx = \int_0^\infty x^2 \cdot \lambda e^{-\lambda x} dx $$
You should spend a little time making sure you believe me on that second equality.  Now, this is exactly the same as $E[X^2]$ for the exponential distribution, which we know from class/textbook:  $E[X^2]=\frac{2}{\lambda^2}$.  (You can verify this yourself; integrate by-parts twice.) All together, we have
$$40.9^2 = \frac{2}{\lambda^2}-(0)^2$$

Solving, we find $\lambda$:

```{r}
sqrt(2/40.9^2)
```

### 5.40

Looks like most folks got the solution, so I won't repeat it here.  Here's a more interesting variant: what's the sampling distribution if we sample *without* replacement?

|       |         |         |    |
| --- | --- | ---  | --- |
| $X$   | 0       |  5      |   10  |
| $P(X)$| $\frac{{5 \choose 3} }{{10 \choose 3}}$        |   $\frac{{3 \choose 1}\cdot {5 \choose 2} + {3 \choose 2}\cdot{5 \choose 1}+{3 \choose 3} }{{10 \choose 3}}$      |  $\frac{{2 \choose 1}\cdot {8 \choose 2} + {2 \choose 2}\cdot{8 \choose 1}+{3 \choose 3} }{{10 \choose 3}}$   |


### 5.50

a) Since there's no "negative time", this distribution couldn't be symmetric.  Thus, not normal.  We'd be dead in the water without the CLT!

b) 

```{r}
pnorm(125, 120, 110/sqrt(50)) - pnorm(100, 120, 110/sqrt(50))
```

c) Note:  I assume this problem meant to say "mean courtship", not "total courtship".  If the book really meant total, then there's nearly 0% chance of exceeding 150 min total (since there are 50 pairs!).

For "mean courtship":
```{r}
1-pnorm(150, 120, 110/sqrt(50))
```

d) Absolutely not.  We're sampling from a non-normal population, so we need a large sample size (n>=30) for the CLT to kick in.  Yee haw!




## HW 3

### 2.48

a) P(A2 | A1) = P(A2 and A1)/P(A1) = .06/.12

```{r}
.06/.12
```


b) P(A1 and A2 and A3 | A1) = .01/.12

```{r}
.01/.12
```




c) P(exactly one defect | at least one defect)

First:  P(at least one ) = .14  and P(exactly 1) = .05

```{r}
.05/.14
```


d) P(A3^C | A1 and A2)

P(A1 and A2) = .06
P(A1 and A2 and A3^C) = .05

```{r}
.05/.06
```



### 2.64

a) P(W) = P(W and S)+P(W and M)+P(W and L)

```{r}
.6*.8+.3*.5+.1*.3
```

b)

P(S | W ) = P(S and W)/P(W)

```{r}
(.6*.8)/(.66)
```

Similarly,

P(M | W) =
```{r}
.3*.5/.66
```

P(L | W) = 

```{r}
.1*.3/.66
```


### 2.80

      P( 1or2 or 3and 4) 
           = P(1or2) + P(3and4) - P(1or2 and 3and4)
           = P(1or2) + P(3and4) - P(1or2)*P(3and4)

```{r}
.9+.9-.9^2 + .8^2 - (.9+.9-.9^2)*.8^2
```


## HW 2


### 2.26

a)  P(A1^C) = 1-.12 = .88
b)  Since P(A1 or A2) = P(A1)+P(A2)-P(A1 and A2), we plug in:

.13 = .12 + . 07 - P(A1 and A2) 

Thus P(A1 and A2) = .06

c)  P(A1 and A2 and A3^C) = P(A1 and A2) - P(A1 and A2 and A3) = .05

d) P(at most 2) = 1 - P(A1 and A2 and A3) = .99

### 2.30

a)  8P3

```{r}
8*7*6
```

b)  30C6

```{r}
choose(30,6)
```

c) 8C2 10C2 12C2

```{r}
choose(8,2)*choose(10,2)*choose(12,2)
```

d) 

```{r}
83160/593775
```

e) P(all Zin or all Merlot or all Cab)

```{r}
(choose(8,6)+choose(10,6)+choose(12,6))/choose(30,6)
```


### 2.34

a)
```{r}
choose(25,5)
```


b) 
```{r}
choose(6,2)*choose(19,3)
```


c)

P(4)+P(5)
```{r}
(choose(19,4)*choose(6,1)+choose(19,5)*choose(6,0))/choose(25,5)
```


## HW 1

### 1.38

a)  Reported median is 125.
b)  New reported median is 130.  Conclusion:  even though the median is **not** sensitive to small fluctuations in data, it **can be affected** by grouping or rounding.

Related topic:  The shape of a histogram can change drastically depending on how wide your bins/categories are!

### 1.42

- Mean:  

$$\bar{y} = \frac{\sum{x_i+c}}{n} = \frac{x_1+x_2+\cdots +x_n+c+c+\cdots +c}{n}$$
$$=\frac{\sum{x_i}}{n}+\frac{nc}{n} $$

$$= \bar{x}+c$$


- Median:  Re-order the elements $x_i$ so that $x_1 \leq x_2 \leq \dots \leq x_n$.

If $n$ is odd, the median element is $x_{(n+1)/2}$.

Since this ordering also places the $y_i$s in ascending order, and so 

$$y_{(n+1)/2} = x_{(n+1)/2} + c$$

is the median.  If $n$ is even we can say the same about $\frac{x_{n/2}+x_{n/2+1}}{2}$

b. Similar arguments show that 

$$\bar{y} = c\bar{x}$$

and

$$ y_{(n+1)/2} = c\cdot x_{(n+1)/2}$$


### 1.44

a) range = 182.6-180.3 = 2.3


b) (note that I did not follow the "hint"; ok if you did)

$$s^2 = \frac{(180.5-181.4083)^2 + \cdots + (180.5 - 181.4083)^2}{11} = .5245$$

c)  $s = \sqrt{s^2} = .7242$

d)  

$$ \frac{394913.6 - \frac{4738894}{12}}{12-1} = .5245 $$








