
---
title: "Math 220 Spring 2022 HW Solutions"
date: "Updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: yeti

      
  
---

```{r setup, include=FALSE}

                                                                          #
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)      #
                                                                          #
library(tidyverse)           #The tidyverse has all of our data tools     #
library(ggthemes)            #ggthemes makes our graphs look nicer        #
                                                                          # 
theme_set(                                #this sets the theme for all    #
  theme_tufte() +                         #plots that we make             #
  theme(text=element_text(family="sans")) + 
  theme(axis.text.x = element_text(angle=-90))                            #
  )                                                                       #
                                                                          #
                                                                          #
                                                                          #

```


# Abbreviated Even Solutions

## HW 1

### 1.38

a)  Reported median is 125.
b)  New reported median is 130.  Conclusion:  even though the median is **not** sensitive to small fluctuations in data, it **can be affected** by grouping or rounding.

Related topic:  The shape of a histogram can change drastically depending on how wide your bins/categories are!

### 1.42

- Mean:  

$$\bar{y} = \frac{\sum{x_i+c}}{n} = \frac{x_1+x_2+\cdots +x_n+c+c+\cdots +c}{n}$$
$$=\frac{\sum{x_i}}{n}+\frac{nc}{n} $$

$$= \bar{x}+c$$


- Median:  Re-order the elements $x_i$ so that $x_1 \leq x_2 \leq \dots \leq x_n$.

If $n$ is odd, the median element is $x_{(n+1)/2}$.

Since this ordering also places the $y_i$s in ascending order, and so 

$$y_{(n+1)/2} = x_{(n+1)/2} + c$$

is the median.  If $n$ is even we can say the same about $\frac{x_{n/2}+x_{n/2+1}}{2}$

b. Similar arguments show that 

$$\bar{y} = c\bar{x}$$

and

$$ y_{(n+1)/2} = c\cdot x_{(n+1)/2}$$


### 1.44

a) range = 182.6-180.3 = 2.3


b) (note that I did not follow the "hint"; ok if you did)

$$s^2 = \frac{(180.5-181.4083)^2 + \cdots + (180.5 - 181.4083)^2}{11} = .5245$$

c)  $s = \sqrt{s^2} = .7242$

d)  

$$ \frac{394913.6 - \frac{4738894}{12}}{12-1} = .5245 $$



## HW 2

### 2.26

a)  P(A1^C) = 1-.12 = .88
b)  Since P(A1 or A2) = P(A1)+P(A2)-P(A1 and A2), we plug in:

.13 = .12 + . 07 - P(A1 and A2) 

Thus P(A1 and A2) = .06

c)  P(A1 and A2 and A3^C) = P(A1 and A2) - P(A1 and A2 and A3) = .05

d) P(at most 2) = 1 - P(A1 and A2 and A3) = .99

### 2.30

a)  8P3

```{r}
8*7*6
```

b)  30C6

```{r}
choose(30,6)
```

c) 8C2 10C2 12C2

```{r}
choose(8,2)*choose(10,2)*choose(12,2)
```

d) 

```{r}
83160/593775
```

e) P(all Zin or all Merlot or all Cab)

```{r}
(choose(8,6)+choose(10,6)+choose(12,6))/choose(30,6)
```

### 2.48

a) P(A2 | A1) = P(A2 and A1)/P(A1) = .06/.12

```{r}
.06/.12
```


b) P(A1 and A2 and A3 | A1) = .01/.12

```{r}
.01/.12
```




c) P(exactly one defect | at least one defect)

First:  P(at least one ) = .14  and P(exactly 1) = .05

```{r}
.05/.14
```


d) P(A3^C | A1 and A2)

P(A1 and A2) = .06
P(A1 and A2 and A3^C) = .05

```{r}
.05/.06
```




## HW 3

### 2.64

a) P(W) = P(W and S)+P(W and M)+P(W and L)

```{r}
.6*.8+.3*.5+.1*.3
```

b)

P(S | W ) = P(S and W)/P(W)

```{r}
(.6*.8)/(.66)
```

Similarly,

P(M | W) =
```{r}
.3*.5/.66
```

P(L | W) = 

```{r}
.1*.3/.66
```


### 3.32

```{r}
x <- c(16,18,20)
P <- c(.2,.5,.3)
```

E[X]:
```{r}
sum(x*P)
```

E[X^2]:
```{r}
sum(x^2*P)
```

V(X): (using shortcut)
```{r}
333.2 - 18.2^2
```

b)  $Y=70X-650$, then E[Y]:
```{r}
70*18.2 - 650
```

c) Var[y]:
```{r}
70^2*1.96
```

(note that stdev = sqrt(V) = 137.2)


d) Note:  not linear!!!

```{r}
sum((x-.008*x^2)*P)
```

### 3.56

Note: X~binom(n=25, p=.02)

a) 

```{r}
dbinom(1,25,.02)
```

b)

```{r}
1-dbinom(0,25,.02)
```

c) 

```{r}
1-pbinom(1,25,.02)
```

d)  

E[X]:  

```{r}
.02*25
```

Stdev[X]:

```{r}
sqrt(25*.02*.98)
```

```{r}
.5+2*.7
```


Thus need:  X = 0,1

```{r}
pbinom(1,25,.02)
```

d) Avg time = Y =

$$ Y = \frac{3(25-X)+4.5*X}{25}=\frac{75+1.5X}{25}$$

Thus E[Y] = 
```{r}
(75+1.5*.5)/25
```




## HW 4

### 3.82

a)

```{r}
dpois(1,.2)
```

b)

```{r}
1-ppois(1,.2)
```

c)  Each disk:

```{r}
dpois(0,.2)
```

Together:

```{r}
(dpois(0,.2))^2
```


### 4.14

a) 

$\mu$:
```{r}
(7.5+20)/2
```
$\sigma^2$:
```{r}
(20-7.5)^2/12
```

b) 

$f(x) = 1/(20-7.5)$

```{r}
1/(20-7.5)
```

Thus 

$F(x) = .08(x-7.5)$ for $7.5\leq x \leq 20$,

$F(x)=0$ for $x<7.5$, and 

$F(x)=1$ for $x>20$.

c) 

```{r}
(15-10)*.08
```

d)

$\sigma$:
```{r}
sqrt((20-7.5)^2/12)
```

Thus for 1 stdev: $P(13.75-3.61 < X < 13.75 + 3.61)$ =

```{r}
.08*(13.75+3.61 - 7.5) - .08*(13.75-3.61 - 7.5)
```

And for 2 stdev:  there's 100% chance.

### 4.40

a)

$P(X<=40)$
```{r}
pnorm(40,43, 4.5)
```

$P(X>60)$
```{r}
1-pnorm(60,43, 4.5)
```

b) 

```{r}
qnorm(.25, 43, 4.5)
```






## HW 5

```{r}
data <- c(35, 49, 52, 54, 65, 51, 51, 47, 86, 36, 46, 33, 39, 45, 39, 63, 95, 35, 30, 48)

sort(data)

```


### 6.2

a)  45.95

b) 47.5

c) 16.81

d) 4/20

### 6.8

a) 68/80

b) (.85)^2


## HW 5

### 6.22 b

thetaa = 3.116  (solution in class notes)


### 7.14

Crit val:
```{r}
qnorm(.975)
```
z = 1.96

low:
```{r}
1427 - 1.96*325/sqrt(514)
```
hi:
```{r}
1427 + 1.96*325/sqrt(514)
```

We're 95% confident that true mean FEV levels are btwn 1398.9 and 1455.1.

b)  

n = 
```{r}
(1.96*320/25)^2
```

Use n=630.

### 7.32

crit val:

```{r}
qt(1-.005, 19)
```
t = 2.861

```{r}
1584 - 2.861*607/sqrt(20)
```
```{r}
1584 + 2.861*607/sqrt(20)
```




## HW 7

### 5.46

a)  

```{r}
1.6/sqrt(16)
```

b)

```{r}
1.6/sqrt(64)
```


c)  Larger n -> more likely to be close to the mean, since smaller stderr.




### 8.12

a)  

       H0:   mu = 1300
       Ha:   mu > 1300
       
b)  We'll reject H0 if xbar is any greater than:

```{r}
qnorm(.99, 1300, 60/sqrt(10))
```

Thus, if xbar = 1340, we fail to reject H0.


c)  If the true mean is 1350, then P(xbar < 1344) = 

```{r}
pnorm(1344, 1350, 60/sqrt(10))
```

Thus $\beta$=37.6%.

### 8.40

1)  Hypotheses

H0:  mu=48
Ha:  mu>48

2)  Test Stat

```{r}
(51.3-48)/(1.2/sqrt(10))
```

3) p-val

```{r}
1-pt(8.696, 9)
```

4)  Since p<.05, we reject H0.  There's strong evidence that mean strength is higher than 48 MPa.









